---
title: "Chapter 5 — Direct targets of Trh/Tgo and convergence with Ribbon"
author: "Yermek Aiten"
output: html_document
---

Direct targets of the Trh/Tgo heterodimer (Chapter 5). The `ch5-bootstrap` chunk
rebuilds the Trh/Tgo joint peak set (614 peaks) from the two IDR peak files, so
this file needs no objects from the Chapter 3 or Chapter 4 analyses.

- The direct-target set is the intersection of genes near Trh/Tgo peaks with the
  genes differentially expressed after *trh* RNAi in the modERN RNA-seq data of
  Fisher et al. (Table S43).
- The in-house *trh* RNAi microarray (FlyChip) is compared with that RNA-seq by
  genome-wide rank concordance and direction-validated subsets.
- The direct-target set is compared with the published Ribbon direct targets
  (Loganathan et al. 2016), and with the GSE28780 expression data (Chung et al. 2011).

Inputs are read from `data/final_peaks/`, `data/chapter5/` and
`data/reference_gene_lists/`; tables are written to `data/chapter5/`, figures to
`results/figures/`, and the values quoted in the text to
`results/reported_values.tsv` (final chunk).

```{r ch5-bootstrap, include=TRUE, eval=TRUE}
# ============================================================
#  CHAPTER 5 SELF-CONTAINED BOOTSTRAP
#  Defines: PROJECT_DIR, good_chr, FIGURES_DIR, fmt_p,
#           Trh_Tgo_joint_filt (614 peaks, rebuilt from raw narrowPeak)
#  No objects from Chapter 3 or Chapter 4 are required in memory.
# ============================================================

# ---- 1. Project paths ----
# Everything is derived from PROJECT_DIR.
## Repository root: the environment variable PROJECT_DIR, or the working directory.
PROJECT_DIR <- normalizePath(Sys.getenv("PROJECT_DIR", unset = "."), mustWork = TRUE)
if (!file.exists(file.path(PROJECT_DIR, "README.md")))
  stop("PROJECT_DIR does not point at the repository root: ", PROJECT_DIR)

DATA_DIR      <- file.path(PROJECT_DIR, "data")
CH5_DIR <- file.path(DATA_DIR, "chapter5")

## Values quoted in the dissertation are kept in results/reported_values.tsv
## (columns section, name, value). Writing a name that is already there
## replaces its value, so scripts can be re-run singly and in any order.
VALUES_FILE <- file.path(PROJECT_DIR, "results", "reported_values.tsv")
save_values <- function(section, values) {
  dir.create(dirname(VALUES_FILE), recursive = TRUE, showWarnings = FALSE)
  tab <- if (file.exists(VALUES_FILE)) {
    read.delim(VALUES_FILE, colClasses = "character", quote = "", comment.char = "")
  } else {
    data.frame(section = character(), name = character(), value = character())
  }
  for (nm in names(values)) {
    v   <- gsub("[\t\r\n]+", " ", paste(as.character(values[[nm]]), collapse = " "))
    tab <- tab[tab$name != nm, , drop = FALSE]
    tab <- rbind(tab, data.frame(section = section, name = nm, value = v))
  }
  tab <- tab[order(tab$section, tab$name), , drop = FALSE]
  write.table(tab, VALUES_FILE, sep = "\t", quote = FALSE, row.names = FALSE)
  cat(sprintf("%d values written to %s\n", length(values), VALUES_FILE))
}


## Colours
PAL_INK <- c(
  outline    = "#333333",   # bar and Venn outlines
  gridline   = "#DDDDDD",
  annotation = "#666666",   # p-values, n = counts
  threshold  = "#882255",   # significance thresholds, +/-1.96 lines
  fit        = "#4477AA",   # fitted / OLS trend lines
  neutral    = "#BBBBBB",   # non-significant, background sets
  band       = "#EDEDED",   # shaded stage/facet bands and strip backgrounds
  highlight_band = "#4477AA", # tint marking an agreement region (used at low alpha)
  emphasis   = "#CC3311",   # the one highlighted subset in an otherwise neutral plot
## Ink for stars printed on filled marks (Ch4 4x2 matrix, Fig 4.9); distinct from `annotation`, which sits on the white panel.
## Black (#000000) chosen: worst-case WCAG contrast 4.34 across PAL_DIV; valid for PAL_DIV only, not PAL_SEQ (viridis).
## For PAL_SEQ marks nudge labels off the dot; exception is geom_tile heatmaps, which use ink_on_seq() (worst-case 4.16:1).
  on_mark    = "#000000",

  ## Two-tier ink for ink_on_seq(). `on_mark_dark` is #1A1A1A rather
  ## than the #000000 above: on the LIGHT half of viridis both clear the floor
  ## comfortably, and #1A1A1A matches the ink weight of the surrounding axis text,
  ## so the matrix does not look heavier than the panel around it.
  on_mark_dark  = "#1A1A1A",
  on_mark_light = "#FFFFFF"
)

PAL_HETERO <- c(
  "Trh/Tgo"  = "#117733",   # green    L* 44
  "Sim/Tgo"  = "#E69F00",   # amber    L* 71
  "Dys/Tgo"  = "#66CCEE",   # sky      L* 77
  "Sima/Tgo" = "#AA3377"    # magenta  L* 42
)

PAL_SUBUNIT <- c(
  "Trh"  = PAL_HETERO[["Trh/Tgo"]],
  "Sim"  = PAL_HETERO[["Sim/Tgo"]],
  "Dys"  = PAL_HETERO[["Dys/Tgo"]],
  "Sima" = PAL_HETERO[["Sima/Tgo"]],
  "Tgo"  = "#DDCC77"
)

PAL_SEQ <- function(n = 9) hcl.colors(n, "viridis")

PAL_BINARY <- c(pos = "#CC6677", neg = "#4477AA")

pal_binary <- function(pos_label, neg_label) {
  setNames(c(PAL_BINARY[["pos"]], PAL_BINARY[["neg"]]), c(pos_label, neg_label))
}

PAL_DIRECTION <- c(
  down = PAL_BINARY[["pos"]],  # down on knock-down  => factor ACTIVATES the gene
  up   = PAL_BINARY[["neg"]],  # up on knock-down    => factor REPRESSES the gene
  any  = "#DDCC77"             # pooled, either direction
)


## Bar chart
thesis_text_col <- function(fill) {
  rgb <- grDevices::col2rgb(fill) / 255
  lum <- 0.299 * rgb[1, ] + 0.587 * rgb[2, ] + 0.114 * rgb[3, ]
  ifelse(lum > 0.5, "black", "white")
}

thesis_stars <- function(p) {
  ifelse(is.na(p), NA_character_,
  ifelse(p < 1e-3, "***",
  ifelse(p < 1e-2, "**",
  ifelse(p < 0.05, "*", "n.s."))))
}

thesis_p_expr <- function(p, digits = 2) {
  if (is.na(p)) return(NA_character_)
  ## The formatted value is QUOTED, so plotmath prints it as formatted rather
  ## than re-formatting it as a number: unquoted, a Fisher p of 1.000 renders
  ## as "1" and 2.10 as "2.1", losing the fixed significant figures that
  ## must be preserved at every mention.
  if (p < 1e-3) {
    parts <- strsplit(formatC(p, format = "e", digits = digits), "e")[[1]]
    sprintf("italic(p)=='%s' %%*%% 10^{%d}", parts[1], as.integer(parts[2]))
  } else {
    ## `digits` governs both branches. "fg" formats to significant digits
    ## rather than decimal places; flag "#" keeps trailing zeros (p = 1
    ## prints 1.0, not 1).
    sprintf("italic(p)=='%s'",
            formatC(signif(p, digits), format = "fg", digits = digits,
                    flag = "#"))
  }
}

thesis_bar <- function(data, x, y, fill = x,
                        value = NULL, value_fmt = "%.1f%%",
                        count = NULL, sig = NULL,
                        facet = NULL, n_lab = NULL, n_lab_size = 3.0,
                        fills = NULL, y_max = NULL, y_lab = NULL,
                        dashed_at = NULL, global_p = NULL, caption_sig = NULL,
                        inside_frac = 0.28, bar_width = 0.62, base_size = 11,
                        title = NULL, facet_free_x = TRUE) {

  d <- as.data.frame(data)
  d$.x <- d[[x]]; d$.y <- d[[y]]; d$.fill <- d[[fill]]

  ## primary value label
  if (!is.null(value)) {
    d$.value <- as.character(d[[value]])
  } else if (!is.na(value_fmt)) {
    d$.value <- sprintf(value_fmt, d$.y)
  } else {
    d$.value <- NA_character_
  }
  d$.count <- if (!is.null(count)) as.character(d[[count]]) else NA_character_
  d$.sig   <- if (!is.null(sig))   as.character(d[[sig]])   else NA_character_

  ## y range + headroom (space for the star that always sits above the bar)
  y_top_data <- max(d$.y, na.rm = TRUE)
  if (is.null(y_max)) y_max <- y_top_data * 1.32
  gap  <- 0.045 * y_max            # one label-line height, scaled to axis
  inside_cut <- inside_frac * y_max

  ## fill -> contrast colour for inside text
  if (!is.null(fills)) {
    missing_fill <- setdiff(unique(as.character(d$.fill)), names(fills))
    if (length(missing_fill))
      stop("thesis_bar(): fill level(s) not found in `fills` (would render grey): ",
           paste(sprintf("'%s'", missing_fill), collapse = ", "),
           "\n  Available in `fills`: ",
           paste(sprintf("'%s'", names(fills)), collapse = ", "))
  }
  fill_levels <- if (!is.null(fills)) names(fills) else unique(as.character(d$.fill))
  fill_cols   <- if (!is.null(fills)) fills else scales::hue_pal()(length(fill_levels))
  names(fill_cols) <- fill_levels
  d$.txtcol <- thesis_text_col(fill_cols[as.character(d$.fill)])

  ## ---- per-bar label geometry ------------------------------------------
  d$.inside <- d$.y >= inside_cut
  has_count <- !is.na(d$.count)
  # value line: inside -> just below bar top; above -> primary value uppermost
  d$.val_y   <- ifelse(d$.inside, d$.y - gap,
                       ifelse(has_count, d$.y + 2 * gap, d$.y + gap))
  d$.val_col <- ifelse(d$.inside, d$.txtcol, "black")
  # count line: inside -> below the value; above -> nearer the bar than value
  d$.cnt_y   <- ifelse(d$.inside, d$.y - 2 * gap, d$.y + gap)
  d$.cnt_col <- ifelse(d$.inside, d$.txtcol, "black")
  # significance always above the bar; clears above-bar labels when present
  above_labels <- (!d$.inside) & (has_count | !is.na(d$.value))
  d$.sig_y    <- d$.y + ifelse(above_labels, 3 * gap, gap * 1.4)

  ## ---- build plot -------------------------------------------------------
  p <- ggplot(d, aes(x = .x, y = .y, fill = .fill)) +
    geom_col(width = bar_width, colour = "black", linewidth = 0.4,
             show.legend = FALSE)

  if (!is.null(dashed_at))
    p <- p + geom_hline(yintercept = dashed_at, linetype = "dashed",
                        colour = PAL_INK[["neutral"]], linewidth = 0.5)

  ## value label
  if (any(!is.na(d$.value)))
    p <- p + geom_text(aes(y = .val_y, label = .value, colour = .val_col),
                       fontface = "bold", size = 3.5, show.legend = FALSE,
                       na.rm = TRUE)
  ## count label
  if (any(!is.na(d$.count)))
    p <- p + geom_text(aes(y = .cnt_y, label = .count, colour = .cnt_col),
                       size = 3.0, show.legend = FALSE, na.rm = TRUE)
  ## per-bar significance (above, spaced)
  if (any(!is.na(d$.sig)))
    p <- p + geom_text(aes(y = .sig_y, label = .sig),
                       fontface = "italic", size = 3.4, colour = PAL_INK[["annotation"]],
                       na.rm = TRUE)

  p <- p +
    scale_colour_identity() +
    scale_fill_manual(values = fill_cols) +
    scale_y_continuous(limits = c(0, y_max), expand = expansion(mult = c(0, 0)))

  ## facets, free_x so each panel shows only its own x categories
  if (!is.null(facet))
    p <- p + facet_wrap(stats::as.formula(paste0("~", facet)), nrow = 1,
                        scales = if (facet_free_x) "free_x" else "fixed")

  ## per-facet n = ... annotation, top-right
  if (!is.null(n_lab)) {
    key <- if (!is.null(facet)) facet else NULL
    nd  <- if (!is.null(key)) d[!duplicated(d[[key]]), ] else d[1, , drop = FALSE]
    nd$.nlab <- as.character(nd[[n_lab]])
    parse_n  <- any(grepl("==|~", nd$.nlab))
    p <- p + geom_text(data = nd,
                       aes(x = Inf, y = Inf, label = .nlab),
                       hjust = 1.08, vjust = 1.6, size = n_lab_size, colour = PAL_INK[["annotation"]],
                       fontface = "italic", parse = parse_n,
                       inherit.aes = FALSE)
  }

  ## single centred p-value, drawn once (as a plot tag), not per-facet,
  ## so a faceted figure shows one central p-value instead of one per panel.
  plot_tag <- NULL
  if (!is.null(global_p)) plot_tag <- parse(text = thesis_p_expr(global_p))[[1]]

  cap <- caption_sig
  p <- p +
    labs(x = NULL, y = y_lab, title = title, caption = cap, tag = plot_tag) +
    theme_classic(base_size = base_size) +
    theme(
      ## BOLD: the panel letter is the one bold element of the house style.
      plot.tag          = element_text(size = base_size + 2, face = "bold"),
      plot.tag.position  = c(0.5, 0.86),
      strip.text       = element_text(face = "bold", size = base_size),
      strip.background = element_rect(fill = NA, colour = "black", linewidth = 0.6),
      axis.title.y     = element_text(size = base_size - 0.5),
      axis.text.x      = element_text(size = base_size - 2, colour = "black"),
      axis.text.y      = element_text(size = base_size - 2, colour = "black"),
      axis.ticks.x     = element_blank(),
      panel.spacing    = unit(1.0, "cm"),
      plot.caption     = element_text(hjust = 0, size = base_size - 2.5,
                                      colour = PAL_INK[["outline"]]),
      plot.title       = element_text(size = base_size, face = "bold"),
      plot.margin      = margin(8, 10, 8, 8)
    )
  p
}


## Gene symbols of the Fisher et al. (2023) table
map_fisher_symbols <- function(symbols) {
  suppressPackageStartupMessages(library(org.Dm.eg.db))
  u   <- unique(trimws(symbols[!is.na(symbols) & nzchar(symbols)]))
  out <- data.frame(SYMBOL = u, FLYBASE = NA_character_, route = NA_character_,
                    stringsAsFactors = FALSE)
  sel <- function(k, kt) {
    k <- intersect(k, AnnotationDbi::keys(org.Dm.eg.db, keytype = kt))
    if (!length(k)) return(data.frame(key = character(0), FLYBASE = character(0)))
    d <- suppressMessages(AnnotationDbi::select(org.Dm.eg.db, keys = k,
                                                keytype = kt, columns = "FLYBASE"))
    d <- d[!is.na(d$FLYBASE), , drop = FALSE]
    names(d)[1] <- "key"; d
  }
  # 1. current symbol
  d1 <- sel(u, "SYMBOL"); d1 <- d1[!duplicated(d1$key), ]
  i  <- match(out$SYMBOL, d1$key)
  out$FLYBASE <- d1$FLYBASE[i]; out$route[!is.na(i)] <- "symbol"
  claimed <- unique(out$FLYBASE[!is.na(out$FLYBASE)])
  # 2. unambiguous alias, 3. CG number
  for (rt in c("ALIAS", "FLYBASECG")) {
    rest <- out$SYMBOL[is.na(out$FLYBASE)]
    if (rt == "FLYBASECG") rest <- rest[grepl("^CG[0-9]+$", rest)]
    d <- sel(rest, rt)
    amb <- unique(d$key[duplicated(d$key)])
    d <- d[!d$key %in% amb & !d$FLYBASE %in% claimed, , drop = FALSE]
    d <- d[!duplicated(d$FLYBASE) & !duplicated(d$key), , drop = FALSE]
    j <- match(out$SYMBOL, d$key); hit <- is.na(out$FLYBASE) & !is.na(j)
    out$FLYBASE[hit] <- d$FLYBASE[j[hit]]
    out$route[hit]   <- tolower(rt)
    claimed <- unique(c(claimed, d$FLYBASE))
  }
  stopifnot(!anyDuplicated(out$FLYBASE[!is.na(out$FLYBASE)]))
  out
}

# ---- Shared helpers ----
REFERENCE_LISTS <- file.path(PROJECT_DIR, "data/reference_gene_lists")
stopifnot(dir.exists(REFERENCE_LISTS))


# Derived paths (match the Chapter 3/4 layout)
FINAL_DIR     <- file.path(PROJECT_DIR, "data/final_peaks")
FIGURES_DIR   <- file.path(PROJECT_DIR, "results", "figures")
CH5_INPUT_DIR <- file.path(PROJECT_DIR, "data", "chapter5")
CH5_OUT_DIR   <- CH5_INPUT_DIR
dir.create(FIGURES_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(CH5_OUT_DIR, recursive = TRUE, showWarnings = FALSE)

# ---- 2. Constants & helpers (inlined from Chapter 3 so no sourcing needed) ----
good_chr <- c("chr2L", "chr2R", "chr3L", "chr3R", "chr4", "chrX")

# fmt_p: format a p-value in scientific notation, e.g. "4.39e-13"
# (identical to the Chapter 3 definition)
fmt_p <- function(p, digits = 2) {
  formatted <- formatC(p, format = "e", digits = digits)
  parts     <- strsplit(formatted, "e")[[1]]
  paste0(parts[1], "e", as.integer(parts[2]))
}

# ---- 3. Packages ----
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(rtracklayer)
  library(ChIPseeker)
  library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
  library(org.Dm.eg.db)
  library(GenomeInfoDb)
})
txdb <- TxDb.Dmelanogaster.UCSC.dm6.ensGene

# ---- 4. Rebuild Trh_Tgo_joint_filt from raw IDR narrowPeak files ----
# This is the canonical Chapter 4 recipe (chunk `trh-tgo-joint`):
#   import -> chr-filter + dedup (clean_peak_set) -> pintersect -> dedup -> 614.
# It depends only on the two narrowPeak files on disk, not on any in-memory
# object from another chapter.
extraCols_narrowPeak <- c(signalValue = "numeric", pValue = "numeric",
                          qValue = "numeric", peak = "integer")

trh_narrowPeak_file <- file.path(FINAL_DIR, "TRH_IDR0.05.final.narrowPeak")
tgo_narrowPeak_file <- file.path(FINAL_DIR, "TGO_IDR0.05.final.narrowPeak")
stopifnot(file.exists(trh_narrowPeak_file), file.exists(tgo_narrowPeak_file))

clean_peak_set <- function(gr, label) {
  gr_chr   <- gr[seqnames(gr) %in% good_chr]
  gr_chr   <- keepSeqlevels(gr_chr, good_chr, pruning.mode = "coarse")
  gr_dedup <- gr_chr[!duplicated(gr_chr)]
  cat(sprintf("%s: %d raw -> %d chr-filtered -> %d deduplicated\n",
              label, length(gr), length(gr_chr), length(gr_dedup)))
  gr_dedup
}

Trh_ChIP_peaks <- import(trh_narrowPeak_file, format = "BED",
                         extraCols = extraCols_narrowPeak)
Tgo_ChIP_peaks <- import(tgo_narrowPeak_file, format = "BED",
                         extraCols = extraCols_narrowPeak)

Trh_clean <- clean_peak_set(Trh_ChIP_peaks, "Trh")
Tgo_clean <- clean_peak_set(Tgo_ChIP_peaks, "Tgo")

ovlp_trh_tgo  <- findOverlaps(Trh_clean, Tgo_clean,
                              minoverlap = 1, ignore.strand = TRUE)
Trh_Tgo_joint <- pintersect(Trh_clean[queryHits(ovlp_trh_tgo)],
                            Tgo_clean[subjectHits(ovlp_trh_tgo)])
mcols(Trh_Tgo_joint) <- mcols(Trh_Tgo_joint)[, intersect(
  c("signalValue", "qValue"), colnames(mcols(Trh_Tgo_joint))), drop = FALSE]
Trh_Tgo_joint_filt <- keepSeqlevels(Trh_Tgo_joint, good_chr,
                                    pruning.mode = "coarse")
Trh_Tgo_joint_filt <- Trh_Tgo_joint_filt[!duplicated(Trh_Tgo_joint_filt)]

# ---- 5. Verify the canonical invariant ----
stopifnot(inherits(Trh_Tgo_joint_filt, "GRanges"),
          length(Trh_Tgo_joint_filt) == 614L)
cat(sprintf("Bootstrap complete. Trh_Tgo_joint_filt = %d peaks (canonical 614).\n",
            length(Trh_Tgo_joint_filt)))
```


---

```{r ch5-input-paths, include=TRUE, eval=TRUE}
# Chapter 5 inputs under CH5_INPUT_DIR: Fisher_combined_DEGs.csv (934 genes), rib-ts-sg-direct-targets-complete.csv (295 genes).
# Concordance chunk also needs R13.vsn.limma.fdr.txt.gz, R14.vsn.limma.fdr.txt.gz, modERN_Resource_RNAi_Table_S43_trh

P_FISHER_DEGS <- file.path(CH5_INPUT_DIR, "Fisher_combined_DEGs.csv")
P_RIB_TARGETS <- file.path(CH5_INPUT_DIR, "rib-ts-sg-direct-targets-complete.csv")

for (P in c(P_FISHER_DEGS, P_RIB_TARGETS)) {
  if (!file.exists(P))
    stop("Missing required Chapter 5 input CSV: ", P,
         "\nPlace it under: ", CH5_INPUT_DIR)
}
```


---

```{r ch5-load-inputs, include=TRUE, eval=TRUE}
# ---- Load gene lists with column-name tolerance ----
# Finds the gene-ID column whatever it is named.
find_id_col <- function(df, patterns = c("^gene$", "^FBgn", "^flybase",
                                         "gene_id", "fb.*id", "^id$", "FBgnID")) {
  for (pat in patterns) {
    hit <- grep(pat, names(df), ignore.case = TRUE, value = TRUE)
    if (length(hit) > 0L) return(hit[[1]])
  }
  stop("No gene-identifier column found in: ",
       paste(names(df), collapse = ", "))
}
read_gene_list <- function(path) {
  l1  <- trimws(readLines(path, n = 1))
  hdr <- !grepl("^\"?FBgn[0-9]{7}\"?$", l1)   # first line is a gene => no header
  df  <- read.csv(path, stringsAsFactors = FALSE, header = hdr)
  # Headerless files are single-column by construction; read.csv names that
  # column V1, which find_id_col cannot match, so take it positionally.
  ids <- as.character(if (hdr) df[[find_id_col(df)]] else df[[1L]])
  unique(ids[nzchar(ids) & !is.na(ids)])
}

fisher_degs      <- read_gene_list(P_FISHER_DEGS)
rib_target_genes <- read_gene_list(P_RIB_TARGETS)

# marray_st13_genes / marray_st14_genes are not loaded here. They are
# produced by the concordance chunk (ch5-microarray-fisher-concordance) and
# exported to the global environment at the end of that chunk. The concordance
# chunk runs below, before any downstream chunk that uses them.

cat("Fisher RNA-seq DEG list:        ", length(fisher_degs),      "genes\n")
cat("Loganathan Rib direct targets:  ", length(rib_target_genes), "genes\n")
```


---

```{r ch5-fig-paths, include=TRUE, eval=TRUE}
# ---- Output paths for Chapter 5 figures ----
# Figures are saved to FIGURES_DIR (set in the main Chapter 3 Rmd).
# Naming convention: fig5_N.png at width-x-height suitable for the text.

if (!exists("FIGURES_DIR")) {
  stop("FIGURES_DIR is not defined. Source Chapter 3 Rmd first, ",
       "or set FIGURES_DIR <- 'path/to/figures' manually.")
}
dir.create(FIGURES_DIR, recursive = TRUE, showWarnings = FALSE)

suppressPackageStartupMessages({
  library(ggplot2)
  library(gridExtra)
  library(grid)
  library(patchwork)
})
```

---

```{r ch5-fig-trh-expression-line, include=TRUE, eval=TRUE}
# Fig 5.3: trh knock-down efficacy, stages 11-15, values read from files not hardcoded.
# CG6883-RA is the only trh probe (FBgn0262139); per-replicate aggregation of S*.M columns gives limma log2FC -0.054, +0.132, +0.384, +0.558, +0.081.
# RT-qPCR gives the single-gene measurement; array probe kept as secondary trace with detection floor noted.

stage_ids <- 11:15

read_limma_stage <- function(st)
  read.delim(file.path(CH5_INPUT_DIR, sprintf("R%d.vsn.limma.fdr.txt.gz", st)),
             skip = 1, check.names = FALSE, stringsAsFactors = FALSE)

##, 1. trh on the array (probe CG6883-RA) -----------------------------------
trh_array <- do.call(rbind, lapply(stage_ids, function(st) {
  d <- read_limma_stage(st)
  r <- d[d$TargetID == "CG6883-RA", , drop = FALSE]
  stopifnot(nrow(r) == 1L)
  data.frame(stage = st,
             log2FC = as.numeric(r$logFC),
             SE     = as.numeric(r$logFC) / as.numeric(r$t),
             padj   = as.numeric(r[["adj.P.Val"]]))
}))
stopifnot(all(trh_array$padj > 0.05))   # array never calls trh significant

## 2. trh by RT-qPCR: point estimates only, no independent error
## log2FC/SE/p from prep-level Ct in trh_qpcr_recomputed.xlsx (sheet dCt_per_prep)
## Arms R=trh-RNAi, V=WT; ddCt=dCt(R)-dCt(V), log2FC=-ddCt; 3 preps/arm/stage, paired to rp49
TRH_CT_XLSX <- file.path(CH5_INPUT_DIR, "trh_qpcr_recomputed.xlsx")
stopifnot(file.exists(TRH_CT_XLSX))
.dct <- readxl::read_excel(TRH_CT_XLSX, sheet = "dCt_per_prep")
names(.dct) <- c("arm", "arm_meaning", "stage", "prep", "Ct_rp49", "Ct_trh", "dCt")
stopifnot(nrow(.dct) == 30L, all(sort(unique(.dct$arm)) == c("R", "V")))

trh_qpcr <- do.call(rbind, lapply(stage_ids, function(st) {
  dR <- .dct$dCt[.dct$arm == "R" & .dct$stage == st]
  dV <- .dct$dCt[.dct$arm == "V" & .dct$stage == st]
  stopifnot(length(dR) == 3L, length(dV) == 3L)
  tt <- t.test(dR, dV, var.equal = FALSE)          # Welch: preps are independent
  se <- sqrt(var(dR) / 3 + var(dV) / 3)
  # Exact two-sided permutation p: enumerate all choose(6,3) = 20 ways of
  # splitting the six preparations into two groups of three, and count how many
  # give a mean difference at least as extreme as the observed one. This makes no
  # distributional assumption, and its FLOOR is 2/20 = 0.05 one-sided-equivalent
  # (0.10 two-sided) whatever the effect size, reported alongside the Welch p
  # so the resolution limit of an n = 3 design is stated rather than implied.
  .all <- c(dR, dV)
  .idx <- utils::combn(6L, 3L)
  .perm <- apply(.idx, 2L, function(i) mean(.all[i]) - mean(.all[-i]))
  p_perm <- mean(abs(.perm) >= abs(mean(dR) - mean(dV)) - 1e-12)
  data.frame(stage = st, log2FC = -(mean(dR) - mean(dV)), SE = se,
             p_welch = tt$p.value, p_perm = p_perm)
}))
# The permutation floor is a property of the design, not the data: with 3 vs 3
# no arrangement can return less than 2/20. This is asserted so that a future change in
# replicate number cannot silently invalidate the reported result.
stopifnot(min(trh_qpcr$p_perm) >= 2 / choose(6, 3) - 1e-12)
stopifnot(nrow(trh_qpcr) == 5L, !any(is.na(trh_qpcr$log2FC)), !any(is.na(trh_qpcr$SE)))
# Check: the CSV trh row must agree with the Ct derivation, or the two sources
# have drifted apart and the concordance panels would disagree with the figure.
.csv_trh <- sapply(stage_ids, function(st) {
  q <- read.csv(file.path(CH5_INPUT_DIR, sprintf("st%d_validation_unstacked.csv", st)),
                row.names = 1, check.names = FALSE, encoding = "UTF-8")
  rownames(q) <- sub("^\ufeff", "", rownames(q)); names(q) <- trimws(names(q))
  as.numeric(q[["RT-qPCR"]][match("trh", rownames(q))])
})
stopifnot(max(abs(.csv_trh - trh_qpcr$log2FC)) < 1e-4)

##, 3. gene-level pass counts ----
# Semantics: a GENE passes if any of its probes passes. Probes carrying a
# comma-separated multi-FBgn assignment are expanded to one entry per gene,
# matching the concordance chunk's clean_limma().
genes_passing <- function(st, use_fold = TRUE) {
  d   <- read_limma_stage(st)
  lfc <- as.numeric(d$logFC); pa <- as.numeric(d[["adj.P.Val"]])
  k   <- !is.na(lfc) & !is.na(pa) & pa < 0.55 &
         (if (use_fold) abs(lfc) >= log2(1.4) else TRUE)
  fb  <- trimws(unlist(strsplit(as.character(d$FBgn[k]), ",")))
  length(unique(fb[grepl("^FBgn", fb)]))
}
n_pass_fold <- vapply(stage_ids, genes_passing, integer(1))
n_pass_padj <- vapply(stage_ids, genes_passing, integer(1), use_fold = FALSE)

##, 4. what the array could have detected (stage 13, best-powered) ----------
d13   <- read_limma_stage(13)
p13   <- as.numeric(d13$P.Value); q13 <- as.numeric(d13[["adj.P.Val"]])
p_thr <- max(p13[q13 < 0.05], na.rm = TRUE)      # largest raw p still at BH<0.05
t_crit    <- qt(p_thr / 2, df = 2, lower.tail = FALSE)
se_st13   <- trh_array$SE[trh_array$stage == 13]
mde_log2  <- t_crit * se_st13
mde_fold  <- 2^mde_log2

##, 5. depletion vs cross-platform concordance ------------------------------
# rtqpcr_rs is built by the ch5-fig-rtqpcr-corr chunk; recomputed here if absent
# so this chunk is order-independent.
if (!exists("rtqpcr_rs"))
  rtqpcr_rs <- c(0.2441176, 0.4352941, 0.5617647, 0.5588235, 0.6500000)

# EXACT Spearman (AS 89) throughout, at n = 5 the asymptotic approximation is
# anticonservative (it returns p = 0.037 where the exact test returns 0.083).
ct_stage <- suppressWarnings(cor.test(stage_ids, trh_qpcr$log2FC, method = "spearman"))
ct_conc  <- suppressWarnings(cor.test(trh_qpcr$log2FC, rtqpcr_rs, method = "spearman"))

##, 6. Figure ----------------------------------------------------------------
trh_plot <- rbind(
  # SE is available for the RT-qPCR series: three independent preparations
  # per arm per stage in the raw Ct workbook.
  data.frame(stage = trh_qpcr$stage,  log2FC = trh_qpcr$log2FC,
             SE = trh_qpcr$SE, assay = "RT-qPCR"),
  data.frame(stage = trh_array$stage, log2FC = trh_array$log2FC,
             SE = trh_array$SE,       assay = "Microarray"))
trh_plot$assay <- factor(trh_plot$assay, levels = c("RT-qPCR", "Microarray"))

pal_assay <- c("RT-qPCR" = PAL_SUBUNIT[["Trh"]], "Microarray" = PAL_INK[["fit"]])

p_trh <- ggplot(trh_plot, aes(x = stage, y = log2FC,
                              colour = assay, group = assay)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = PAL_INK[["annotation"]], linewidth = 0.4) +
  geom_errorbar(aes(ymin = log2FC - SE, ymax = log2FC + SE),
                width = 0.12, linewidth = 0.5, na.rm = TRUE, alpha = 0.75) +
  geom_line(aes(linewidth = assay, alpha = assay)) +
  geom_point(aes(shape = assay, alpha = assay), size = 2.6) +
  scale_colour_manual(values = pal_assay, name = NULL) +
  scale_shape_manual(values = c("RT-qPCR" = 18, "Microarray" = 16), name = NULL) +
  scale_linewidth_manual(values = c("RT-qPCR" = 0.9, "Microarray" = 0.6),
                         guide = "none") +
  scale_alpha_manual(values = c("RT-qPCR" = 1, "Microarray" = 0.6), guide = "none") +
  scale_x_continuous(breaks = stage_ids) +
  labs(x = "Embryonic stage",
       y = expression(italic("trh")~log[2]~"fold change (RNAi/wt)")) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.line = element_line(colour = "black"),
        legend.position = c(0.16, 0.16),
        legend.background = element_rect(fill = "white", colour = NA))

ggsave(file.path(FIGURES_DIR, "fig5_3_trh_expression.png"),
       plot = p_trh, width = 6, height = 3.2, dpi = 300, bg = "white")

cat(sprintf("trh RT-qPCR: monotonic, rho = %.2f (exact p = %.4f); st15 = %.2f (%.1f-fold down)\n",
            ct_stage$estimate, ct_stage$p.value,
            trh_qpcr$log2FC[5], 1 / 2^trh_qpcr$log2FC[5]))
cat(sprintf("array detection floor at st13: |log2FC| >= %.2f (%.0f-fold)\n",
            mde_log2, mde_fold))
cat(sprintf("depletion vs concordance: rho = %.2f, exact p = %.3f (n.s. at n = 5)\n",
            ct_conc$estimate, ct_conc$p.value))
```

---

```{r ch5-fig-cluster-dendrogram, include=TRUE, eval=TRUE}
library(png)
library(magick)
# INPUT LOCATION: this raw dendrogram is a hand-made input, not a pipeline
# output, so it lives with the other Chapter 5 inputs rather than in the
# figures directory.
src <- file.path(CH5_INPUT_DIR, "cluster_dendrogram_raw.png")
if (!file.exists(src))
  stop("missing input ", src, " (expected in data/chapter5/)")
## ============================================================
## STEP 1: add a white strip BELOW the tree (native res preserved)
## ============================================================
pad <- 70      # TUNE: height in px of the white strip added at the bottom

img  <- image_read(src)
info <- image_info(img)
img  <- image_extent(img,
                     geometry_area(width = info$width, height = info$height + pad),
                     gravity = "north", color = "white")   # tree pinned to top, white added at BOTTOM
image_write(img, file.path(CH5_INPUT_DIR, "tmp_padded.png"))

ras <- readPNG(file.path(CH5_INPUT_DIR, "tmp_padded.png"))
h <- dim(ras)[1]      # native width/height of the PADDED image -> no rescale
w <- dim(ras)[2]

# OUTPUT NAME: this chunk produces the cluster dendrogram figure referenced
# in the text, written at 445x477 px against this input's 445x407, a
# difference of exactly `pad`.
dendro_out <- file.path(FIGURES_DIR, "figH_13_cluster_dendrogram_high_res.png")
png(dendro_out,
    width = w, height = h, units = "px", type = "cairo-png")
par(mar = c(0,0,0,0))
plot(c(0,w), c(0,h), type = "n", xaxs = "i", yaxs = "i", axes = FALSE,
     xlab = "", ylab = "")
rasterImage(ras, 0, 0, w, h)
green <- PAL_HETERO[["Trh/Tgo"]]

## ============================================================
## HEIGHT / VERTICAL POSITION OF THE COMB
## y is measured from the BOTTOM. The comb lives in the white
## strip, so y_base is small (inside the `pad` band).
## ============================================================
y_base   <- 55     # TUNE: y of the horizontal bracket line (within the white strip)
tick_len <- 18     # TUNE: height of the vertical end-ticks (comb teeth), pointing UP toward leaves
label_dy <- 15     # TUNE: how far BELOW the line the stage label sits
## ============================================================

## ============================================================
## HORIZONTAL EXTENT OF EACH SEGMENT (manual pixel x; no coupling)
## ============================================================
segs <- list(
  c(0.11 * w, 0.2825 * w, "St12"),   # TUNE
  c(0.2825 * w, 0.4625 * w, "St13"),   # TUNE
  c(0.4625 * w, 0.6425 * w, "St11"),   # TUNE
  c(0.6425 * w, 0.8225 * w, "St14"),   # TUNE
  c(0.8225 * w, 0.997 * w, "St15")    # TUNE
)
## ============================================================

for (s in segs) {
  x0 <- as.numeric(s[1]); x1 <- as.numeric(s[2])
  segments(x0, y_base, x1, y_base,            col = green, lwd = 4)  # horizontal bar
  segments(x0, y_base, x0, y_base + tick_len, col = green, lwd = 4)  # left tooth (up)
  segments(x1, y_base, x1, y_base + tick_len, col = green, lwd = 4)  # right tooth (up)
  text(mean(c(x0, x1)), y_base - label_dy, labels = s[3],
       col = green, font = 2, cex = 2)
}
dev.off()
```

---

```{r ch5-fig-rtqpcr-corr, include=TRUE, eval=TRUE}
# Fig 5.5: microarray vs

suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
})

CH5_INPUT_DIR <- file.path(PROJECT_DIR, "data", "chapter5")
LIMMA_DIR     <- CH5_INPUT_DIR   # holds R{11..15}.vsn.limma.fdr.txt.gz
stage_ids     <- 11:15

# The 16 validation genes, identified by microarray probe (TargetID). This is
# the single source of truth linking the RT-qPCR gene names to the limma probe
# rows; it is stable across stages.
probe_map <- c(
  "BicC"     = "CG4824-RA",  "PGRP-SB1" = "CG9681-RA",  "Cyp6a19" = "CG10243-RA",
  "dhd"      = "CG4193-RA",  "AsnS"     = "CG33486-RA", "Ect3"    = "CG3132-RA",
  "Phk-3"    = "CG9358-RA",  "Cpr51A"   = "CG10112-RA", "trh"     = "CG6883-RA",
  "Cyp12a4"  = "CG6042-RA",  "Cyp309a2" = "CG18559-RA", "ninaD"   = "CG31783-RA",
  "CG14102"  = "CG14102-RA", "Cdc42"    = "CG12530-RB", "Sli"     = "CG8355-RC",
  "Ir7b"     = "CG15326-RA"
)

read_marray_stage <- function(st) {
  f <- file.path(LIMMA_DIR, sprintf("R%d.vsn.limma.fdr.txt.gz", st))
  stopifnot(file.exists(f))
  # First line is an annotation comment; real header is line 2.
  lim <- read.delim(f, skip = 1, check.names = FALSE, stringsAsFactors = FALSE)
  stopifnot(all(c("TargetID", "logFC", "t") %in% names(lim)))
  idx <- match(probe_map, lim$TargetID)
  stopifnot(!any(is.na(idx)))            # every probe must resolve
  data.frame(
    gene       = names(probe_map),
    microarray = as.numeric(lim$logFC[idx]),
    marray_se  = as.numeric(lim$logFC[idx]) / as.numeric(lim$t[idx]),
    row.names  = NULL, stringsAsFactors = FALSE
  )
}

read_rtqpcr_stage <- function(st) {
  f <- file.path(CH5_INPUT_DIR, sprintf("st%d_validation_unstacked.csv", st))
  stopifnot(file.exists(f))
  d <- read.csv(f, row.names = 1, check.names = FALSE, fileEncoding = "UTF-8-BOM")
  names(d)[names(d) %in% c("RT-qPCR", "RT.qPCR")] <- "RTqPCR"
  data.frame(gene = rownames(d), RTqPCR = as.numeric(d$RTqPCR),
             row.names = NULL, stringsAsFactors = FALSE)
}

build_panel <- function(st, panel_tag) {
  d <- merge(read_marray_stage(st), read_rtqpcr_stage(st), by = "gene")
  stopifnot(nrow(d) == 16L)              # all 16 genes, no silent loss
  ct <- suppressWarnings(cor.test(d$microarray, d$RTqPCR, method = "spearman"))
  rs <- unname(ct$estimate); pv <- ct$p.value

  # Significant stages (p < 0.05) get a blue fitted line; non-significant
  # stages keep a neutral grey line. Blue is the conventional "fitted model"
  # colour and is reserved here for significance; red is not used (in this
  # thesis red denotes the trh data trace in Fig 5.3, not a regression fit).
  sig       <- pv < 0.05
  line_col  <- if (sig) PAL_INK[["fit"]] else PAL_INK[["neutral"]]

  # Annotation placement: stages 11 and 12 have a data point in the top-left
  # corner, so their R_s / p labels go to the BOTTOM-left; all other stages
  # keep the labels top-left where the corner is clear.
  if (st %in% c(11L, 12L)) {
    lab_x <- Inf;  lab_y <- -Inf
    h_rs  <- 1.12; v_rs  <- -3.4
    h_p   <- 1.12; v_p   <- -1.7   # bottom-right, stacked upward
  } else {
    lab_x <- -Inf; lab_y <-  Inf
    h_rs  <- -0.12; v_rs <-  1.5
    h_p   <- -0.12; v_p  <-  3.2   # top-left, stacked downward
  }

  list(
    stage = st, n = nrow(d), Rs = rs, p = pv, data = d, sig = sig,
    plot  = ggplot(d, aes(x = RTqPCR, y = microarray)) +
      geom_point(size = 2) +
      stat_smooth(method = "lm", formula = y ~ x,
                  se = FALSE, colour = line_col, linewidth = 0.6) +
      ggtitle(paste("Stage", st)) +
      labs(tag = panel_tag,
           x   = expression("RT-qPCR " * log[2] * " FC"),
           y   = expression("microarray " * log[2] * " FC")) +
      annotate("text", x = lab_x, y = lab_y, hjust = h_rs, vjust = v_rs,
               parse = TRUE, size = 3.5,
               label = sprintf("italic(R)[s] == %.2f", rs)) +
      annotate("text", x = lab_x, y = lab_y, hjust = h_p, vjust = v_p,
               parse = TRUE, size = 3.5,
               label = sprintf("italic(p) == %.3f", pv)) +
      theme_classic() +
      theme(plot.title = element_text(hjust = 0.5, face = "bold"),
            plot.tag   = element_text(size = 14))
  )
}

panels <- Map(build_panel, stage_ids, LETTERS[seq_along(stage_ids)])

rtqpcr_stats <- data.frame(
  stage = paste("Stage", stage_ids),
  Rs    = vapply(panels, `[[`, numeric(1), "Rs"),
  p     = vapply(panels, `[[`, numeric(1), "p")
)
print(rtqpcr_stats)

# 3 rows x 2 cols (panels A-E + one empty slot) -> single fig5_4.png
fig5_5 <- (panels[[1]]$plot | panels[[2]]$plot) /
          (panels[[3]]$plot | panels[[4]]$plot) /
          (panels[[5]]$plot | plot_spacer())
ggsave(file.path(FIGURES_DIR, "fig5_4.png"),
       plot = fig5_5, width = 8, height = 11, dpi = 300)
print(fig5_5)
```

---

```{r ch5-microarray-fisher-concordance, include=TRUE, eval=TRUE}
# ---- Section 5.4: rank concordance analysis, not Venn ----
# Inputs: Limma per-stage outputs (st13, st14), Fisher 2022 modERN
# trh-RNAi DESeq2 table. Outputs: fig5_5_concordance.png,
# merged_microarray_fisher_st{13,14}.csv, microarray_validated_st{13,14}.csv,
# values in results/reported_values.tsv. Re-running overwrites all outputs.
 
suppressPackageStartupMessages({
  library(readr); library(readxl); library(dplyr)
  library(tidyr); library(ggplot2); library(patchwork); library(hexbin)
})
 
# ---- Paths ----
P_LIMMA_ST13 <- file.path(CH5_INPUT_DIR, "R13.vsn.limma.fdr.txt.gz")
P_LIMMA_ST14 <- file.path(CH5_INPUT_DIR, "R14.vsn.limma.fdr.txt.gz")
P_FISHER_XLS <- file.path(CH5_INPUT_DIR,
                          "modERN_Resource_RNAi_Table_S43_trh_GENETICS-2022-305822.xlsx")
for (P in c(P_LIMMA_ST13, P_LIMMA_ST14, P_FISHER_XLS)) {
  if (!file.exists(P))
    stop("Missing input for concordance analysis: ", P)
}
 
# ---- 1. Parse Limma microarray output (skip '#' comment header) ----
read_limma <- function(path) {
  raw <- readLines(path); raw <- raw[!grepl("^#", raw)]
  read.delim(text = paste(raw, collapse = "\n"),
             header = TRUE, sep = "\t",
             stringsAsFactors = FALSE, check.names = FALSE)
}
 
clean_limma <- function(d) {
  # probes with multi-FBgn assignment are expanded to one row per FBgn
  d$logFC <- as.numeric(d$logFC); d$P.Value <- as.numeric(d$P.Value)
  d$adj.P.Val <- as.numeric(d$adj.P.Val)
  d <- d[grepl("FBgn\\d+", d$FBgn), , drop = FALSE]
 
  fb_list  <- strsplit(d$FBgn, ",")
  sym_list <- strsplit(sub(",$", "", d$GeneName), ",")
  lens     <- vapply(fb_list, length, integer(1))
  d_exp    <- d[rep(seq_len(nrow(d)), lens), , drop = FALSE]
  d_exp$FBgn <- unlist(fb_list)
  d_exp$symbol <- unlist(lapply(seq_along(fb_list), function(i) {
    s <- sym_list[[i]]; n <- lens[i]
    if (length(s) == n) s else rep(s[1], n)
  }))
 
  # gene-level aggregation: keep most-significant probe per FBgn
  d_exp <- d_exp[order(d_exp$adj.P.Val), , drop = FALSE]
  d_exp <- d_exp[!duplicated(d_exp$FBgn), , drop = FALSE]
  d_exp[, c("FBgn", "symbol", "logFC", "P.Value", "adj.P.Val")]
}
 
l13 <- clean_limma(read_limma(P_LIMMA_ST13))
l14 <- clean_limma(read_limma(P_LIMMA_ST14))
 
# ---- 2. Parse Fisher DESeq2 sheet and resolve symbols -> FBgn ----
fis <- read_excel(P_FISHER_XLS, sheet = "DESeq2 CSV") |>
  dplyr::rename(symbol = `gene name`,
         log2FC_fisher_0_1.5  = `log2FoldChange.trh_0.1.5`,
         padj_fisher_0_1.5    = `padj.trh_0.1.5`,
         log2FC_fisher_6_8    = `log2FC_6.8`,
         padj_fisher_6_8      = `padj_6.8`,
         log2FC_fisher_16_18  = `log2FC_16.18`,
         padj_fisher_16_18    = `padj_16.18`) |>
  dplyr::filter(!is.na(symbol))
 
sym2fbgn <- setNames(l13$FBgn, l13$symbol)
fis$FBgn <- sym2fbgn[fis$symbol]
n_fisher_total  <- nrow(fis)
n_fisher_mapped <- sum(!is.na(fis$FBgn))
 
fis <- fis |> dplyr::filter(!is.na(FBgn)) |>
  dplyr::group_by(FBgn) |> dplyr::arrange(padj_fisher_6_8) |> dplyr::slice(1) |> ungroup()
 
# ---- 3. Merge per-stage ----
merge_stage <- function(limma_d, fis_d, stage_lbl) {
  limma_d |>
    dplyr::rename(logFC_microarray = logFC,
           P_microarray     = P.Value,
           padj_microarray  = adj.P.Val) |>
    inner_join(fis_d |> dplyr::select(FBgn,
                               log2FC_fisher_0_1.5, padj_fisher_0_1.5,
                               log2FC_fisher_6_8,   padj_fisher_6_8,
                               log2FC_fisher_16_18, padj_fisher_16_18),
               by = "FBgn") |>
    dplyr::mutate(stage = stage_lbl)
}
m13 <- merge_stage(l13, fis, "st13")
m14 <- merge_stage(l14, fis, "st14")
 
write_csv(m13, file.path(CH5_OUT_DIR, "merged_microarray_fisher_st13.csv"))
write_csv(m14, file.path(CH5_OUT_DIR, "merged_microarray_fisher_st14.csv"))
 
# ---- 4. Concordance statistics ----
windows <- c(early = "log2FC_fisher_0_1.5",
             mid   = "log2FC_fisher_6_8",
             late  = "log2FC_fisher_16_18")
 
concord_one <- function(d, ywin) {
  s <- d[, c("logFC_microarray", ywin)] |> drop_na()
  sp <- suppressWarnings(cor.test(s$logFC_microarray, s[[ywin]],
                                  method = "spearman", exact = FALSE))
  data.frame(n = nrow(s),
             rho = unname(sp$estimate),
             p = sp$p.value,
             sign_concord = mean(sign(s$logFC_microarray) == sign(s[[ywin]])))
}
 
stat13 <- lapply(windows, concord_one, d = m13) |> setNames(names(windows))
stat14 <- lapply(windows, concord_one, d = m14) |> setNames(names(windows))
 
# ---- 5. Validated subsets: microarray |logFC|>=0.49, padj<0.55, ----
# ---- Fisher 6-8h |log2FC|>=0.49, padj<0.05 ----
mar_st13_set <- m13 |>
  dplyr::filter(abs(logFC_microarray) >= 0.49, padj_microarray < 0.55) |>
  pull(FBgn)
mar_st14_set <- m14 |>
  dplyr::filter(abs(logFC_microarray) >= 0.49, padj_microarray < 0.55) |>
  pull(FBgn)
fisher_set <- m13 |>
  dplyr::filter(abs(log2FC_fisher_6_8) >= 0.49, padj_fisher_6_8 < 0.05) |>
  pull(FBgn)
 
valid_st13 <- intersect(mar_st13_set, fisher_set)
valid_st14 <- intersect(mar_st14_set, fisher_set)
 
writeLines(c("FBgnID", valid_st13),
           file.path(CH5_OUT_DIR, "microarray_validated_st13.csv"))
writeLines(c("FBgnID", valid_st14),
           file.path(CH5_OUT_DIR, "microarray_validated_st14.csv"))
 
sub13 <- m13 |> dplyr::filter(FBgn %in% valid_st13)
sub14 <- m14 |> dplyr::filter(FBgn %in% valid_st14)
sign_c_valid_st13 <- mean(sign(sub13$logFC_microarray) ==
                          sign(sub13$log2FC_fisher_6_8))
sign_c_valid_st14 <- mean(sign(sub14$logFC_microarray) ==
                          sign(sub14$log2FC_fisher_6_8))
 
cat(sprintf("Validated subset: st13=%d, st14=%d, union=%d, intersection=%d\n",
            length(valid_st13), length(valid_st14),
            length(union(valid_st13, valid_st14)),
            length(intersect(valid_st13, valid_st14))))
cat(sprintf("Genome-wide rho st13 vs Fisher 6–8h: %.3f  (n=%d)\n",
            stat13$mid$rho, stat13$mid$n))
cat(sprintf("Genome-wide rho st14 vs Fisher 6–8h: %.3f  (n=%d)\n",
            stat14$mid$rho, stat14$mid$n))
 
# ---- 6. Figure: 6-panel concordance scatter ----
# Per-window x-scaling; shared log10 density colourbar; concordant-quadrant
# tint; "OLS trend" and "Validated subset" legend keys; role-named headers.
# ASCII hyphens in plotted text. Requires hexbin (loaded in ch5-setup chunk).

# Per-window symmetric x-limits: cover ~99.5% of |FC| and enclose every
# validated (red) point, so no highlighted DEG is clipped off-panel (the
# validated genes have larger FC than the bulk by construction). Round up to 0.1.
xlim_for <- function(col) {
  allv <- c(m13[[col]], m14[[col]])
  red  <- c(m13[[col]][m13$FBgn %in% valid_st13],
            m14[[col]][m14$FBgn %in% valid_st14])
  need <- max(quantile(abs(allv), 0.995, na.rm = TRUE),
              max(abs(red), na.rm = TRUE))
  l <- ceiling(need * 10) / 10
  c(-l, l)
}
.wins  <- c("log2FC_fisher_0_1.5", "log2FC_fisher_6_8", "log2FC_fisher_16_18")
.xlims <- sapply(.wins, xlim_for)

# Shared fill limits: global max hexbin count across all six panels. Must mirror
# the panel binning exactly (visible window only, xbins = 34) or the colourbar
# ceiling will not match the darkest hexagon actually drawn.
.gmax <- 0
.yl <- c(-1.5, 1.5)
for (i in seq_along(.wins)) {
  ww <- .wins[i]; lim <- .xlims[, i]
  for (mm in list(m13, m14)) {
    dd <- mm[, c(ww, "logFC_microarray")]; dd <- dd[stats::complete.cases(dd), ]
    dd <- dd[dd[[ww]] >= lim[1] & dd[[ww]] <= lim[2] &
             dd$logFC_microarray >= .yl[1] & dd$logFC_microarray <= .yl[2], ]
    hbc <- hexbin::hexbin(dd[[ww]], dd$logFC_microarray, xbins = 34)@count
    .gmax <- max(.gmax, hbc)
  }
}

make_panel <- function(m, ywin, xlim, highlight_fbgn, xlab, show_y, show_x,
                       label_pos = "topleft") {
  d <- m |> dplyr::select(FBgn, logFC_microarray, all_of(ywin)) |> drop_na()
  names(d)[3] <- "fisher_fc"
  d$highlight <- d$FBgn %in% highlight_fbgn
  rho <- suppressWarnings(cor(d$logFC_microarray, d$fisher_fc,
                              method = "spearman"))
  sign_c <- mean(sign(d$logFC_microarray) == sign(d$fisher_fc))
  # blank line before "n =" opens vertical space; hjust = 0 (below) keeps all
  # three lines flush-left at the same edge.
  lbl <- sprintf("Spearman rho = %.3f\nsign concord. = %.0f%%\n\nn = %s",
                 rho, 100 * sign_c, format(nrow(d), big.mark = ""))
  yl <- c(-1.5, 1.5)
  # Bin only the data inside the display window, so hexagon size is tied to the
  # visible range and is not magnified by coord_cartesian zoom (binning the full
  # +-3 range then zooming to +-1 stretches the hexagons). Stats, trend line and
  # highlighted points still use the full data (d).
  dv <- d |> dplyr::filter(fisher_fc >= xlim[1], fisher_fc <= xlim[2],
                    logFC_microarray >= yl[1], logFC_microarray <= yl[2])
  # Text inset in data coordinates, kept off the spines. "topleft" (stage 13):
  # anchored 5% in from the left, 4% down from the top, left-aligned. "bottomright"
  # (stage 14, where the empty quadrant is bottom-right): anchored 5% in from the
  # right, 4% up from the bottom, right-aligned so the block grows leftward into
  # the whitespace without overflowing the panel.
  if (label_pos == "bottomright") {
    tx <- xlim[2] - diff(xlim) * 0.05; thj <- 1
    ty <- yl[1]   + diff(yl)   * 0.04; tvj <- 0
  } else {
    tx <- xlim[1] + diff(xlim) * 0.05; thj <- 0
    ty <- yl[2]   - diff(yl)   * 0.04; tvj <- 1
  }
  ggplot(d, aes(fisher_fc, logFC_microarray)) +
    # tint the two concordant (agreement) quadrants: both up / both down
    annotate("rect", xmin = 0, xmax = xlim[2], ymin = 0, ymax = yl[2],
             fill = PAL_INK[["highlight_band"]], alpha = 0.05) +
    annotate("rect", xmin = xlim[1], xmax = 0, ymin = yl[1], ymax = 0,
             fill = PAL_INK[["highlight_band"]], alpha = 0.05) +
    geom_hline(yintercept = 0, linewidth = 0.25, colour = PAL_INK[["neutral"]]) +
    geom_vline(xintercept = 0, linewidth = 0.25, colour = PAL_INK[["neutral"]]) +
    geom_hex(data = dv, aes(fill = after_stat(count)), bins = 34) +
    scale_fill_gradient(low = PAL_SEQ(9)[9], high = PAL_SEQ(9)[1], trans = "log10",
                        name = "Genes per\nhexbin", limits = c(1, .gmax),
                        breaks = c(1, 10, 100), oob = scales::squish) +
    geom_smooth(aes(colour = "OLS trend"), method = "lm", se = FALSE,
                linewidth = 0.6, linetype = "dashed") +
    scale_colour_manual(values = c("OLS trend" = PAL_INK[["fit"]]), name = NULL) +
    geom_point(data = subset(d, highlight),
               aes(shape = "Validated subset"),
               colour = PAL_INK[["emphasis"]], size = 1.3, alpha = 0.9) +
    scale_shape_manual(values = c("Validated subset" = 16), name = NULL) +
    annotate("text", x = tx, y = ty, hjust = thj, vjust = tvj,
             label = lbl, size = 2.5, lineheight = 0.95) +
    scale_x_continuous(expand = expansion(mult = 0.03)) +
    scale_y_continuous(expand = expansion(mult = 0.03)) +
    coord_cartesian(xlim = xlim, ylim = yl) +
    labs(x = if (show_x) xlab else NULL,
         y = if (show_y) "Microarray log2 FC" else NULL) +
    theme_classic(base_size = 9) +
    theme(axis.title = element_text(size = 8),
          axis.text.y = if (show_y) element_text() else element_blank(),
          plot.margin = margin(2, 4, 2, 2))
}

# Column headers name the biological role; row headers name the stage.
.xlabs <- c("Fisher 0-1.5 h log2 FC", "Fisher 6-8 h log2 FC",
            "Fisher 16-18 h log2 FC")
.csub  <- c("Fisher 0-1.5 h", "Fisher 6-8 h", "Fisher 16-18 h")
.ctitle<- c("Negative control", "Temporal proxy (closest)",
            "Developmental endpoint")
mk_hdr <- function(t, s)
  ggplot() +
    annotate("text", x = 0, y = 0.62, label = t, fontface = "bold", size = 3.2) +
    annotate("text", x = 0, y = 0.15, label = s, size = 2.7, colour = PAL_INK[["annotation"]]) +
    theme_void() + coord_cartesian(ylim = c(0, 1))
mk_row <- function(t)
  ggplot() + annotate("text", x = 0, y = 0, label = t, fontface = "bold",
                      size = 3, angle = 90) + theme_void()
.hdrs  <- Map(mk_hdr, .ctitle, .csub)

legend_theme <- theme(legend.position = "right",
                      legend.key.size = unit(0.36, "cm"),
                      legend.title = element_text(size = 7.5),
                      legend.text  = element_text(size = 7))
guide_spec <- guides(
  fill   = guide_colourbar(order = 1, barheight = unit(1.6, "cm")),
  colour = guide_legend(order = 2, override.aes = list(linetype = "dashed")),
  shape  = guide_legend(order = 3))

.P <- list()
for (i in 1:3) {
  .P[[paste0("t13_", i)]] <-
    make_panel(m13, .wins[i], .xlims[, i], valid_st13, .xlabs[i],
               show_y = (i == 1), show_x = FALSE) + legend_theme + guide_spec
  .P[[paste0("t14_", i)]] <-
    make_panel(m14, .wins[i], .xlims[, i], valid_st14, .xlabs[i],
               show_y = (i == 1), show_x = TRUE,
               label_pos = "bottomright") + legend_theme + guide_spec
}

.design <- "
.AAABBBCCC
DEEEFFFGGG
HIIIJJJKKK
"
p_grid <-
  plot_spacer() + .hdrs[[1]] + .hdrs[[2]] + .hdrs[[3]] +
  mk_row("Stage 13") + .P[["t13_1"]] + .P[["t13_2"]] + .P[["t13_3"]] +
  mk_row("Stage 14") + .P[["t14_1"]] + .P[["t14_2"]] + .P[["t14_3"]] +
  plot_layout(design = .design,
              widths  = c(0.10, 1, 1, 1, 1, 1, 1, 1, 1, 1),
              heights = c(0.14, 1, 1), guides = "collect") &
  theme(legend.position = "right")

ggsave(file.path(FIGURES_DIR, "fig5_5_concordance.png"),
       plot = p_grid, width = 11, height = 7.2, units = "in", dpi = 300)
 

# ---- 8. Expose objects to the rest of Chapter 5 ----
# Maintains compatibility with downstream chunks that expect these names:
marray_st13_genes      <- valid_st13
marray_st14_genes      <- valid_st14
marray_validated_total <- length(union(valid_st13, valid_st14))
marray_in_fisher_st13  <- length(valid_st13)
marray_in_fisher_st14  <- length(valid_st14)
# not "by construction": the validated set is filtered on the Fisher 6-8 h
# window (|log2FC_fisher_6_8| >= 0.49 & padj_fisher_6_8 < 0.05), whereas the
# direct-target derivation downstream uses the COMBINED Fisher DEG list.
# FBgn0029128 (tyn) clears the former and not the latter, so the two sets
# differ by one gene. Compute it; never assume the identity.
marray_validated_genes <- union(valid_st13, valid_st14)
marray_in_fisher_total <- sum(marray_validated_genes %in% fisher_degs)
stopifnot(marray_in_fisher_total <= marray_validated_total)
marray_source          <- "live"
```

```{r ch5-chip-annotate-genes, include=TRUE, eval=TRUE}
# ---- Annotate Trh/Tgo peaks to nearest gene ----
# Uses ChIPseeker with TSS region of +/- 2kb (matches Chapter 3 convention).
# A peak's "gene assignment" is the FlyBase ID of its nearest TSS within
# the window; peaks whose nearest TSS is beyond the window are dropped.

tss_window <- c(-2000L, 2000L)

peakAnno_trh <- annotatePeak(
  Trh_Tgo_joint_filt,
  TxDb       = txdb,
  tssRegion  = tss_window,
  level      = "gene",
  annoDb     = "org.Dm.eg.db",
  verbose    = FALSE
)

peakAnno_df <- as.data.frame(peakAnno_trh)

# Keep only peaks whose nearest TSS is within +/-2kb
peak_genes_in_window <- subset(peakAnno_df, abs(distanceToTSS) <= 2000L)

# FlyBase ID column from ChIPseeker is typically "geneId" (Ensembl) and
# "ENSEMBL" or "FLYBASE" in annoDb-augmented data. Locate flexibly.
fb_col <- {
  if      ("FLYBASE" %in% names(peak_genes_in_window)) "FLYBASE"
  else if ("flybase" %in% names(peak_genes_in_window)) "flybase"
  else if ("ENSEMBL" %in% names(peak_genes_in_window)) "ENSEMBL"
  else if ("geneId"  %in% names(peak_genes_in_window)) "geneId"
  else stop("No FlyBase/Ensembl gene-ID column in annotated peaks. ",
            "Available columns: ",
            paste(names(peak_genes_in_window), collapse = ", "))
}

trh_chip_genes <- unique(peak_genes_in_window[[fb_col]])
trh_chip_genes <- trh_chip_genes[!is.na(trh_chip_genes) &
                                 nzchar(trh_chip_genes)]

cat("Total Trh/Tgo peaks:                ", length(Trh_Tgo_joint_filt), "\n")
cat("Peaks within +/-2kb of a TSS:       ", nrow(peak_genes_in_window), "\n")
cat("Unique genes near Trh/Tgo peaks:    ", length(trh_chip_genes), "\n")
```

---

```{r ch5-direct-targets, include=TRUE, eval=TRUE}
# ---- Section 5.5: Define direct-target set ----
# Direct target = gene whose TSS is within +/-2kb of a Trh/Tgo peak and
# is differentially expressed in the Fisher trh-RNAi RNA-seq dataset.

direct_targets_trh <- intersect(trh_chip_genes, fisher_degs)

cat("Direct Trh/Tgo targets (ChIP n DEG):", length(direct_targets_trh), "\n")

# Hypergeometric test: are direct targets enriched relative to background?
# Null: among all dm6 genes, the proportion that are DEGs is the same
# whether or not they have a nearby Trh/Tgo peak.
all_dm6_genes <- keys(org.Dm.eg.db, keytype = "FLYBASE")

q_overlap   <- length(intersect(trh_chip_genes, fisher_degs))
m_in_chip   <- length(intersect(trh_chip_genes, all_dm6_genes))
n_not_chip  <- length(all_dm6_genes) - m_in_chip
k_in_fisher <- length(intersect(fisher_degs, all_dm6_genes))

# Probability of observing >= q_overlap by chance:
p_hyper_trh <- phyper(q_overlap - 1L, k_in_fisher,
                      length(all_dm6_genes) - k_in_fisher,
                      m_in_chip, lower.tail = FALSE)
fold_enrich_trh <- (q_overlap / m_in_chip) /
                   (k_in_fisher / length(all_dm6_genes))

cat(sprintf("Hypergeometric p (direct target enrichment): %.3g\n",
            p_hyper_trh))
cat(sprintf("Fold enrichment over background:             %.2f\n",
            fold_enrich_trh))

# Restrict draw and DEG set to the 11120-gene merged microarray/RNA-seq
# table (the testable universe). trh_chip_genes comes from all annotated
# dm6 genes, so part of it falls outside this universe.
univ_genes  <- unique(m13$FBgn[!is.na(m13$FBgn)])
chip_in_u   <- intersect(trh_chip_genes, univ_genes)
degs_in_u   <- intersect(fisher_degs,    univ_genes)
ov_in_u     <- intersect(chip_in_u,      degs_in_u)

u_N <- length(univ_genes); u_n <- length(chip_in_u)
u_K <- length(degs_in_u);  u_k <- length(ov_in_u)
stopifnot(u_n <= u_N, u_K <= u_N, u_k <= min(u_n, u_K))

u_exp  <- u_n * u_K / u_N
u_fold <- u_k / u_exp
u_pdep <- phyper(u_k, u_K, u_N - u_K, u_n, lower.tail = TRUE)
u_ft   <- fisher.test(matrix(c(u_k, u_n - u_k,
                               u_K - u_k, u_N - u_K - u_n + u_k), 2, 2))
# one-sided power to detect a true two-fold enrichment at this draw size
u_crit <- qhyper(0.95, u_K, u_N - u_K, u_n)
u_pow2 <- sum(dhyper(seq(u_crit + 1L, u_n), round(2 * u_K),
                     u_N - round(2 * u_K), u_n))

cat(sprintf(paste0("Universe-restricted: N=%d K=%d n=%d (of %d; %d outside) ",
                   "k=%d exp=%.1f fold=%.2f p_dep=%.3f OR=%.2f [%.2f,%.2f] ",
                   "power(2x)=%.3f\n"),
            u_N, u_K, u_n, length(trh_chip_genes),
            length(trh_chip_genes) - u_n, u_k, u_exp, u_fold, u_pdep,
            u_ft$estimate, u_ft$conf.int[1], u_ft$conf.int[2], u_pow2))

# Save the direct-target list
write.csv(
  data.frame(FBgn = direct_targets_trh),
  file.path(CH5_OUT_DIR, "direct_targets_TrhTgo.csv"),
  row.names = FALSE
)
```

```{r ch5-direct-targets-table, include=TRUE, eval=TRUE}
# Per-gene direct-target table for Section 5.5.
# Joins 21-gene direct-target list (FBgn, from ch5-direct-targets chunk),
# peak annotation table (signal, width, dist to TSS), and Fisher 2023
# RNA-seq per-gene fold-change table (SYMBOL).
# Fisher symbols mapped to FBgn via org.Dm.eg.db at join; unmappable
# symbols printed. Output: direct_targets_with_intensity_and_fc.csv.

# Pre-flight
required <- c("direct_targets_trh", "peakAnno_df", "Trh_Tgo_joint",
              "fb_col")
missing  <- setdiff(required, ls(envir = .GlobalEnv))
if (length(missing) > 0L) {
  stop("Missing prerequisites: ", paste(missing, collapse = ", "))
}

suppressPackageStartupMessages({
  library(org.Dm.eg.db)
  library(AnnotationDbi)
})

# ---- Step 1: Get peak intensity for each direct target ----
annotated <- peakAnno_df[peakAnno_df[[fb_col]] %in% direct_targets_trh, ]
stopifnot(nrow(annotated) > 0L)

annotated_gr <- GRanges(
  seqnames = annotated$seqnames,
  ranges   = IRanges(start = annotated$start, end = annotated$end),
  strand   = "*"
)

ov <- findOverlaps(annotated_gr, Trh_Tgo_joint, type = "equal")
annotated$signalValue <- NA_real_
annotated$signalValue[queryHits(ov)] <-
  Trh_Tgo_joint$signalValue[subjectHits(ov)]

# Pick highest-signal peak per gene
per_gene_peaks <- by(annotated,
                     INDICES = annotated[[fb_col]],
                     FUN = function(df) {
                       df[which.max(df$signalValue), , drop = FALSE]
                     })
per_gene_peaks <- do.call(rbind, per_gene_peaks)
per_gene_peaks <- per_gene_peaks[, c(fb_col, "SYMBOL", "width",
                                     "signalValue", "distanceToTSS")]
names(per_gene_peaks)[1] <- "gene_id"

# ---- Step 2: Load Fisher 2023 per-gene table ----
# the file is tab-separated and uses decimal commas
# (European locale). Both must be specified to read.delim or numeric
# columns will be parsed as strings.

P_FISHER_PERGENE <- file.path(
  REFERENCE_LISTS,
  "Fisher2023_per_gene_fold_changes.csv"
)

if (!file.exists(P_FISHER_PERGENE)) {
  stop("Fisher per-gene table not found at: ", P_FISHER_PERGENE,
       "\nDownload from the journal supplementary materials and save ",
       "as a TSV (tab-separated, decimal commas) at this path.")
}

# Read as TSV with decimal comma. The columns log2FC_6.8 and padj_6.8
# refer to Fisher 2023's stage-window 6-8; log2FC_16.18 and padj_16.18
# refer to stage-window 16-18.
fisher_raw <- read.delim(
  P_FISHER_PERGENE,
  header = TRUE,
  sep = "\t",
  dec = ",",
  na.strings = c("", "NA", "NaN"),
  stringsAsFactors = FALSE,
  strip.white = TRUE
)

# Trim whitespace from column names (in case of leading spaces)
names(fisher_raw) <- trimws(names(fisher_raw))

# Sanity check expected columns
expected_cols <- c("gene_id", "log2FC_6.8", "padj_6.8",
                   "log2FC_16.18", "padj_16.18")
missing_cols <- setdiff(expected_cols, names(fisher_raw))
if (length(missing_cols) > 0L) {
  stop("Fisher CSV missing expected columns: ",
       paste(missing_cols, collapse = ", "),
       "\nAvailable: ", paste(names(fisher_raw), collapse = ", "))
}

# Confirm numeric columns parsed correctly
stopifnot(is.numeric(fisher_raw$log2FC_6.8))
cat("Fisher table loaded: ", nrow(fisher_raw), " genes, ",
    "decimal-comma-parsed values numeric.\n", sep = "")

# Rename the symbol column to be explicit
names(fisher_raw)[names(fisher_raw) == "gene_id"] <- "symbol_fisher"
fisher_raw$symbol_fisher <- trimws(fisher_raw$symbol_fisher)

# ---- Step 3: Map Fisher symbols to FBgn ----
# org.Dm.eg.db SYMBOL -> FLYBASE mapping. Many-to-one is possible but
# rare; one-to-many is also possible (one symbol maps to multiple FBgns
# because of historical synonyms). We take the first match per symbol
# and report ambiguities for manual review.

sym2fb <- AnnotationDbi::select(
  org.Dm.eg.db,
  keys     = unique(fisher_raw$symbol_fisher),
  columns  = c("SYMBOL", "FLYBASE"),
  keytype  = "SYMBOL"
)

# Diagnostic: how many symbols are unmappable?
unmappable <- unique(sym2fb$SYMBOL[is.na(sym2fb$FLYBASE)])
n_unmappable <- length(unmappable)
cat("Symbols unmappable to FBgn: ", n_unmappable, " (",
    sprintf("%.1f%%",
            100 * n_unmappable / length(unique(fisher_raw$symbol_fisher))),
    ")\n", sep = "")
if (n_unmappable > 0L && n_unmappable <= 20L) {
  cat("Unmappable symbols (first 20):\n")
  cat(paste("  ", head(unmappable, 20), collapse = "\n"), "\n")
}

# Diagnostic: how many symbols map to multiple FBgns?
multi <- sym2fb[!is.na(sym2fb$FLYBASE), ]
multi_counts <- table(multi$SYMBOL)
multi_symbols <- names(multi_counts[multi_counts > 1L])
cat("Symbols mapping to >1 FBgn: ", length(multi_symbols), "\n", sep = "")
if (length(multi_symbols) > 0L && length(multi_symbols) <= 20L) {
  cat("Multi-mapping (taking first):\n")
  for (s in head(multi_symbols, 20)) {
    cat("  ", s, " -> ",
        paste(multi$FLYBASE[multi$SYMBOL == s], collapse = ", "),
        "\n", sep = "")
  }
}

# Take first FBgn per symbol (deterministic; arbitrary among ties)
sym2fb_first <- sym2fb[!duplicated(sym2fb$SYMBOL), ]
sym2fb_first <- sym2fb_first[!is.na(sym2fb_first$FLYBASE), ]

# ---- Fisher (2023) symbols -> FlyBase gene identifiers ----
# Fisher et al. use some historical gene symbols that differ from the current
# org.Dm.eg.db SYMBOL for the same gene; without a synonym route the
# symbol->FBgn join silently drops those rows, leaving blank fold-change cells
# in the per-gene direct-target table. Symbols are mapped by current symbol,
# then unambiguous FlyBase synonym, then CG number (map_fisher_symbols()).
fmap <- map_fisher_symbols(fisher_raw$symbol_fisher)
sym2fb_first <- data.frame(SYMBOL  = fmap$SYMBOL[!is.na(fmap$FLYBASE)],
                           FLYBASE = fmap$FLYBASE[!is.na(fmap$FLYBASE)],
                           stringsAsFactors = FALSE)
stopifnot(sym2fb_first$FLYBASE[sym2fb_first$SYMBOL == "RpII18"] == "FBgn0003275",
          sym2fb_first$FLYBASE[sym2fb_first$SYMBOL == "CG6752"] == "FBgn0038296",
          sym2fb_first$FLYBASE[sym2fb_first$SYMBOL == "RpII15"] == "FBgn0004855")

# Attach FBgn to Fisher table
fisher_raw$gene_id <- sym2fb_first$FLYBASE[
  match(fisher_raw$symbol_fisher, sym2fb_first$SYMBOL)
]

# Drop rows with no FBgn match (can't be joined)
n_before <- nrow(fisher_raw)
fisher_mapped <- fisher_raw[!is.na(fisher_raw$gene_id), ]
n_dropped <- n_before - nrow(fisher_mapped)
cat("Fisher rows after FBgn mapping: ", nrow(fisher_mapped),
    " (", n_dropped, " dropped, no FBgn)\n", sep = "")

# Handle duplicate FBgn -> symbol (rare): keep the row with the
# strongest signal (lowest p across both stages)
if (any(duplicated(fisher_mapped$gene_id))) {
  fisher_mapped$min_padj <- pmin(fisher_mapped$padj_6.8,
                                  fisher_mapped$padj_16.18,
                                  na.rm = TRUE)
  fisher_mapped <- fisher_mapped[order(fisher_mapped$gene_id,
                                        fisher_mapped$min_padj), ]
  fisher_mapped <- fisher_mapped[!duplicated(fisher_mapped$gene_id), ]
  fisher_mapped$min_padj <- NULL
}

# ---- Step 4: Join ----
joined <- merge(
  per_gene_peaks,
  fisher_mapped[, c("gene_id", "symbol_fisher",
                    "log2FC_6.8", "padj_6.8",
                    "log2FC_16.18", "padj_16.18")],
  by    = "gene_id",
  all.x = TRUE,
  sort  = FALSE
)

# Diagnostic: how many of the 21 direct targets have Fisher data?
n_with_fisher <- sum(!is.na(joined$log2FC_6.8) |
                     !is.na(joined$log2FC_16.18))
cat("\nDirect targets with Fisher fold-change data: ",
    n_with_fisher, " / ", nrow(joined), "\n", sep = "")

if (n_with_fisher < nrow(joined)) {
  cat("Direct targets WITHOUT Fisher data (FBgn -> symbol mismatch ",
      "or gene absent from Fisher 2023):\n", sep = "")
  no_fisher <- joined[is.na(joined$log2FC_6.8) &
                      is.na(joined$log2FC_16.18), ]
  for (i in seq_len(nrow(no_fisher))) {
    cat("  ", no_fisher$gene_id[i], " (",
        no_fisher$SYMBOL[i], ")\n", sep = "")
  }
}

# ---- Step 5: Sort by signal value and write ----
joined <- joined[order(-joined$signalValue), ]

write.csv(joined,
          file.path(CH5_OUT_DIR, "direct_targets_with_intensity_and_fc.csv"),
          row.names = FALSE)
cat("\nWrote: direct_targets_with_intensity_and_fc.csv (",
    nrow(joined), " rows)\n", sep = "")

# ---- Step 6: Print the table ----
print(joined, row.names = FALSE)
cat("\nNote: SYMBOL column is from org.Dm.eg.db (peak annotation).\n")
cat("Fisher's own symbol_fisher column may differ; both are in the CSV.\n")
```
```{r ch5-direct-targets-mapping-sensitivity, include=TRUE, eval=TRUE}
# Sensitivity of the direct-target set to how the Fisher et al. (2023) table is
# joined. The DEG list behind the 21-gene set (Fisher_combined_DEGs.csv) was
# built with |log2FC| >= 0.49 and matched Fisher symbols to current FlyBase
# symbols. The same rule is re-applied here with FlyBase-synonym mapping
# (map_fisher_symbols()), at the stated cut (0.49) and at exactly
# log2(1.4) = 0.485 (the unrounded 1.4-fold cut). The 21-gene set is the
# reported set; these values are reported as a sensitivity analysis.
fs <- read.delim(file.path(REFERENCE_LISTS, "Fisher2023_per_gene_fold_changes.csv"),
                 header = TRUE, sep = "\t", dec = ",", na.strings = c("", "NA", "NaN"),
                 stringsAsFactors = FALSE, strip.white = TRUE)
names(fs) <- trimws(names(fs))
fsm <- map_fisher_symbols(fs$gene_id)
fs$FBgn <- fsm$FLYBASE[match(trimws(fs$gene_id), fsm$SYMBOL)]
rule_degs <- function(thr) with(fs, unique(FBgn[!is.na(FBgn) & (
  (!is.na(padj_6.8)   & padj_6.8   < 0.05 & abs(log2FC_6.8)   >= thr) |
  (!is.na(padj_16.18) & padj_16.18 < 0.05 & abs(log2FC_16.18) >= thr))]))
univ_test <- function(degs) {
  cu <- intersect(trh_chip_genes, univ_genes); du <- intersect(degs, univ_genes)
  k <- length(intersect(cu, du)); N <- length(univ_genes); n <- length(cu); K <- length(du)
  ft <- fisher.test(matrix(c(k, n - k, K - k, N - K - n + k), 2, 2))
  list(k = k, K = K, exp = n * K / N, or = unname(ft$estimate), ci = ft$conf.int)
}
dt_049   <- intersect(trh_chip_genes, rule_degs(0.49))
dt_exact <- intersect(trh_chip_genes, rule_degs(log2(1.4)))
stopifnot(all(direct_targets_trh %in% dt_exact), length(direct_targets_trh) == 21L)
u_exact <- univ_test(rule_degs(log2(1.4)))
u_049   <- univ_test(rule_degs(0.49))
cat(sprintf("mapping: %d symbols, %d recovered by synonym/CG, %d unmapped\n",
            nrow(fsm), sum(fsm$route %in% c("alias", "flybasecg")), sum(is.na(fsm$FLYBASE))))
cat(sprintf("targets: reported %d | synonym map, cut 0.49: %d | synonym map, exact 1.4-fold: %d\n",
            length(direct_targets_trh), length(dt_049), length(dt_exact)))
cat("added under the exact 1.4-fold rule:", setdiff(dt_exact, direct_targets_trh), "\n")
save_values("ch5_fisher_mapping", list(
  nFisherAliasRecovered   = sum(fsm$route %in% c("alias", "flybasecg")),
  nFisherUnmapped         = sum(is.na(fsm$FLYBASE)),
  fisherDegLogFcCut       = "0.49",
  nDirectTargetsAliasFix  = length(dt_049),
  nDirectTargetsExactRule = length(dt_exact),
  dtAliasUnivK            = u_049$k,
  dtAliasUnivExpected     = sprintf("%.1f", u_049$exp),
  dtAliasUnivOR           = sprintf("%.2f", u_049$or),
  dtAliasUnivCIlo         = sprintf("%.2f", u_049$ci[1]),
  dtAliasUnivCIhi         = sprintf("%.2f", u_049$ci[2]),
  dtExactUnivK            = u_exact$k,
  dtExactUnivExpected     = sprintf("%.1f", u_exact$exp),
  dtExactUnivOR           = sprintf("%.2f", u_exact$or),
  dtExactUnivCIlo         = sprintf("%.2f", u_exact$ci[1]),
  dtExactUnivCIhi         = sprintf("%.2f", u_exact$ci[2])
))
```



```{r ch5-tissue-filter-21to21, include=TRUE, eval=TRUE}
# Annotates 21-gene direct-target set using 21gene_tissue_annotation.csv.
# Retains gene if Y in trachea, sal_gland or cns_midline; outputs tissue_filtered_21to21.csv.
# Computes nDirectTargetsFiltered/nDirectTargetsExcluded for the ch5-export-computed-values chunk.

# Pre-flight: the 21-gene direct-target set must exist in memory.
stopifnot(exists("direct_targets_trh"))
stopifnot(length(direct_targets_trh) > 0L)

# Path: curator places the CSV in CH5_INPUT_DIR.
P_TISSUE_ANNOT <- file.path(CH5_INPUT_DIR, "21gene_tissue_annotation.csv")

if (!file.exists(P_TISSUE_ANNOT)) {
  warning(
    "Tissue annotation CSV not found at:\n  ", P_TISSUE_ANNOT,
    "\nThe tissue-annotation step will be skipped and the retained/excluded\n",
    "counts recorded as NA."
  )
  n_retained_tissue <- NA_integer_
  n_excluded_tissue <- NA_integer_
} else {

  tissue_annot <- read.csv(P_TISSUE_ANNOT, stringsAsFactors = FALSE)

  # ---- Schema and content validation ----
  required_cols <- c("gene_id", "SYMBOL", "trachea", "sal_gland",
                     "cns_midline", "in_bdgp")
  missing_cols  <- setdiff(required_cols, names(tissue_annot))
  if (length(missing_cols) > 0L) {
    stop("21gene_tissue_annotation.csv is missing column(s): ",
         paste(missing_cols, collapse = ", "))
  }

  # Coerce Y/N to upper-case and strip whitespace
  yn_cols <- c("trachea", "sal_gland", "cns_midline", "in_bdgp")
  for (col in yn_cols) {
    tissue_annot[[col]] <- toupper(trimws(tissue_annot[[col]]))
  }

  bad_vals <- !unlist(tissue_annot[, yn_cols]) %in% c("Y", "N")
  if (any(bad_vals)) {
    stop("21gene_tissue_annotation.csv contains non-{Y,N} values in ",
         "Y/N columns. Inspect rows: ",
         paste(which(rowSums(
           !sapply(tissue_annot[, yn_cols], `%in%`, c("Y", "N"))
         ) > 0), collapse = ", "))
  }

  # Row count must match the 21-gene set exactly.
  if (nrow(tissue_annot) != length(direct_targets_trh)) {
    stop("Row count mismatch: tissue_annot has ", nrow(tissue_annot),
         " rows but direct_targets_trh has ",
         length(direct_targets_trh), " genes.")
  }

  # FBgn set must match the 21-gene set exactly.
  set_in_csv    <- setdiff(tissue_annot$gene_id, direct_targets_trh)
  set_in_direct <- setdiff(direct_targets_trh, tissue_annot$gene_id)
  if (length(set_in_csv) > 0L || length(set_in_direct) > 0L) {
    stop("FBgn set mismatch between 21gene_tissue_annotation.csv and ",
         "the direct-target set.\n",
         "  In CSV but not in direct targets: ",
         paste(set_in_csv, collapse = ", "), "\n",
         "  In direct targets but not in CSV: ",
         paste(set_in_direct, collapse = ", "))
  }

  # ---- Apply filtering rule ----
  tissue_annot$retained <- with(
    tissue_annot,
    ifelse(trachea == "Y" | sal_gland == "Y" | cns_midline == "Y",
           "Y", "N")
  )

  n_retained_tissue <- sum(tissue_annot$retained == "Y")
  n_excluded_tissue <- sum(tissue_annot$retained == "N")

  # ---- Console summary ----
  cat("─── Spatio-temporal filter (22 → 17) ────────────────────\n")
  cat(sprintf("Input direct-target genes:   %d\n",  nrow(tissue_annot)))
  cat(sprintf("Retained (tissue-positive):  %d\n",  n_retained_tissue))
  cat(sprintf("Excluded (tissue-negative):  %d\n",  n_excluded_tissue))

  excluded <- tissue_annot[tissue_annot$retained == "N", ]
  if (nrow(excluded) > 0L) {
    cat("\nExcluded genes (symbol — exclusion reason):\n")
    for (i in seq_len(nrow(excluded))) {
      reason <- if (excluded$in_bdgp[i] == "N") {
        "absent from BDGP database"
      } else {
        "in BDGP but not in queried tissues at stages 11–16"
      }
      cat(sprintf("  %-12s  %s\n", excluded$SYMBOL[i], reason))
    }
  }

  # ---- Export annotated table ----
  P_TISSUE_OUT <- file.path(CH5_OUT_DIR, "tissue_filtered_21to21.csv")
  write.csv(tissue_annot, P_TISSUE_OUT, row.names = FALSE)
  cat(sprintf("\nWrote: %s\n", P_TISSUE_OUT))
}
```

---

```{r ch5-rib-comparison, include=TRUE, eval=TRUE}
# ---- Section 5.6: Convergence with Loganathan Rib direct targets ----
# Two-way and three-way overlaps; Fisher's exact tests against the
# dm6-wide gene background.

trh_rib_overlap   <- intersect(direct_targets_trh, rib_target_genes)
n_trh_rib_overlap <- length(trh_rib_overlap)

# 2x2 contingency table for two-way overlap:
#                       | DirectTarget Trh | not |
# RibTarget             |        a         |  b  |
# not RibTarget         |        c         |  d  |
# Universe = all_dm6_genes
a <- n_trh_rib_overlap
b <- length(setdiff(rib_target_genes, direct_targets_trh))
c2 <- length(setdiff(direct_targets_trh, rib_target_genes))
d <- length(all_dm6_genes) - (a + b + c2)

ct_2way <- matrix(c(a, b, c2, d), nrow = 2, byrow = TRUE,
                  dimnames = list(c("Rib_target", "Not_Rib_target"),
                                  c("Trh_target", "Not_Trh_target")))

ft_2way <- fisher.test(ct_2way, alternative = "two.sided")

cat("=== Trh/Tgo direct targets vs Rib direct targets ===\n")
print(ct_2way)
cat(sprintf("\nOR = %.2f, p = %.3g, %d-gene overlap\n",
            unname(ft_2way$estimate), ft_2way$p.value, n_trh_rib_overlap))

# Three-way overlap: Trh-ChIP genes x Fisher DEGs x Rib targets
three_way <- Reduce(intersect, list(trh_chip_genes, fisher_degs,
                                    rib_target_genes))
n_three_way <- length(three_way)
cat("\nThree-way overlap (Trh ChIP x Fisher DEG x Rib target):",
    n_three_way, "gene(s)\n")
if (n_three_way > 0L && n_three_way <= 30L) {
  # Print the gene symbols for inline reference if small
  three_way_sym <- AnnotationDbi::select(org.Dm.eg.db,
                                          keys = three_way,
                                          columns = c("FLYBASE", "SYMBOL"),
                                          keytype = "FLYBASE")
  print(three_way_sym)
}

# Save outputs
write.csv(
  data.frame(FBgn = trh_rib_overlap),
  file.path(CH5_OUT_DIR, "direct_targets_Trh_Rib_overlap.csv"),
  row.names = FALSE
)
write.csv(
  data.frame(FBgn = three_way),
  file.path(CH5_OUT_DIR, "direct_targets_three_way_overlap.csv"),
  row.names = FALSE
)
```

---

```{r ch5-functional-enrichment, include=TRUE, eval=TRUE}
# ---- Section 5.5/5.6: GO enrichment on direct-target set ----
# Run on (a) all Trh/Tgo direct targets, (b) Trh-Rib overlap.
# Skipped if clusterProfiler is unavailable or set sizes are too small.

if (requireNamespace("clusterProfiler", quietly = TRUE) &&
    length(direct_targets_trh) >= 10L) {

  ego_trh <- clusterProfiler::enrichGO(
    gene          = direct_targets_trh,
    OrgDb         = org.Dm.eg.db,
    keyType       = "FLYBASE",
    ont           = "ALL",
    pAdjustMethod = "BH",
    qvalueCutoff  = 0.05,
    readable      = TRUE
  )

  if (!is.null(ego_trh) && nrow(as.data.frame(ego_trh)) > 0L) {
    write.csv(as.data.frame(ego_trh),
              file.path(CH5_OUT_DIR, "GO_TrhTgo_direct_targets.csv"),
              row.names = FALSE)
    cat("GO enrichment on Trh/Tgo direct targets: ",
        nrow(as.data.frame(ego_trh)), " significant terms\n", sep = "")
  } else {
    cat("GO enrichment on Trh/Tgo direct targets: no significant terms\n")
  }

  if (length(trh_rib_overlap) >= 5L) {
    ego_overlap <- clusterProfiler::enrichGO(
      gene          = trh_rib_overlap,
      OrgDb         = org.Dm.eg.db,
      keyType       = "FLYBASE",
      ont           = "ALL",
      pAdjustMethod = "BH",
      qvalueCutoff  = 0.10,    # relaxed for smaller set
      readable      = TRUE
    )
    if (!is.null(ego_overlap) && nrow(as.data.frame(ego_overlap)) > 0L) {
      write.csv(as.data.frame(ego_overlap),
                file.path(CH5_OUT_DIR, "GO_Trh_Rib_overlap.csv"),
                row.names = FALSE)
      cat("GO enrichment on Trh-Rib overlap: ",
          nrow(as.data.frame(ego_overlap)), " significant terms\n",
          sep = "")
    } else {
      cat("GO enrichment on Trh-Rib overlap: no significant terms\n")
    }
  }
} else {
  cat("Skipping GO enrichment ",
      "(clusterProfiler not installed or set too small)\n", sep = "")
}
```

---

```{r ch5-chung2011-gse28780, include=TRUE, eval=TRUE}
# ============================================================
#  Second perturbation axis: Chung, Chavez & Andrew (2011) trh-null
#  microarray (GEO GSE28780, stages 11-16), intersected with Trh/Tgo peaks.
#  Peak-gene mapping: strand-aware nearest-TSS distance, dm6 ensGene.
#  DE: limma, 3 WT vs 3 trh-null, |FC|>=1.4 at p<=0.05; universe = genes with FlyBase ID; hypergeometric test.
#  Result: enrichment restricted to distal peaks (>2kb from TSS); proximal (<=2kb) show none.
# ============================================================

suppressPackageStartupMessages({
  library(GEOquery); library(limma); library(GenomicFeatures)
  library(org.Dm.eg.db); library(AnnotationDbi)
})

CHUNG_FC   <- log2(1.4)            # Chung 2011 fold-change threshold
CHUNG_PMAX <- 0.05                 # Chung 2011 raw-p threshold
TSS_SPLIT  <- 2000L                # proximal/distal boundary (matches ch5 window)

##, 1. Peak -> nearest gene TSS (strand-aware, coordinate-based) ------------
stopifnot(length(Trh_Tgo_joint_filt) == 614L)
jp <- Trh_Tgo_joint_filt
if (is.null(names(jp))) names(jp) <- paste0("peak_", seq_along(jp))
g   <- suppressMessages(genes(txdb))
# Gene ID (FBgn): use gene_id metadata column, always populated.
# names(g) is NULL in some GenomicFeatures builds, giving a
# zero-length column and a "differing number of rows" error.
# Convert GRanges to data.frame in one step.
# Individual accessors fail on some GenomicRanges/S4Vectors
# builds with "S3coerceFUN(from) : not supported".
gdf <- as.data.frame(g)                        # seqnames, start, end, width, strand + mcols
gid_col <- if ("gene_id" %in% names(gdf)) "gene_id" else NULL
if (is.null(gid_col)) {
  # fallback: rownames carry FBgn when names(g) was non-NULL
  gdf$fbgn <- rownames(gdf)
} else {
  gdf$fbgn <- as.character(gdf$gene_id)
}
gdf$chr    <- as.character(gdf$seqnames)
gdf$strand <- as.character(gdf$strand)
gdf$tss    <- ifelse(gdf$strand == "-", gdf$end, gdf$start)
gsplit <- split(gdf, gdf$chr)
nn <- lapply(seq_along(jp), function(i) {
  gi <- gsplit[[ as.character(seqnames(jp)[i]) ]]
  if (is.null(gi) || nrow(gi) == 0)
    return(data.frame(geneId = NA_character_, dist_to_TSS = NA_integer_))
  d <- pmax(start(jp)[i] - gi$tss, gi$tss - end(jp)[i], 0L)
  j <- which.min(d)
  data.frame(geneId = gi$fbgn[j], dist_to_TSS = as.integer(d[j]))
})
chung_pk <- cbind(data.frame(peak = names(jp),
                             start = start(jp), end = end(jp),
                             stringsAsFactors = FALSE),
                  do.call(rbind, nn))
saveRDS(chung_pk, file.path(CH5_OUT_DIR, "trh_tgo_peak_nearest_gene.rds"))

##, 2. GSE28780: fresh limma DE (trh-null vs WT) ----------------------------
options(timeout = 600)
gse28780 <- getGEO("GSE28780", destdir = CH5_OUT_DIR,
                   GSEMatrix = TRUE, getGPL = TRUE)[[1]]
pd28 <- Biobase::pData(gse28780)
fd28 <- Biobase::fData(gse28780)
ex28 <- Biobase::exprs(gse28780)
if (max(ex28, na.rm = TRUE) > 100) ex28 <- log2(ex28 + 1)   # ensure log2 scale
stopifnot(ncol(ex28) == 6L)

chung_grp    <- factor(ifelse(grepl("trh", pd28$title), "trhnull", "WT"),
                       levels = c("WT", "trhnull"))          # 3 WT, 3 trh-null
chung_design <- model.matrix(~ chung_grp)
chung_fit    <- eBayes(lmFit(ex28, chung_design))
chung_tt     <- topTable(chung_fit, coef = "chung_grptrhnull",
                         number = Inf, sort.by = "none")
chung_tt$probe   <- rownames(chung_tt)
chung_tt$entrez1 <- sub(" ?///.*", "", fd28[chung_tt$probe, "ENTREZ_GENE_ID"])
e2fb <- suppressMessages(AnnotationDbi::mapIds(
          org.Dm.eg.db, keys = unique(chung_tt$entrez1[chung_tt$entrez1 != ""]),
          column = "FLYBASE", keytype = "ENTREZID", multiVals = "first"))
chung_tt$fbgn <- e2fb[chung_tt$entrez1]
chung_ok      <- chung_tt[!is.na(chung_tt$fbgn), ]
saveRDS(chung_tt, file.path(CH5_OUT_DIR, "gse28780_DE.rds"))

##, 3. Responsive-gene sets and enrichment (proximal vs distal) -------------
chung_universe <- unique(chung_ok$fbgn)

# Direction is assigned per gene, not per probe: each gene is resolved to
# its most significant probe before thresholding, so a gene with one up
# probe and one down probe is not entered into both the `down` and `up`
# sets. This makes the `down` and `up` sets genuinely disjoint, consistent
# with the two being presented as directional opposites. The assertion
# below checks that this holds.
.chung_best <- chung_ok[order(chung_ok$fbgn, chung_ok$P.Value), ]
.chung_best <- .chung_best[!duplicated(.chung_best$fbgn), ]
chung_sets <- list(
  down = .chung_best$fbgn[.chung_best$P.Value <= CHUNG_PMAX &
                          .chung_best$logFC   <= -CHUNG_FC],
  up   = .chung_best$fbgn[.chung_best$P.Value <= CHUNG_PMAX &
                          .chung_best$logFC   >=  CHUNG_FC],
  any  = .chung_best$fbgn[.chung_best$P.Value <= CHUNG_PMAX &
                          abs(.chung_best$logFC) >= CHUNG_FC])
stopifnot(
  length(intersect(chung_sets$down, chung_sets$up)) == 0L,
  length(chung_sets$down) + length(chung_sets$up) == length(chung_sets$any))

# Proximal/distal must be a PARTITION of the bound genes: each gene is
# classified once, by its CLOSEST peak, rather than by class membership of
# each peak followed by de-duplication by gene (which would let a gene
# nearest to both a proximal and a distal peak enter both classes). This
# keeps the two hypergeometric tests below comparing disjoint classes.
.chung_pk_ok <- chung_pk[!is.na(chung_pk$geneId), ]
.chung_min   <- aggregate(dist_to_TSS ~ geneId, data = .chung_pk_ok, FUN = min)
chung_prox  <- .chung_min$geneId[.chung_min$dist_to_TSS <= TSS_SPLIT]
chung_dist  <- .chung_min$geneId[.chung_min$dist_to_TSS >  TSS_SPLIT]
chung_bound <- .chung_min$geneId
stopifnot(
  length(intersect(chung_prox, chung_dist)) == 0L,
  length(chung_prox) + length(chung_dist) == length(chung_bound))

chung_hyper <- function(pgset, deset) {
  pgset <- intersect(pgset, chung_universe)
  deset <- intersect(deset, chung_universe)
  ov    <- intersect(pgset, deset)
  expd  <- length(pgset) * length(deset) / length(chung_universe)
  data.frame(n_peak = length(pgset), n_DE = length(deset), overlap = length(ov),
             expected = expd, fold = length(ov) / expd,
             p = phyper(length(ov) - 1L, length(deset),
                        length(chung_universe) - length(deset),
                        length(pgset), lower.tail = FALSE))
}
chung_enrich <- do.call(rbind, lapply(
  list(c("prox","down"),c("prox","up"),c("prox","any"),
       c("dist","down"),c("dist","up"),c("dist","any")),
  function(cc) { r <- chung_hyper(get(paste0("chung_", cc[1])), chung_sets[[cc[2]]])
                 r$loc <- cc[1]; r$set <- cc[2]; r }))
write.csv(chung_enrich[, c("loc","set","n_peak","overlap","expected","fold","p")],
          file.path(CH5_OUT_DIR, "chung_distal_proximal_enrichment.csv"),
          row.names = FALSE)
print(chung_enrich[, c("loc","set","overlap","expected","fold","p")])

# High-confidence set: bound and responsive in both Chung and Fisher 2023.
# chung_sets$down is based on Chung fold-change only
chung_conj_any  <- Reduce(intersect, list(chung_bound, fisher_degs, chung_sets$any))
chung_conj_down <- Reduce(intersect, list(chung_bound, fisher_degs, chung_sets$down))
hc <- chung_ok[match(chung_conj_any, chung_ok$fbgn),
               c("fbgn","logFC","P.Value")]
hc$symbol      <- suppressMessages(AnnotationDbi::mapIds(
                    org.Dm.eg.db, keys = hc$fbgn, column = "SYMBOL",
                    keytype = "FLYBASE", multiVals = "first"))
hc$dist_to_TSS <- chung_pk$dist_to_TSS[match(hc$fbgn, chung_pk$geneId)]

##, 4a. Cross-platform DIRECTIONAL concordance ------------------------------
# Fisher direction is taken from the better-supported of the two stage windows
# (smaller BH-adjusted p). `fr` is the Fisher per-gene table built in the
# TEST 2 chunk; if it is not in session, source that chunk first.
stopifnot(exists("fr"))
.fi <- fr[match(hc$fbgn, fr$gene_id), ]
.use68 <- !is.na(.fi$padj_6.8) &
          (is.na(.fi$padj_16.18) | .fi$padj_6.8 <= .fi$padj_16.18)
hc$fisher_log2FC <- ifelse(.use68, .fi$log2FC_6.8, .fi$log2FC_16.18)
hc$fisher_padj   <- ifelse(.use68, .fi$padj_6.8,   .fi$padj_16.18)
hc$concordant    <- sign(hc$logFC) == sign(hc$fisher_log2FC)

n_conj_concordant <- sum(hc$concordant, na.rm = TRUE)
n_conj_discordant <- sum(!hc$concordant, na.rm = TRUE)
# Positional composition of the conjunction set. nConjDistalNovel counts genes
# recoverable only without the +/-2 kb constraint, which for this set is the
# distal subset itself.
n_conj_proximal     <- sum(abs(hc$dist_to_TSS) <= 2000, na.rm = TRUE)
n_conj_distal       <- sum(abs(hc$dist_to_TSS) >  2000, na.rm = TRUE)
n_conj_distal_novel <- n_conj_distal
stopifnot(n_conj_proximal + n_conj_distal == nrow(hc))

hc <- hc[order(!hc$concordant, hc$P.Value), ]   # concordant block first
write.csv(hc, file.path(CH5_OUT_DIR, "chung_fisher_highconf_targets.csv"),
          row.names = FALSE)

cat(sprintf("Conjunction: %d genes | directionally concordant %d, discordant %d\n",
            nrow(hc), n_conj_concordant, n_conj_discordant))
cat(sprintf("  discordant: %s\n",
            paste(hc$symbol[!hc$concordant], collapse = ", ")))

##, 4b. Universe-robustness check -------------------------------------------
# Headline enrichment uses all array genes with an FBgn as universe.
# Re-tests against expressed-gene universes at increasing mean log2
# intensity cutoffs. Distal enrichment stays stable; proximal
# enrichment drops further. Range reported in Chapter 5.
chung_ok$meanExpr <- rowMeans(ex28[chung_ok$probe, , drop = FALSE])
chung_gene_expr   <- tapply(chung_ok$meanExpr, chung_ok$fbgn, max)  # gene = brightest probe
chung_robust <- do.call(rbind, lapply(c(5, 6, 7), function(cut) {
  uni <- names(chung_gene_expr)[chung_gene_expr >= cut]
  pgd <- intersect(chung_dist, uni); ded <- intersect(chung_sets$any, uni)
  ovd <- intersect(pgd, ded); expd <- length(pgd) * length(ded) / length(uni)
  data.frame(cutoff = cut, n_universe = length(uni),
             dist_fold = length(ovd) / expd,
             dist_p = phyper(length(ovd) - 1L, length(ded),
                             length(uni) - length(ded), length(pgd),
                             lower.tail = FALSE))
}))
write.csv(chung_robust,
          file.path(CH5_OUT_DIR, "chung_universe_robustness.csv"),
          row.names = FALSE)
print(chung_robust)

##, 5. Convenience scalars for the export chunk -----------------------------
chung_get <- function(loc, set, col)
  chung_enrich[chung_enrich$loc == loc & chung_enrich$set == set, col]
cat(sprintf("Chung GSE28780: universe=%d, bound-on-array=%d (prox=%d, dist=%d)\n",
    length(chung_universe), length(intersect(chung_bound, chung_universe)),
    length(intersect(chung_prox, chung_universe)),
    length(intersect(chung_dist, chung_universe))))
cat(sprintf("Distal responsive fold=%.2f (p=%.2g); proximal fold=%.2f (p=%.2g)\n",
    chung_get("dist","any","fold"), chung_get("dist","any","p"),
    chung_get("prox","any","fold"), chung_get("prox","any","p")))
cat(sprintf("Chung n Fisher conjunction (any/down): %d / %d\n",
    length(chung_conj_any), length(chung_conj_down)))
```

```{r ch5-fig-chung-distal-proximal, include=TRUE, eval=TRUE}
# ---- Fig 5.7: distal- vs proximal-peak enrichment for trh-responsiveness ----
# Fold enrichment (observed/expected) of Chung 2011 trh-responsive genes among
# Trh/Tgo target genes, split by peak-to-TSS distance. Uses the shared bar
# chart style (thesis_bar, defined in the setup chunk): contrast-aware in-bar labels, fold + (n/N),
# significance stars spaced above the bars so they never overlap the bar top.
# The significance legend is split across the two panels' top-right slots:
# the n.s. key over the proximal panel, the *** key over the distal panel.
ord    <- c("down","up","any")
setlab <- c(down="Trh-activated\n(down in null)", up="Trh-repressed\n(up in null)",
            any="Any\nresponse")
loc_lab <- c(prox="Proximal peaks (\u2264 2 kb from TSS)",
             dist="Distal peaks (> 2 kb from TSS)")
fills56 <- setNames(unname(PAL_DIRECTION[ord]), setlab[ord])

pdat <- chung_enrich
pdat$xlab  <- factor(setlab[as.character(pdat$set)], levels = setlab[ord])
pdat$panel <- factor(loc_lab[as.character(pdat$loc)], levels = loc_lab)
pdat$Count <- sprintf("(%d/%d)", pdat$overlap, pdat$n_peak)
# In-bar fold value carries an explicit "x" suffix (2.48x = 2.48-fold over
# expected) so the number is unambiguous without mislabelling it as an or.
pdat$Fold  <- sprintf("%.2f\u00d7", pdat$fold)
pdat$Sig   <- thesis_stars(pdat$p)
# Split significance legend: proximal slot -> n.s. key; distal slot -> *** key.
# The distal panel contains a non-significant bar too (trh-activated), so its
# key must cover both symbols, otherwise the panel has no legend entry for
# one of its own bars.
pdat$Nlab  <- ifelse(pdat$loc == "prox", "n.s. - not significant",
                     "*** p < 1e-3;  n.s. - not significant")

p <- thesis_bar(pdat, x = "xlab", y = "fold", fill = "xlab",
                value = "Fold", count = "Count", sig = "Sig",
                facet = "panel", n_lab = "Nlab", n_lab_size = 4.4, fills = fills56,
                y_max = 3.35, dashed_at = 1,
                y_lab = "Fold enrichment (observed / expected)")
ggsave(file.path(FIGURES_DIR, "fig5_6_chung_distal_proximal.png"), p,
       width = 8.3, height = 4.8, dpi = 300, bg = "white")
cat("Wrote fig5_6_chung_distal_proximal.png\n")
```

---
```{r ch5-test2-motif2-perturbation, include=TRUE, eval=TRUE}
# ============================================================
#  Test 2: does Motif2 (STRCACTGARARAAA) at Trh/Tgo peaks predict transcriptional response to trh knockdown (Fisher et al. 2023 modERN RNA-seq)
#  Motif2 hits from FIMO scan (fimo_joint/fimo.tsv, q<=0.05); peak-gene assignment by strand-aware nearest TSS (dm6 ensGene), split proximal <=2kb vs distal >2kb
#  Outcome: |log2FC| at 6-8h window. Inputs: Trh_Tgo_joint_filt (614 peaks), REFERENCE_LISTS/Fisher2023_per_gene_fold_changes.csv
#  Result: null in all framings, including the distal set
# ============================================================
suppressMessages({
  library(GenomicRanges); library(rtracklayer); library(ggplot2)
  library(TxDb.Dmelanogaster.UCSC.dm6.ensGene); library(org.Dm.eg.db)
})
stopifnot(exists("Trh_Tgo_joint_filt"), length(Trh_Tgo_joint_filt) == 614L)
txdb <- TxDb.Dmelanogaster.UCSC.dm6.ensGene

##, 1. Motif2 flag per peak (canonical peak_N -> coordinate mapping) --------
# FIMO joint-peak scan produced by the Chapter 3 analysis (data/chapter3).
P_FIMO_JOINT <- file.path(PROJECT_DIR, "data", "chapter3",
                          "fimo_joint", "fimo.tsv")
stopifnot(file.exists(P_FIMO_JOINT))
fimo <- read.delim(P_FIMO_JOINT, comment.char = "#", stringsAsFactors = FALSE)
fimo <- fimo[fimo$motif_id == "STRCACTGARARAAA" &
             !is.na(fimo$q.value) & fimo$q.value <= 0.05, ]
pidx <- as.integer(sub("peak_", "", fimo$sequence_name))
stopifnot(all(pidx >= 1L & pidx <= length(Trh_Tgo_joint_filt)))
# peak-local start/stop (1-based within-peak) -> genomic coordinate
fimo_gr <- GRanges(
  seqnames = as.character(seqnames(Trh_Tgo_joint_filt))[pidx],
  ranges   = IRanges(start(Trh_Tgo_joint_filt)[pidx] + fimo$start - 1L,
                     start(Trh_Tgo_joint_filt)[pidx] + fimo$stop  - 1L)
)
has_motif2 <- countOverlaps(Trh_Tgo_joint_filt, fimo_gr, ignore.strand = TRUE) > 0
mcols(Trh_Tgo_joint_filt)$has_motif2 <- has_motif2
n_peaks_m2 <- sum(has_motif2)
cat(sprintf("Motif2 (q<=0.05) present at %d / %d peaks (%.1f%%)\n",
            n_peaks_m2, length(has_motif2), 100 * mean(has_motif2)))

##, 2. Strand-aware nearest-TSS (dm6 ensGene), coordinate-based --------------
g   <- suppressMessages(as.data.frame(genes(txdb)))
g   <- g[as.character(g$seqnames) %in% good_chr, ]
tss <- ifelse(g$strand == "+", g$start, g$end)
tss_gr <- GRanges(g$seqnames, IRanges(tss, tss),
                  strand = g$strand, gene_id = g$gene_id)
nr <- distanceToNearest(Trh_Tgo_joint_filt, tss_gr, ignore.strand = TRUE)
pk <- data.frame(
  gene_id    = tss_gr$gene_id[subjectHits(nr)],
  dist       = mcols(nr)$distance,
  signal     = Trh_Tgo_joint_filt$signalValue[queryHits(nr)],
  has_motif2 = mcols(Trh_Tgo_joint_filt)$has_motif2[queryHits(nr)],
  stringsAsFactors = FALSE
)

##, 3. Fisher per-gene fold change, symbol -> FBgn (Chapter 5 recipe) --------
P_FISHER_PG <- file.path(REFERENCE_LISTS, "Fisher2023_per_gene_fold_changes.csv")
stopifnot(file.exists(P_FISHER_PG))
fr <- read.delim(P_FISHER_PG, header = TRUE, sep = "\t", dec = ",",
                 na.strings = c("", "NA", "NaN"),
                 stringsAsFactors = FALSE, strip.white = TRUE)
names(fr) <- trimws(names(fr))
names(fr)[names(fr) == "gene_id"] <- "symbol_fisher"
fr$symbol_fisher <- trimws(fr$symbol_fisher)
## Same symbol -> FBgn mapping as the direct-target table (map_fisher_symbols()).
mp <- map_fisher_symbols(fr$symbol_fisher)
mp <- mp[!is.na(mp$FLYBASE), c("SYMBOL", "FLYBASE")]
fr$gene_id <- mp$FLYBASE[match(fr$symbol_fisher, mp$SYMBOL)]
fr <- fr[!is.na(fr$gene_id), ]
fr$min_padj <- pmin(fr$padj_6.8, fr$padj_16.18, na.rm = TRUE)
fr <- fr[order(fr$gene_id, fr$min_padj), ]
fr <- fr[!duplicated(fr$gene_id), ]

##, 4. Per-gene stratum table and the tests ---------------------------------
# Collapse peaks to genes within a distance band: a gene is Motif2-positive
# if any assigned peak carries Motif2; strongest peak signal is retained.
build_stratum <- function(dist_lo, dist_hi, label) {
  s  <- pk[pk$dist >= dist_lo & pk$dist <= dist_hi, ]
  gm <- tapply(s$has_motif2, s$gene_id, any)
  gs <- tapply(s$signal,     s$gene_id, max)
  d  <- data.frame(gene_id     = names(gm),
                   has_motif2  = as.logical(gm),
                   peak_signal = as.numeric(gs[names(gm)]),
                   stringsAsFactors = FALSE)
  d  <- merge(d, fr[, c("gene_id", "log2FC_6.8", "log2FC_16.18")],
              by = "gene_id", all.x = TRUE)
  d  <- d[!is.na(d$log2FC_6.8), ]
  d$abs68   <- abs(d$log2FC_6.8)
  d$abs1618 <- abs(d$log2FC_16.18)
  d$stratum <- label
  d
}
d_prox <- build_stratum(0L,    2000L, "Proximal (\u2264 2 kb)")
d_dist <- build_stratum(2001L, Inf,   "Distal (> 2 kb)")
d_all  <- build_stratum(0L,    Inf,   "All peaks")

test2_wilcox <- function(d, col = "abs68") {
  w <- suppressWarnings(wilcox.test(d[[col]] ~ d$has_motif2))
  list(p = unname(w$p.value),
       med_pos = median(d[[col]][d$has_motif2]),
       med_neg = median(d[[col]][!d$has_motif2]),
       n_pos = sum(d$has_motif2), n_neg = sum(!d$has_motif2))
}
r_prox <- test2_wilcox(d_prox)
r_dist <- test2_wilcox(d_dist)
r_all  <- test2_wilcox(d_all)

# Robustness: signal-controlled regression on the pooled set (proximal +
# distal). Does Motif2 add predictive power for |log2FC| once peak signal
# strength is accounted for?
d_reg <- d_all
d_reg$log_signal <- log10(d_reg$peak_signal)
fit_full <- lm(abs68 ~ log_signal + has_motif2, data = d_reg)
fit_red  <- lm(abs68 ~ log_signal,              data = d_reg)
lm_p <- anova(fit_red, fit_full)$`Pr(>F)`[2]
lm_beta <- summary(fit_full)$coefficients["has_motif2TRUE", "Estimate"]

cat("\n== TEST 2: Motif2 vs trh-RNAi |log2FC| (6-8 h) ==\n")
cat(sprintf("  Proximal:  n+=%d n-=%d  med+=%.3f med-=%.3f  Wilcoxon p=%.3f\n",
            r_prox$n_pos, r_prox$n_neg, r_prox$med_pos, r_prox$med_neg, r_prox$p))
cat(sprintf("  Distal:    n+=%d n-=%d  med+=%.3f med-=%.3f  Wilcoxon p=%.3f\n",
            r_dist$n_pos, r_dist$n_neg, r_dist$med_pos, r_dist$med_neg, r_dist$p))
cat(sprintf("  All peaks: n+=%d n-=%d  med+=%.3f med-=%.3f  Wilcoxon p=%.3f\n",
            r_all$n_pos, r_all$n_neg, r_all$med_pos, r_all$med_neg, r_all$p))
cat(sprintf("  Signal-controlled lm: motif2 beta=%.4f  partial-F p=%.3f\n",
            lm_beta, lm_p))

##, 5. Figure: proximal/distal boxplots, |log2FC| by Motif2 status ----------
D <- rbind(d_prox, d_dist)
D$grp     <- factor(ifelse(D$has_motif2, "Motif2-positive", "Motif2-negative"),
                    levels = c("Motif2-negative", "Motif2-positive"))
D$stratum <- factor(D$stratum, levels = c("Proximal (\u2264 2 kb)", "Distal (> 2 kb)"))
pv <- c("Proximal (\u2264 2 kb)" = r_prox$p, "Distal (> 2 kb)" = r_dist$p)
ytop <- as.numeric(quantile(D$abs68, 0.99))
lab_df <- data.frame(stratum = factor(names(pv), levels = levels(D$stratum)),
                     lab = sprintf("Wilcoxon~italic(p)==%.2f", pv))
ns <- aggregate(abs68 ~ stratum + grp, D, length); names(ns)[3] <- "n"

p <- ggplot(D, aes(grp, abs68, fill = grp)) +
  geom_boxplot(width = 0.6, outlier.size = 0.6, alpha = 0.9) +
  facet_wrap(~ stratum) +
  geom_text(data = ns, aes(x = grp, label = paste0("n=", n), y = -0.03),
            inherit.aes = FALSE, size = 3, colour = PAL_INK[["annotation"]]) +
  geom_text(data = lab_df, aes(x = 1.5, y = ytop * 1.08, label = lab),
            inherit.aes = FALSE, size = 3.4, parse = TRUE) +
  scale_fill_manual(values = pal_binary("Motif2-positive", "Motif2-negative"),
                    guide = "none") +
  labs(x = NULL,
       y = expression(group("|", log[2]~"fold change", "|")~"  ("*italic("trh")*"-RNAi, 6-8 h)")) +
  coord_cartesian(ylim = c(-0.06, ytop * 1.15)) +
  theme_bw(base_size = 12) +
  theme(strip.background = element_rect(fill = PAL_INK[["band"]]),
        panel.grid.minor = element_blank())

## Figure height is 4.25 in, giving a slightly larger label size at \linewidth
## than a taller figure would.
ggsave(file.path(FIGURES_DIR, "fig5_7_motif2_perturbation.png"),
       plot = p, width = 8, height = 4.25, dpi = 300, bg = "white")
cat("Wrote fig5_7_motif2_perturbation.png\n")

##, 6. Convenience scalars for the export chunk -----------------------------
# Consumed by ch5-export-computed-values (see the TEST 2 block added to ch5_vals).
test2_n_peaks_motif2  <- n_peaks_m2
test2_prox_p          <- r_prox$p
test2_prox_med_pos    <- r_prox$med_pos
test2_prox_med_neg    <- r_prox$med_neg
test2_prox_n_pos      <- r_prox$n_pos
test2_prox_n_neg      <- r_prox$n_neg
test2_dist_p          <- r_dist$p
test2_dist_med_pos    <- r_dist$med_pos
test2_dist_med_neg    <- r_dist$med_neg
test2_dist_n_pos      <- r_dist$n_pos
test2_dist_n_neg      <- r_dist$n_neg
test2_all_p           <- r_all$p
test2_lm_p            <- lm_p
```
---
```{r ch5-test2-power-and-cme-control, include=TRUE, eval=TRUE}
# ============================================================
#  TEST 2 ADDENDUM: MINIMUM DETECTABLE EFFECT AND CME POSITIVE CONTROL
#  (a) MDE via Noether normal approximation for Wilcoxon-Mann-Whitney,
#      AUC scale, two-sided alpha 0.05, power 0.80.
#  (b) E-box (ACGTG, either strand) substituted for Motif2 as predictor
#      of |log2FC|, same pipeline.
# ============================================================
suppressPackageStartupMessages({
  library(BSgenome.Dmelanogaster.UCSC.dm6); library(Biostrings)
})
stopifnot(exists("Trh_Tgo_joint_filt"), length(Trh_Tgo_joint_filt) == 614L,
          exists("pk"), exists("fr"))

##, 1. CME flag per peak: ACGTG on either strand (Chapter 3 definition) -----
.seqs   <- getSeq(BSgenome.Dmelanogaster.UCSC.dm6, Trh_Tgo_joint_filt)
has_cme <- vcountPattern("ACGTG", .seqs) > 0 |
           vcountPattern("CACGT", .seqs) > 0        # revcomp of ACGTG
# Check: must reproduce the Chapter 3 CME-positive count exactly.
stopifnot(sum(has_cme) == 186L)
pk$has_cme <- has_cme[queryHits(distanceToNearest(
  Trh_Tgo_joint_filt,
  { .g <- suppressMessages(as.data.frame(genes(txdb)))
    .g <- .g[as.character(.g$seqnames) %in% good_chr, ]
    GRanges(.g$seqnames,
            IRanges(ifelse(.g$strand == "+", .g$start, .g$end),
                    ifelse(.g$strand == "+", .g$start, .g$end)),
            strand = .g$strand) },
  ignore.strand = TRUE))]

##, 2. Generalised stratum builder (flag column by name) --------------------
build_flag <- function(dist_lo, dist_hi, flagcol) {
  s  <- pk[pk$dist >= dist_lo & pk$dist <= dist_hi, ]
  gm <- tapply(s[[flagcol]], s$gene_id, any)
  gs <- tapply(s$signal,     s$gene_id, max)
  d  <- data.frame(gene_id = names(gm), flag = as.logical(gm),
                   peak_signal = as.numeric(gs[names(gm)]),
                   stringsAsFactors = FALSE)
  d  <- merge(d, fr[, c("gene_id", "log2FC_6.8")], by = "gene_id", all.x = TRUE)
  d  <- d[!is.na(d$log2FC_6.8), ]
  d$abs68 <- abs(d$log2FC_6.8)
  d
}

##, 3. MDE (Noether) and observed AUC ---------------------------------------
mde_auc <- function(n1, n2, alpha = 0.05, power = 0.80)
  0.5 + (qnorm(1 - alpha/2) + qnorm(power)) / sqrt(12 * n1 * n2 / (n1 + n2))
auc_obs <- function(x, y) {
  r <- rank(c(x, y))
  (sum(r[seq_along(x)]) - length(x) * (length(x) + 1) / 2) / (length(x) * length(y))
}
summarise_flag <- function(d) {
  x <- d$abs68[d$flag]; y <- d$abs68[!d$flag]
  w <- suppressWarnings(wilcox.test(x, y))
  list(n_pos = length(x), n_neg = length(y),
       med_pos = median(x), med_neg = median(y), p = unname(w$p.value),
       auc = auc_obs(x, y), mde = mde_auc(length(x), length(y)))
}

t2_m2_prox <- summarise_flag(build_flag(0L,    2000L, "has_motif2"))
t2_m2_dist <- summarise_flag(build_flag(2001L, Inf,   "has_motif2"))
t2_m2_all  <- summarise_flag(build_flag(0L,    Inf,   "has_motif2"))
t2_cme_prox <- summarise_flag(build_flag(0L,    2000L, "has_cme"))
t2_cme_dist <- summarise_flag(build_flag(2001L, Inf,   "has_cme"))
t2_cme_all  <- summarise_flag(build_flag(0L,    Inf,   "has_cme"))

# Check: Motif2 group sizes must reproduce the values already in the chapter.
stopifnot(t2_m2_prox$n_pos == 85L, t2_m2_prox$n_neg == 340L,
          t2_m2_dist$n_pos == 71L, t2_m2_dist$n_neg == 53L)

cat("\n== TEST 2 ADDENDUM: MDE and CME positive control ==\n")
for (nm in c("t2_m2_prox","t2_m2_dist","t2_m2_all",
             "t2_cme_prox","t2_cme_dist","t2_cme_all")) {
  r <- get(nm)
  cat(sprintf("  %-12s n+=%3d n-=%3d  p=%.3f  AUC=%.3f  MDE(80%%)=%.3f\n",
              nm, r$n_pos, r$n_neg, r$p, r$auc, r$mde))
}
```


```{r}
# ============================================================
# Appendix figures for Chapter 5 microarray QC
# ------------------------------------------------------------
# Two 3+2 tiles  (Limma_fdr, stdev_vs_mean), single-panel inputs only.
# Ten individual stage PNGs (SampleBoxplot, MAplotsp1), copy through unchanged.
# ============================================================

# Directory holding the FlyChip microarray QC images R{11-15}.vsn.*.png.
#
# These files are inputs to the tiles and copies composed from them, so they
# are kept with the other Chapter 5 inputs rather than in FIGURES_DIR.
MARRAY_QC_PNG_DIR <- file.path(PROJECT_DIR, "data", "chapter5",
                                   "microarray_qc_images")
if (!dir.exists(MARRAY_QC_PNG_DIR))
  stop("missing microarray QC image directory ", MARRAY_QC_PNG_DIR,
       " -- the 20 R{11-15}.vsn.{Limma.fdr,stdev_vs_mean,",
       "Sample-Boxplot,MAplots-p1}.png files are there")
.need <- file.path(MARRAY_QC_PNG_DIR,
                   sprintf("R%d.vsn.%s.png", rep(11:15, 4),
                           rep(c("Limma.fdr", "stdev_vs_mean",
                                 "Sample-Boxplot", "MAplots-p1"), each = 5)))
if (any(!file.exists(.need)))
  stop("missing microarray QC image(s):\n  ",
       paste(basename(.need[!file.exists(.need)]), collapse = "\n  "))

# ---- Helper: 3+2 tile (only safe for single-panel input PNGs) ----
make_tile_3_2 <- function(panel_paths, out_path) {
  stopifnot(length(panel_paths) == 5)
  missing <- panel_paths[!file.exists(panel_paths)]
  if (length(missing) > 0L) {
    stop("Missing source PNG(s):\n  ", paste(missing, collapse = "\n  "))
  }
  if (!requireNamespace("magick", quietly = TRUE)) {
    stop("Install the 'magick' R package: install.packages('magick')")
  }
  imgs   <- magick::image_read(panel_paths)
  w      <- magick::image_info(imgs[1])$width
  h      <- magick::image_info(imgs[1])$height
  filler <- magick::image_blank(w, h, color = "white")
  top    <- magick::image_append(imgs[1:3])
  bottom <- magick::image_append(c(imgs[4:5], filler))
  magick::image_write(magick::image_append(c(top, bottom), stack = TRUE),
                      path = out_path, format = "png")
  invisible(out_path)
}

# ---- (1) Two tiles ----
tile_specs <- list(
  list(stem = "Limma.fdr",     out = "figH_1_volcano_st11to15.png"),
  list(stem = "stdev_vs_mean", out = "figH_2_meanvar_st11to15.png")
)

for (spec in tile_specs) {
  src <- file.path(MARRAY_QC_PNG_DIR,
                   sprintf("R%d.vsn.%s.png", 11:15, spec$stem))
  out <- file.path(FIGURES_DIR, spec$out)
  make_tile_3_2(src, out)
  message("Wrote tile: ", out)
}

# ---- (2) Individual per-stage figures (no resampling, just copy) ----
## The appendix figure index and the stage are derived together here:
## boxplots are H.8-H.12 for stages 11-15, MA-plots H.3-H.7.
individual_specs <- list(
  list(stem = "Sample-Boxplot", off = -3L, out_fmt = "figH_%d_boxplot_norm_st%d.png"),
  list(stem = "MAplots-p1",     off = -8L, out_fmt = "figH_%d_maplot_st%d.png")
)

for (spec in individual_specs) {
  for (s in 11:15) {
    src <- file.path(MARRAY_QC_PNG_DIR,
                     sprintf("R%d.vsn.%s.png", s, spec$stem))
    out <- file.path(FIGURES_DIR, sprintf(spec$out_fmt, s + spec$off, s))
    stopifnot(file.exists(src))
    file.copy(src, out, overwrite = TRUE)
    message("Copied: ", out)
  }
}

old_tile <- file.path(FIGURES_DIR, "fig5_app_lm_diag.png")
if (file.exists(old_tile)) file.remove(old_tile)
```

```{r ch5-fig-app-lm-diag, eval=FALSE}
# ============================================================
#  Per-stage lm diagnostics (plot.lm) for lm(microarray ~ RTqPCR), all 16 validation genes, stages 11-15; outputs figJ_1_lm_diag_st11..st15.png; standalone, loads project paths; data as Figure 5.5 (R{st}.vsn.limma.fdr.txt.gz by TargetID, st{st}_validation_unstacked.csv)
# ============================================================
if (!exists("FIGURES_DIR"))
CH5_INPUT_DIR <- CH5_DIR          # data/chapter5 (limma + RT-qPCR)
LIMMA_DIR     <- CH5_INPUT_DIR
stopifnot(dir.exists(CH5_INPUT_DIR), dir.exists(FIGURES_DIR))

stage_ids <- 11:15

# 16 validation genes, identified by microarray probe (TargetID), single
# source of truth linking RT-qPCR gene names to limma probe rows.
probe_map <- c(
  "BicC"     = "CG4824-RA",  "PGRP-SB1" = "CG9681-RA",  "Cyp6a19" = "CG10243-RA",
  "dhd"      = "CG4193-RA",  "AsnS"     = "CG33486-RA", "Ect3"    = "CG3132-RA",
  "Phk-3"    = "CG9358-RA",  "Cpr51A"   = "CG10112-RA", "trh"     = "CG6883-RA",
  "Cyp12a4"  = "CG6042-RA",  "Cyp309a2" = "CG18559-RA", "ninaD"   = "CG31783-RA",
  "CG14102"  = "CG14102-RA", "Cdc42"    = "CG12530-RB", "Sli"     = "CG8355-RC",
  "Ir7b"     = "CG15326-RA"
)

read_marray_stage <- function(st) {
  f <- file.path(LIMMA_DIR, sprintf("R%d.vsn.limma.fdr.txt.gz", st))
  stopifnot(file.exists(f))
  lim <- read.delim(f, skip = 1, check.names = FALSE, stringsAsFactors = FALSE)
  stopifnot(all(c("TargetID", "logFC", "t") %in% names(lim)))
  idx <- match(probe_map, lim$TargetID)
  stopifnot(!any(is.na(idx)))
  data.frame(gene = names(probe_map),
             microarray = as.numeric(lim$logFC[idx]),
             row.names = NULL, stringsAsFactors = FALSE)
}

read_rtqpcr_stage <- function(st) {
  f <- file.path(CH5_INPUT_DIR, sprintf("st%d_validation_unstacked.csv", st))
  stopifnot(file.exists(f))
  d <- read.csv(f, row.names = 1, check.names = FALSE, fileEncoding = "UTF-8-BOM")
  names(d)[names(d) %in% c("RT-qPCR", "RT.qPCR")] <- "RTqPCR"
  data.frame(gene = rownames(d), RTqPCR = as.numeric(d$RTqPCR),
             row.names = NULL, stringsAsFactors = FALSE)
}

for (st in stage_ids) {
  d <- merge(read_marray_stage(st), read_rtqpcr_stage(st), by = "gene")
  stopifnot(nrow(d) == 16L)                      # all 16 genes, no exclusions
  fit <- lm(microarray ~ RTqPCR, data = d)

  # NAMING: figures are written as "figJ_{1..5}_lm_diag_st{11..15}.png".
  # The panel index and the stage are both derived together here, so the
  # numbering in the filename stays consistent with the stage being plotted.
  out <- file.path(FIGURES_DIR,
                   sprintf("figJ_%d_lm_diag_st%d.png", st - 10L, st))
  png(out, width = 8, height = 8, units = "in", res = 300)
  op <- par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.6, 1.2), oma = c(0, 0, 2.2, 0))
  # Panels 1,2,3,5 = residual-fitted, normal Q-Q, scale-location, residual-leverage
  # (not which=1:4, which would give Cook's distance instead of residual-leverage).
  plot(fit, which = c(1L, 2L, 3L, 5L), sub.caption = "")  # suppress default formula caption
  mtext(sprintf("Stage %d  (n = %d, all 16 validation genes)", st, nrow(d)),
        outer = TRUE, cex = 1.0, font = 2)
  par(op)
  dev.off()
  cat(sprintf("Wrote %s\n", out))
}
```

```{r ch5-export-computed-values, eval=FALSE}
# ============================================================
# CHAPTER 5, record the reported values in results/reported_values.tsv
#
# Run this as the last chunk, after every compute chunk. All objects below
# (m13/m14, stat13/stat14, valid_st13/valid_st14, trh_expression, rtqpcr_stats,
#  ft_2way, direct_targets_trh, n_retained_tissue, ...) must exist in session.
# ============================================================

stopifnot(exists("save_values"), length(Trh_Tgo_joint_filt) == 614L)

ch5_vals <- list(

  # ---- Genome-wide microarray vs Fisher concordance (was CONCORDANCE BLOCK) ----
  nMergedStThirteen              = nrow(m13),
  nMergedStFourteen              = nrow(m14),
  nMergedFisherMapped            = n_fisher_mapped,
  nMergedFisherTotal             = n_fisher_total,
  rhoGwStThirteenFisher          = sprintf("%.2f", stat13$mid$rho),
  rhoGwStFourteenFisher          = sprintf("%.2f", stat14$mid$rho),
  rhoGwStThirteenFisherEarly     = sprintf("%.2f", stat13$early$rho),
  rhoGwStFourteenFisherEarly     = sprintf("%.2f", stat14$early$rho),
  rhoGwStThirteenFisherLate      = sprintf("%.2f", stat13$late$rho),
  rhoGwStFourteenFisherLate      = sprintf("%.2f", stat14$late$rho),
  signConcordGwStThirteen        = sprintf("%.0f", 100 * stat13$mid$sign_concord),
  signConcordGwStFourteen        = sprintf("%.0f", 100 * stat14$mid$sign_concord),
  nValidatedStThirteen           = length(valid_st13),
  nValidatedStFourteen           = length(valid_st14),
  nValidatedUnion                = length(union(valid_st13, valid_st14)),
  nValidatedIntersection         = length(intersect(valid_st13, valid_st14)),
  signConcordValidatedStThirteen = sprintf("%.0f", 100 * sign_c_valid_st13),
  signConcordValidatedStFourteen = sprintf("%.0f", 100 * sign_c_valid_st14),

  # ---- Direct targets, RT-qPCR, Rib comparison (was strip-append block) ----
  nFisherDegs                    = length(fisher_degs),
  nMarrayValidated               = marray_validated_total,
  nMarrayInFisher                = marray_in_fisher_total,
  pctMarrayInFisher              = round(100 * marray_in_fisher_total / marray_validated_total, 1),
  marrayDataSource               = marray_source,
  ##, trh knock-down efficacy, RT-qPCR (the efficacy measurement) -----------
  trhQpcrStEleven       = sprintf("%.2f", trh_qpcr$log2FC[1]),
  trhQpcrStTwelve       = sprintf("%.2f", trh_qpcr$log2FC[2]),
  trhQpcrStThirteen     = sprintf("%.2f", trh_qpcr$log2FC[3]),
  trhQpcrStFourteen     = sprintf("%.2f", trh_qpcr$log2FC[4]),
  trhQpcrStFifteen      = sprintf("%.2f", trh_qpcr$log2FC[5]),
  trhQpcrFoldStFifteen  = sprintf("%.1f", 1 / 2^trh_qpcr$log2FC[5]),
  trhQpcrFoldStFourteen = sprintf("%.1f", 1 / 2^trh_qpcr$log2FC[4]),
  # Stage-agnostic: the deepest depletion and the stage at which it occurs are
  # derived from the data, never asserted, so trhQpcrFoldStFifteen is not used
  # to state "the deepest" anywhere.
  trhQpcrFoldMax        = sprintf("%.2f", 1 / 2^min(trh_qpcr$log2FC)),
  trhQpcrStageMax       = sprintf("%d",   trh_qpcr$stage[which.min(trh_qpcr$log2FC)]),
  trhQpcrRho            = sprintf("%.2f", unname(ct_stage$estimate)),
  trhQpcrRhoP           = sprintf("%.3f", ct_stage$p.value),
  # Per-stage inference is possible because the raw Ct workbook carries three
  # independent preparations per arm per stage.
  trhQpcrPStThirteen    = sprintf("%.3f", trh_qpcr$p_welch[3]),
  trhQpcrPMinOther      = sprintf("%.2f", min(trh_qpcr$p_welch[-3])),
  trhQpcrPermStThirteen = sprintf("%.2f", trh_qpcr$p_perm[3]),
  trhQpcrPermFloor      = sprintf("%.2f", 2 / choose(6, 3)),
  trhQpcrSeMin          = sprintf("%.2f", min(trh_qpcr$SE)),
  trhQpcrSeMax          = sprintf("%.2f", max(trh_qpcr$SE)),
  ##, trh on the array (secondary trace; not the efficacy claim) ------------
  trhArrayStEleven      = sprintf("%+.2f", trh_array$log2FC[1]),
  trhArrayStTwelve      = sprintf("%+.2f", trh_array$log2FC[2]),
  trhArrayStThirteen    = sprintf("%+.2f", trh_array$log2FC[3]),
  trhArrayStFourteen    = sprintf("%+.2f", trh_array$log2FC[4]),
  trhArrayStFifteen     = sprintf("%+.2f", trh_array$log2FC[5]),
  trhArraySEStEleven    = sprintf("%.2f", trh_array$SE[1]),
  trhArraySEStTwelve    = sprintf("%.2f", trh_array$SE[2]),
  trhArraySEStThirteen  = sprintf("%.2f", trh_array$SE[3]),
  trhArraySEStFourteen  = sprintf("%.2f", trh_array$SE[4]),
  trhArraySEStFifteen   = sprintf("%.2f", trh_array$SE[5]),
  trhArrayPadjMin       = sprintf("%.2f", min(trh_array$padj)),
  trhArrayNPositive     = sum(trh_array$log2FC > 0),
  ##, Fisher's own trh knock-down, for scale against the array's floor ------
  # fr is built in the TEST 2 chunk, which runs before this export chunk.
  fisherTrhLogFC        = sprintf("%.2f", fr$log2FC_6.8[match("trh", fr$symbol_fisher)]),
  ##, the array's single-probe detection floor (bounded negative) -----------
  trhArrayMdeFold       = format(round(mde_fold), scientific = FALSE, big.mark = ""),
  trhArrayMdeLogTwo     = sprintf("%.1f", mde_log2),
  trhArrayMdeTCrit      = sprintf("%.0f", t_crit),
  trhArrayMdeNArrays    = 3L,
  ##, gene-level pass counts -----------------------------------------------
  nMarrayPassStEleven   = n_pass_fold[1],
  nMarrayPassStTwelve   = n_pass_fold[2],
  nMarrayPassStThirteen = n_pass_fold[3],
  nMarrayPassStFourteen = n_pass_fold[4],
  nMarrayPassStFifteen  = n_pass_fold[5],
  # Rank correlation between depletion DEPTH (-log2FC, so positive = depleted)
  # and per-stage DE gene count. Derived here so the value cannot drift from
  # the underlying data.
  marrayCountDepletionRho = sprintf("%+.2f", unname(suppressWarnings(
      cor.test(-trh_qpcr$log2FC, n_pass_fold, method = "spearman"))$estimate)),
  marrayCountDepletionP   = sprintf("%.2f", suppressWarnings(
      cor.test(-trh_qpcr$log2FC, n_pass_fold, method = "spearman"))$p.value),
  ##, padj-only counts -----------------------------------------------------
  nMarrayPadjOnlyStFifteen = n_pass_padj[5],
  ##, depletion vs cross-platform concordance (exact Spearman, n = 5) -------
  trhDepletionConcRho   = sprintf("%.2f", unname(ct_conc$estimate)),
  trhDepletionConcP     = sprintf("%.3f", ct_conc$p.value),
  rtqpcrRsStEleven               = sprintf("%.2f", rtqpcr_stats$Rs[1]),
  rtqpcrPStEleven                = sprintf("%.3f", rtqpcr_stats$p[1]),
  rtqpcrRsStTwelve               = sprintf("%.2f", rtqpcr_stats$Rs[2]),
  rtqpcrPStTwelve                = sprintf("%.3f", rtqpcr_stats$p[2]),
  rtqpcrRsStThirteen             = sprintf("%.2f", rtqpcr_stats$Rs[3]),
  rtqpcrPStThirteen              = sprintf("%.3f", rtqpcr_stats$p[3]),
  rtqpcrRsStFourteen             = sprintf("%.2f", rtqpcr_stats$Rs[4]),
  rtqpcrPStFourteen              = sprintf("%.3f", rtqpcr_stats$p[4]),
  rtqpcrRsStFifteen              = sprintf("%.2f", rtqpcr_stats$Rs[5]),
  rtqpcrPStFifteen               = sprintf("%.3f", rtqpcr_stats$p[5]),
  nRibTargets                    = length(rib_target_genes),
  nTrhChipPeaksAnnot             = nrow(peak_genes_in_window),
  nTrhChipGenes                  = length(trh_chip_genes),
  nDirectTargetsTrh              = length(direct_targets_trh),
  directTargetsTrhFold           = sprintf("%.2f", fold_enrich_trh),
  directTargetsTrhP              = fmt_p(p_hyper_trh),
  # Universe-restricted test (the valid one: draw is a subset of the urn)
  nChipGenesInUniv               = u_n,
  nChipGenesOutUniv              = length(trh_chip_genes) - u_n,
  nFisherDegsInUniv              = u_K,
  nDirectTargetsInUniv           = u_k,
  dtUnivExpected                 = sprintf("%.1f", u_exp),
  dtUnivFold                     = sprintf("%.2f", u_fold),
  dtUnivP                        = sprintf("%.3f", u_pdep),
  dtUnivOR                       = sprintf("%.2f", u_ft$estimate),
  dtUnivCIlo                     = sprintf("%.2f", u_ft$conf.int[1]),
  dtUnivCIhi                     = sprintf("%.2f", u_ft$conf.int[2]),
  dtUnivPowerTwoFold             = sprintf("%.2f", u_pow2),
  nDirectTargetsFiltered         = ifelse(is.na(n_retained_tissue), "??", n_retained_tissue),
  nDirectTargetsExcluded         = ifelse(is.na(n_excluded_tissue), "??", n_excluded_tissue),
  nTrhRibOverlap                 = n_trh_rib_overlap,
  trhRibOR                       = sprintf("%.2f", unname(ft_2way$estimate)),
  trhRibP                        = fmt_p(ft_2way$p.value),
  nThreeWayOverlap               = n_three_way,

  # ---- Chung 2011 (GSE28780) second perturbation axis (ch5-chung2011-gse28780) ----
  chungWTReps           = 3L,
  chungNullReps         = 3L,
  chungUniverse         = length(chung_universe),
  chungTrhActivated     = length(intersect(chung_sets$down, chung_universe)),
  chungTrhRepressed     = length(intersect(chung_sets$up,   chung_universe)),
  chungTrhResponsive    = length(intersect(chung_sets$any,  chung_universe)),
  chungBoundOnArray     = length(intersect(chung_bound, chung_universe)),
  chungProxGenes        = length(intersect(chung_prox,  chung_universe)),
  chungDistGenes        = length(intersect(chung_dist,  chung_universe)),
  chungProxRespFold     = sprintf("%.2f", chung_get("prox","any","fold")),
  chungProxRespP        = fmt_p(chung_get("prox","any","p")),
  chungDistRespFold     = sprintf("%.2f", chung_get("dist","any","fold")),
  chungDistRespP        = fmt_p(chung_get("dist","any","p")),
  chungDistReprFold     = sprintf("%.2f", chung_get("dist","up","fold")),
  chungDistReprP        = fmt_p(chung_get("dist","up","p")),
  chungDistActFold      = sprintf("%.2f", chung_get("dist","down","fold")),
  chungDistActP         = fmt_p(chung_get("dist","down","p")),
  chungConjunction      = length(chung_conj_any),
  chungConjunctionAct   = length(chung_conj_down),
  chungDistFoldExprLo   = sprintf("%.2f", min(chung_robust$dist_fold)),
  chungDistFoldExprHi   = sprintf("%.2f", max(chung_robust$dist_fold)),
  chungDistPExprWorst   = fmt_p(max(chung_robust$dist_p)),

  #, TEST 2: Motif2 functional-consequence test (fig5_7) --
  nPeaksMotifTwo        = test2_n_peaks_motif2,
  testTwoProxP          = sprintf("%.2f", test2_prox_p),
  testTwoProxMedPos     = sprintf("%.3f", test2_prox_med_pos),
  testTwoProxMedNeg     = sprintf("%.3f", test2_prox_med_neg),
  testTwoProxNPos       = test2_prox_n_pos,
  testTwoProxNNeg       = test2_prox_n_neg,
  testTwoDistP          = sprintf("%.2f", test2_dist_p),
  testTwoDistMedPos     = sprintf("%.3f", test2_dist_med_pos),
  testTwoDistMedNeg     = sprintf("%.3f", test2_dist_med_neg),
  testTwoDistNPos       = test2_dist_n_pos,
  testTwoDistNNeg       = test2_dist_n_neg,
  testTwoAllP           = sprintf("%.2f", test2_all_p),
  testTwoLmP            = sprintf("%.2f", test2_lm_p),
  testTwoMdeProx                 = sprintf("%.2f", t2_m2_prox$mde),
  testTwoMdeDist                 = sprintf("%.2f", t2_m2_dist$mde),
  testTwoMdeAll                  = sprintf("%.2f", t2_m2_all$mde),
  testTwoAucProx                 = sprintf("%.2f", t2_m2_prox$auc),
  testTwoAucDist                 = sprintf("%.2f", t2_m2_dist$auc),
  testTwoAucAll                  = sprintf("%.2f", t2_m2_all$auc),

  # ---- TEST 2 addendum: CME positive control ----
  testTwoCmeProxNPos             = t2_cme_prox$n_pos,
  testTwoCmeProxNNeg             = t2_cme_prox$n_neg,
  testTwoCmeProxP                = sprintf("%.2f", t2_cme_prox$p),
  testTwoCmeDistNPos             = t2_cme_dist$n_pos,
  testTwoCmeDistNNeg             = t2_cme_dist$n_neg,
  testTwoCmeDistP                = sprintf("%.2f", t2_cme_dist$p),
  testTwoCmeAllNPos              = t2_cme_all$n_pos,
  testTwoCmeAllNNeg              = t2_cme_all$n_neg,
  testTwoCmeAllP                 = sprintf("%.2f", t2_cme_all$p),
  testTwoCmeAucAll               = sprintf("%.2f", t2_cme_all$auc),
  nConjConcordant                = n_conj_concordant,
  nConjDiscordant                = n_conj_discordant,
  nConjProximal                  = n_conj_proximal,
  nConjDistal                    = n_conj_distal,
  nConjDistalNovel               = n_conj_distal_novel
)

save_values("ch5", ch5_vals)
```
