
# Control transcription factors (Chapter 4)

Four non-bHLH-PAS control factors are scanned for Motif2 under the same
pipeline used for the four heterodimers in `chapter4_heterodimer_comparison.Rmd`:

- TWI (Twist): bHLH without PAS, mesoderm (single-end)
- VVL (Ventral veinless): POU-homeodomain, tracheal (paired-end)
- RIB (Ribbon): BTB/POZ zinc finger, morphogenesis (paired-end)
- DA (Daughterless): bHLH class I hub, the obligate partner of the class II
  bHLH factors (single-end); peaks from `01_chipseq_processing/peak_calling_idr/production_peaks_da.sh`

Run these chunks after `chapter4_heterodimer_comparison.Rmd`, in the same R
session: the heterodimer peak sets (`Trh_Tgo_joint_filt`, `Sim_Tgo_joint_filt`,
`Dys_Tgo_joint_filt`, `Sima_Tgo_joint_filt`, with their `$Motif2` and
`$overlaps_enhancer` columns), the enhancer set `enhancers_filt`, `EXPORT_DIR`,
`FIGURES_DIR` and `fmt_p` are taken from that session. The control
Motif2 panel reported in the text is computed by
`chapter4/ch4_ctrl_motif2_panel.R` from the FASTA and FIMO files written here.

```{r ctrl-prereq-check, include=TRUE, eval=TRUE}
# ---- Stop if any prerequisite is missing ----
required_objs <- c(
  # Heterodimer peak sets
  "Trh_Tgo_joint_filt",
  "Sim_Tgo_joint_filt",
  "Dys_Tgo_joint_filt",
  "Sima_Tgo_joint_filt",
  # Reference enhancer set
  "enhancers_filt",
  # Paths and helpers
  "PROJECT_DIR", "EXPORT_DIR", "FIGURES_DIR",
  "fmt_p", "good_chr"
)

missing_objs <- setdiff(required_objs, ls(envir = .GlobalEnv))
if (length(missing_objs) > 0L) {
  stop("Missing prerequisites from main Chapter 4 environment: ",
       paste(missing_objs, collapse = ", "),
       "\nSource or knit the main Chapter 4 Rmd before running these chunks.")
}
# Before this file, run in the Chapter 4 Rmd:
# {r annotate-distance-to-tss-all-heterodimers} ->
# {r regulatory-classification} ->
# {r cme-occurrence} ->
# {r load-fimo-heterodimers}

# Confirm the heterodimer peak sets carry the metadata columns we need
for (nm in c("Trh_Tgo_joint_filt", "Sim_Tgo_joint_filt",
             "Dys_Tgo_joint_filt", "Sima_Tgo_joint_filt")) {
  gr <- get(nm)
  stopifnot(inherits(gr, "GRanges"))
  for (col in c("Motif2", "overlaps_enhancer")) {
    if (!col %in% names(mcols(gr))) {
      stop(sprintf(
  "Object %s lacks metadata column $%s; run the main Chapter 4 motif/enhancer-overlap chunks first.",
  nm,
  col
))
    }
  }
}

cat("All Chapter 4 prerequisites present.\n")
cat(sprintf("  Trh_Tgo_joint_filt:  %d peaks\n", length(Trh_Tgo_joint_filt)))
cat(sprintf("  Sim_Tgo_joint_filt:  %d peaks\n", length(Sim_Tgo_joint_filt)))
cat(sprintf("  Dys_Tgo_joint_filt:  %d peaks\n", length(Dys_Tgo_joint_filt)))
cat(sprintf("  Sima_Tgo_joint_filt: %d peaks\n", length(Sima_Tgo_joint_filt)))
cat(sprintf("  enhancers_filt:      %d enhancers\n", length(enhancers_filt)))
```

---

```{r ctrl-setup, include=TRUE, eval=TRUE}
# ---- Paths for the controls extension ----
## Repository root: the environment variable PROJECT_DIR, or the working directory.
PROJECT_DIR <- normalizePath(Sys.getenv("PROJECT_DIR", unset = "."), mustWork = TRUE)
if (!file.exists(file.path(PROJECT_DIR, "README.md")))
  stop("PROJECT_DIR does not point at the repository root: ", PROJECT_DIR)

DATA_DIR      <- file.path(PROJECT_DIR, "data")
RESULTS_DIR   <- file.path(PROJECT_DIR, "results")
FIGURES_DIR   <- file.path(RESULTS_DIR, "figures")
CH4_DIR <- file.path(DATA_DIR, "chapter4")

good_chr <- c("chr2L", "chr2R", "chr3L", "chr3R", "chr4", "chrX")   # dm6 major arms

dir.create(FIGURES_DIR, recursive = TRUE, showWarnings = FALSE)

CTRL_FINAL_DIR <- file.path(PROJECT_DIR, "data/final_peaks")
CTRL_QC_DIR    <- file.path(PROJECT_DIR, "04_qc")
CTRL_FIMO_ROOT <- file.path(PROJECT_DIR, "data/ctrl_tf", "fimo")
CTRL_FASTA_DIR <- file.path(CTRL_FIMO_ROOT, "input_fastas")

dir.create(CTRL_FIMO_ROOT, recursive = TRUE, showWarnings = FALSE)
dir.create(CTRL_FASTA_DIR, recursive = TRUE, showWarnings = FALSE)

suppressPackageStartupMessages({
  library(GenomicRanges)
  library(rtracklayer)
  library(Biostrings)
  library(BSgenome.Dmelanogaster.UCSC.dm6)
})
# ---- Helper: read narrowPeak ----
read_narrowpeak <- function(path) {
  if (!file.exists(path)) stop("narrowPeak file not found: ", path)
  df <- read.table(path, sep = "\t", header = FALSE, stringsAsFactors = FALSE)
  GRanges(
    seqnames    = df[[1]],
    ranges      = IRanges(start = df[[2]] + 1L, end = df[[3]]),
    score       = df[[5]],
    signalValue = df[[7]],
    pValue      = df[[8]],
    qValue      = df[[9]]
  )
}

load_peaks <- function(fname) {
  gr <- read_narrowpeak(file.path(CTRL_FINAL_DIR, fname))
  gr <- keepSeqlevels(gr,
                      intersect(good_chr, seqlevels(gr)),
                      pruning.mode = "coarse")
  sort(gr)
}

# ---- Final IDR peak sets for the four retained controls ----
TWI_peaks <- load_peaks("TWI_IDR0.05.final.narrowPeak")
VVL_peaks <- load_peaks("VVL_IDR0.05.final.narrowPeak")
RIB_peaks <- load_peaks("RIB_IDR0.05.final.narrowPeak")
DA_peaks  <- load_peaks("DA_IDR0.05.final.narrowPeak")

# ---- Annotate enhancer overlap on each control peak set ----
# Reuses the same column name ($overlaps_enhancer) as the heterodimers.
annotate_enhancer_overlap_local <- function(gr, enhancers) {
  gr$overlaps_enhancer <- countOverlaps(gr, enhancers,
                                        ignore.strand = TRUE) > 0
  gr
}

TWI_peaks <- annotate_enhancer_overlap_local(TWI_peaks, enhancers_filt)
VVL_peaks <- annotate_enhancer_overlap_local(VVL_peaks, enhancers_filt)
RIB_peaks <- annotate_enhancer_overlap_local(RIB_peaks, enhancers_filt)
DA_peaks  <- annotate_enhancer_overlap_local(DA_peaks,  enhancers_filt)

cat(sprintf("TWI: %d peaks (%d enhancer-overlapping)\n",
            length(TWI_peaks), sum(TWI_peaks$overlaps_enhancer)))
cat(sprintf("VVL: %d peaks (%d enhancer-overlapping)\n",
            length(VVL_peaks), sum(VVL_peaks$overlaps_enhancer)))
cat(sprintf("RIB: %d peaks (%d enhancer-overlapping)\n",
            length(RIB_peaks), sum(RIB_peaks$overlaps_enhancer)))
cat(sprintf("DA: %d peaks (%d enhancer-overlapping)\n",
            length(DA_peaks), sum(DA_peaks$overlaps_enhancer)))
stopifnot(length(DA_peaks) == 923L, sum(DA_peaks$overlaps_enhancer) == 264L)
```

---

```{r ctrl-qc-table, include=TRUE, eval=TRUE}
# QC summary table for four retained controls.
# NSC, RSC, fragment estimate from phantompeakqualtools (column 3 of .spp_metrics.txt or _spp.clean.txt, first comma-separated value); FRiP and median peak width from production peak-calling SLURM log (DA: production_peaks_da.sh log).
# NSC/RSC thresholds: 1.05, 0.8.

qc_ctrl <- data.frame(
  Factor      = c("TWI", "TWI", "VVL", "VVL", "RIB", "RIB", "DA",  "DA"),
  Replicate   = c("rep1", "rep2", "rep1", "rep2", "rep1", "rep2", "rep1", "rep2"),
  Library     = c("SE",   "SE",   "PE",   "PE",   "PE",   "PE",   "SE",  "SE"),
  Frag_est    = c(210L,   230L,   190L,   160L,   175L,   180L,   200L,  180L),
  NSC         = c(1.017,  1.033,  1.060,  1.057,  1.030,  1.032,  1.012, 1.013),
  RSC         = c(1.027,  1.335,  1.748,  1.487,  1.802,  2.288,  0.916, 0.974),
  Quality_tag = c(1L,     1L,     2L,     1L,     2L,     2L,     0L,    0L),
  FRiP        = c(0.0847, 0.1063,    # TWI: from the production_peaks_twi.sh output
                  0.0241,   0.0226,      # VVL
                  0.0108,   0.0104,      # RIB
                  0.0148,   0.0137),     # DA: from the production_peaks_da.sh output
  stringsAsFactors = FALSE
)

ctrl_factor_summary <- data.frame(
  Factor          = c("TWI", "VVL", "RIB", "DA"),
  N_final_peaks   = c(length(TWI_peaks), length(VVL_peaks), length(RIB_peaks), length(DA_peaks)),
  Median_width_bp = c(median(width(TWI_peaks)),
                      median(width(VVL_peaks)),
                      median(width(RIB_peaks)),
                      median(width(DA_peaks))),
  N_enh_peaks     = c(sum(TWI_peaks$overlaps_enhancer),
                      sum(VVL_peaks$overlaps_enhancer),
                      sum(RIB_peaks$overlaps_enhancer),
                      sum(DA_peaks$overlaps_enhancer)),
  stringsAsFactors = FALSE
)

cat("\n=== Per-replicate QC ===\n")
print(qc_ctrl, row.names = FALSE)
cat("\n=== Per-factor summary ===\n")
print(ctrl_factor_summary, row.names = FALSE)
```

---

```{r ctrl-export-bed-for-fasta, include=TRUE, eval=TRUE}
# ---- Export enhancer-overlapping peaks as BED for FASTA extraction ----
# Uses the centre of each peak ±100 bp, matching the convention used in
# the main Chapter 4 FIMO scanning chunks (200 bp summit-centred windows).

write_centre_bed <- function(gr_full, out_path, half_window = 100L) {
  gr <- gr_full[gr_full$overlaps_enhancer]
  centre <- (start(gr) + end(gr)) %/% 2L
  df <- data.frame(
    chr   = as.character(seqnames(gr)),
    start = pmax(centre - half_window - 1L, 0L),  # 0-based BED
    end   = centre + half_window,
    name  = paste0("peak_", seq_along(gr))
  )
  write.table(df, out_path, sep = "\t", quote = FALSE,
              row.names = FALSE, col.names = FALSE)
  cat("Wrote", nrow(df), "regions to", out_path, "\n")
}

write_centre_bed(TWI_peaks, file.path(CTRL_FASTA_DIR, "TWI_enh_summits_200bp.bed"))
write_centre_bed(VVL_peaks, file.path(CTRL_FASTA_DIR, "VVL_enh_summits_200bp.bed"))
write_centre_bed(RIB_peaks, file.path(CTRL_FASTA_DIR, "RIB_enh_summits_200bp.bed"))
write_centre_bed(DA_peaks,  file.path(CTRL_FASTA_DIR, "DA_enh_summits_200bp.bed"))
```

---

```{bash ctrl-fasta-extract, include=TRUE, eval=FALSE}
#!/usr/bin/env bash
# ---- BED -> FASTA for the four controls ----
set -euo pipefail

# Use $HOME (or the absolute path), never bare ~ inside quoted strings.
PROJECT_DIR="${PROJECT_DIR:-$(pwd)}"
CTRL_FASTA_DIR="${PROJECT_DIR}/data/ctrl_tf/fimo/input_fastas"
DM6_FA="${PROJECT_DIR}/reference/dm6.fa"

# Pre-flight checks
[ -f "${DM6_FA}" ]                      || { echo "MISSING genome FASTA: ${DM6_FA}";  exit 1; }
[ -d "${CTRL_FASTA_DIR}" ]              || { echo "MISSING input dir: ${CTRL_FASTA_DIR}"; exit 1; }
for F in TWI VVL RIB DA; do
  [ -f "${CTRL_FASTA_DIR}/${F}_enh_summits_200bp.bed" ] || \
    { echo "MISSING BED for ${F}; run the R chunks first"; exit 1; }
done

for FACTOR in TWI VVL RIB DA
do
  bedtools getfasta \
    -fi "${DM6_FA}" \
    -bed "${CTRL_FASTA_DIR}/${FACTOR}_enh_summits_200bp.bed" \
    -fo  "${CTRL_FASTA_DIR}/${FACTOR}_enh_summits_200bp.fa" \
    -nameOnly
  # -nameOnly, not -name. `-name` emits ">peak_1::chr3L:106074-106275"; FIMO
  # parses the embedded locator and writes the CHROMOSOME into sequence_name,
  # so unique(sequence_name) counts chromosomes (6, 6, 5, 5) instead of peaks.
  echo "${FACTOR}: $(grep -c '^>' ${CTRL_FASTA_DIR}/${FACTOR}_enh_summits_200bp.fa) sequences"
done
```

---

```{bash ctrl-fimo-run, include=TRUE, eval=FALSE}
#!/usr/bin/env bash
# ---- Run FIMO Motif2 scan on the four controls ----
# Parameters identical to main Chapter 4 FIMO scanning of heterodimers:
# --thresh 1e-4, q-value < 0.05 applied downstream in R.

set -euo pipefail

PROJECT_DIR="${PROJECT_DIR:-$(pwd)}"
CTRL_FASTA_DIR="${PROJECT_DIR}/data/ctrl_tf/fimo/input_fastas"
CTRL_FIMO_ROOT="${PROJECT_DIR}/data/ctrl_tf/fimo"
MOTIF2_MEME="${PROJECT_DIR}/data/chapter3/STRCACTGARARAAA.meme"   # Motif2 PWM

[ -f "${MOTIF2_MEME}" ] || { echo "MISSING Motif2 .meme: ${MOTIF2_MEME}"; exit 1; }

for FACTOR in TWI VVL RIB DA
do
  OUTDIR="${CTRL_FIMO_ROOT}/${FACTOR}_motif2"
  mkdir -p "${OUTDIR}"
  fimo --thresh 1.0E-4 --oc "${OUTDIR}" \
       "${MOTIF2_MEME}" \
       "${CTRL_FASTA_DIR}/${FACTOR}_enh_summits_200bp.fa"
  echo "${FACTOR} FIMO complete -> ${OUTDIR}/fimo.tsv"
done
```

---

```{r ctrl-matched-bg, include=TRUE, eval=TRUE}
# ---- Build a GC- and length-matched non-Tgo enhancer background ----
# Central quantitative test: if Motif2 enrichment at the four heterodimers
# is heterodimer-specific, it should be ABSENT at non-Tgo-bound enhancers
# matched for GC composition and width.

# Pool all four heterodimer peak sets, take their union as "Tgo-bound"
all_hetero <- c(Trh_Tgo_joint_filt,
                Sim_Tgo_joint_filt,
                Dys_Tgo_joint_filt,
                Sima_Tgo_joint_filt)
# Strip the metadata columns to allow c() then reduce()
mcols(all_hetero) <- NULL
all_hetero <- reduce(all_hetero)

# Non-Tgo enhancers
non_tgo_enh <- enhancers_filt[
  countOverlaps(enhancers_filt, all_hetero, ignore.strand = TRUE) == 0L
]
cat("Non-Tgo enhancers available for matching:", length(non_tgo_enh), "\n")

# Reference distribution: GC and width of the 614 Trh/Tgo joint peaks
ref_seqs  <- getSeq(BSgenome.Dmelanogaster.UCSC.dm6, Trh_Tgo_joint_filt)
ref_gc    <- as.numeric(letterFrequency(ref_seqs, letters = "GC", as.prob = TRUE))
ref_width <- width(Trh_Tgo_joint_filt)

cand_seqs  <- getSeq(BSgenome.Dmelanogaster.UCSC.dm6, non_tgo_enh)
cand_gc    <- as.numeric(letterFrequency(cand_seqs, letters = "GC", as.prob = TRUE))
cand_width <- width(non_tgo_enh)

# Bin by GC (5 bins) x width (4 bins) = 20 strata
gc_breaks <- quantile(ref_gc,    probs = seq(0, 1, 0.2))
w_breaks  <- quantile(ref_width, probs = c(0, 0.25, 0.5, 0.75, 1))

bin_seq <- function(gc, w, gc_br, w_br) {
  paste(
    findInterval(gc, gc_br, rightmost.closed = TRUE),
    findInterval(w,  w_br,  rightmost.closed = TRUE),
    sep = "_"
  )
}
ref_bins  <- bin_seq(ref_gc,  ref_width,  gc_breaks, w_breaks)
cand_bins <- bin_seq(cand_gc, cand_width, gc_breaks, w_breaks)

set.seed(777)
matched_idx <- unlist(lapply(unique(ref_bins), function(b) {
  n_need <- sum(ref_bins == b)
  pool   <- which(cand_bins == b)
  if (length(pool) == 0L) return(integer(0))
  sample(pool, min(n_need, length(pool)), replace = FALSE)
}))

matched_bg <- non_tgo_enh[matched_idx]
cat("Matched background size:", length(matched_bg),
    "(target:", length(Trh_Tgo_joint_filt), ")\n")

# Export BED (centre ±100 bp, matching FIMO convention)
bg_centre <- (start(matched_bg) + end(matched_bg)) %/% 2L
df_bg <- data.frame(
  chr   = as.character(seqnames(matched_bg)),
  start = pmax(bg_centre - 100L - 1L, 0L),
  end   = bg_centre + 100L,
  name  = paste0("bg_", seq_along(matched_bg))
)
# matched_bg_200bp.bed in data/ is the draw the reported numbers rest on.
# Re-running this chunk gives a different draw under the same seed (the bin
# membership of the candidate enhancers differs slightly from the original
# session), so that file is not overwritten: a fresh draw is written alongside it.
bg_bed_path <- file.path(CTRL_FASTA_DIR, "matched_bg_200bp.bed")
if (file.exists(bg_bed_path)) {
  bg_bed_path <- file.path(CTRL_FASTA_DIR, "matched_bg_200bp.regenerated.bed")
}
write.table(df_bg, bg_bed_path, sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = FALSE)
cat("Matched-BG BED:", bg_bed_path, "\n")
```

---

```{bash ctrl-matched-bg-fasta-and-fimo, include=TRUE, eval=FALSE}
#!/usr/bin/env bash
# ---- FASTA + FIMO on the matched non-Tgo background ----
set -euo pipefail

PROJECT_DIR="${PROJECT_DIR:-$(pwd)}"
CTRL_FASTA_DIR="${PROJECT_DIR}/data/ctrl_tf/fimo/input_fastas"
CTRL_FIMO_ROOT="${PROJECT_DIR}/data/ctrl_tf/fimo"
DM6_FA="${PROJECT_DIR}/reference/dm6.fa"
MOTIF2_MEME="${PROJECT_DIR}/data/chapter3/STRCACTGARARAAA.meme"

bedtools getfasta \
  -fi "${DM6_FA}" \
  -bed "${CTRL_FASTA_DIR}/matched_bg_200bp.bed" \
  -fo  "${CTRL_FASTA_DIR}/matched_bg_200bp.fa" \
  -nameOnly
# -nameOnly (not -name): with -name FIMO parses the embedded locator and reports the chromosome

OUTDIR="${CTRL_FIMO_ROOT}/matched_bg_motif2"
mkdir -p "${OUTDIR}"
fimo --thresh 1.0E-4 --oc "${OUTDIR}" \
     "${MOTIF2_MEME}" \
     "${CTRL_FASTA_DIR}/matched_bg_200bp.fa"
echo "Matched background FIMO complete -> ${OUTDIR}/fimo.tsv"
```

---

```{r ctrl-fimo-load, include=TRUE, eval=TRUE}
# Read FIMO results, accepting p.value or p_value columns.
# Use p < 1e-4 as the significance criterion, not q < 0.05; p is comparable across set sizes, q is not.
# Same convention enforced in 04_chapter_analyses/chapter4/ch4_motif2_attribution.R.

read_fimo_hits <- function(fimo_tsv, p_thresh = 1e-4) {
  if (!file.exists(fimo_tsv)) {
    warning("FIMO output not found: ", fimo_tsv)
    return(character(0))
  }
  df <- read.table(fimo_tsv, sep = "\t", header = TRUE,
                   comment.char = "#", stringsAsFactors = FALSE)
  p_col <- if      ("p_value"  %in% names(df)) "p_value"
           else if ("p.value"  %in% names(df)) "p.value"
           else stop("FIMO output has neither p_value nor p.value column: ",
                     fimo_tsv)
  if (!"sequence_name" %in% names(df)) {
    stop("FIMO output missing sequence_name column: ", fimo_tsv)
  }
  # CHECK: sequence_name must identify a REGION, not a chromosome.
  # `bedtools getfasta -name` writes ">peak_1::chr3L:106074-106275"; FIMO parses
  # the embedded locator and puts the chromosome in sequence_name, so
  # unique(sequence_name) silently returns the number of chromosomes. That is
  # exactly how the published control panel came to report 6/6/5/5 positives.
  # Regenerate the FASTA with -nameOnly if this fires.
  if (all(grepl("^chr[0-9XYLRUHet]+$", unique(df$sequence_name)))) {
    stop("FIMO sequence_name values are chromosome names, not region ids: ",
         fimo_tsv, "\n  Regenerate the FASTA with `bedtools getfasta -nameOnly`.")
  }
  df <- df[!is.na(df[[p_col]]) & df[[p_col]] < p_thresh, ]
  unique(df$sequence_name)
}

fimo_twi <- file.path(CTRL_FIMO_ROOT, "TWI_motif2",        "fimo.tsv")
fimo_vvl <- file.path(CTRL_FIMO_ROOT, "VVL_motif2",        "fimo.tsv")
fimo_rib <- file.path(CTRL_FIMO_ROOT, "RIB_motif2",        "fimo.tsv")
fimo_da  <- file.path(CTRL_FIMO_ROOT, "DA_motif2",         "fimo.tsv")
fimo_bg  <- file.path(CTRL_FIMO_ROOT, "matched_bg_motif2", "fimo.tsv")

hits_twi <- read_fimo_hits(fimo_twi)
hits_vvl <- read_fimo_hits(fimo_vvl)
hits_rib <- read_fimo_hits(fimo_rib)
hits_da  <- read_fimo_hits(fimo_da)
hits_bg  <- read_fimo_hits(fimo_bg)

# Numerators (Motif2-positive peak count) and denominators
# (enhancer-overlapping peak count) for each set
n_twi_scan <- sum(TWI_peaks$overlaps_enhancer)
n_vvl_scan <- sum(VVL_peaks$overlaps_enhancer)
n_rib_scan <- sum(RIB_peaks$overlaps_enhancer)
n_da_scan  <- sum(DA_peaks$overlaps_enhancer)
n_bg_scan  <- length(matched_bg)

n_twi_m2 <- length(hits_twi)
n_vvl_m2 <- length(hits_vvl)
n_rib_m2 <- length(hits_rib)
n_da_m2  <- length(hits_da)
n_bg_m2  <- length(hits_bg)

# Motif2 control panel: Twi, Vvl, Rib and Da, with the Trh/Tgo reference and
# the matched background, every row a 201 bp window scanned at p < 1e-4. The
# table is written by 04_chapter_analyses/chapter4/ch4_ctrl_motif2_panel.R to
# data/chapter4/table4_ctrl_motif2_frequency.csv; this chunk reads it and
# checks that the counts above agree with it.
panel_csv <- file.path(CH4_DIR, "table4_ctrl_motif2_frequency.csv")
if (!file.exists(panel_csv)) {
  stop("Run 04_chapter_analyses/chapter4/ch4_ctrl_motif2_panel.R first; not found: ",
       panel_csv)
}
summary_tbl <- read.csv(panel_csv, check.names = FALSE, stringsAsFactors = FALSE)
stopifnot(nrow(summary_tbl) == 6L,
          identical(summary_tbl$Factor[2:5], c("TWI", "VVL", "RIB", "DA")))
n_pos <- setNames(summary_tbl$N_Motif2_pos, summary_tbl$Factor)
stopifnot(n_pos[["TWI"]] == n_twi_m2, n_pos[["VVL"]] == n_vvl_m2,
          n_pos[["RIB"]] == n_rib_m2, n_pos[["DA"]] == n_da_m2,
          n_pos[["Matched non-Tgo BG"]] == n_bg_m2)

cat("\n=== Motif2 frequency: controls vs heterodimer reference ===\n")
print(summary_tbl, row.names = FALSE)
```

---

