---
title: "CHAPTER 3"
author: "Yermek Aiten"
subtitle: "Genome-wide binding analysis of Trh/Tgo"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: show
    number_sections: true
toc_depth: 2
---

```{=html}
<style>
body { counter-reset: h1counter 2; }
h1 { counter-increment: h1counter; }
h1::before { content: counter(h1counter) ". "; }

table {
  border-collapse: collapse;
  width: 100%;
  font-size: 0.9em;
}
th {
  background-color: #2c3e50;
  color: white;
  padding: 8px 12px;
  text-align: left;
}
td {
  padding: 6px 12px;
  border-bottom: 1px solid #ddd;
}
tr:nth-child(even) { background-color: #f5f5f5; }
mark {
  background-color: #ffe066;
  padding: 1px 4px;
  border-radius: 3px;
}
</style>
```

```{r project-paths, include=FALSE}
## Repository root: the environment variable PROJECT_DIR, or the working directory.
PROJECT_DIR <- normalizePath(Sys.getenv("PROJECT_DIR", unset = "."), mustWork = TRUE)
if (!file.exists(file.path(PROJECT_DIR, "README.md")))
  stop("PROJECT_DIR does not point at the repository root: ", PROJECT_DIR)

## Values quoted in the dissertation are kept in results/reported_values.tsv
## (columns section, name, value). Writing a name that is already there
## replaces its value, so scripts can be re-run singly and in any order.
VALUES_FILE <- file.path(PROJECT_DIR, "results", "reported_values.tsv")
save_values <- function(section, values) {
  dir.create(dirname(VALUES_FILE), recursive = TRUE, showWarnings = FALSE)
  tab <- if (file.exists(VALUES_FILE)) {
    read.delim(VALUES_FILE, colClasses = "character", quote = "", comment.char = "")
  } else {
    data.frame(section = character(), name = character(), value = character())
  }
  for (nm in names(values)) {
    v   <- gsub("[\t\r\n]+", " ", paste(as.character(values[[nm]]), collapse = " "))
    tab <- tab[tab$name != nm, , drop = FALSE]
    tab <- rbind(tab, data.frame(section = section, name = nm, value = v))
  }
  tab <- tab[order(tab$section, tab$name), , drop = FALSE]
  write.table(tab, VALUES_FILE, sep = "\t", quote = FALSE, row.names = FALSE)
  cat(sprintf("%d values written to %s\n", length(values), VALUES_FILE))
}


## Colours
PAL_INK <- c(
  outline    = "#333333",   # bar and Venn outlines
  gridline   = "#DDDDDD",
  annotation = "#666666",   # p-values, n = counts
  threshold  = "#882255",   # significance thresholds, +/-1.96 lines
  fit        = "#4477AA",   # fitted / OLS trend lines
  neutral    = "#BBBBBB",   # non-significant, background sets
  band       = "#EDEDED",   # shaded stage/facet bands and strip backgrounds
  highlight_band = "#4477AA", # tint marking an agreement region (used at low alpha)
  emphasis   = "#CC3311",   # the one highlighted subset in an otherwise neutral plot
## Ink for stars/text on PAL_DIV marks (Fig 4.9): black #000000, worst-case WCAG contrast 4.34 across 21-point ramp sample.
## Not valid for PAL_SEQ (viridis): no single ink stays legible, so labels sit off the dot instead.
## Exception: geom_tile heatmaps use ink_on_seq() (below), keyed on tile lightness, worst-case 4.16 across the ramp.
  on_mark    = "#000000",

  ## Two-tier ink for ink_on_seq(). `on_mark_dark` is #1A1A1A rather
  ## than the #000000 above: on the LIGHT half of viridis both clear the floor
  ## comfortably, and #1A1A1A matches the ink weight of the surrounding axis text,
  ## so the matrix does not look heavier than the panel around it.
  on_mark_dark  = "#1A1A1A",
  on_mark_light = "#FFFFFF"
)

PAL_HETERO <- c(
  "Trh/Tgo"  = "#117733",   # green    L* 44
  "Sim/Tgo"  = "#E69F00",   # amber    L* 71
  "Dys/Tgo"  = "#66CCEE",   # sky      L* 77
  "Sima/Tgo" = "#AA3377"    # magenta  L* 42
)

PAL_SUBUNIT <- c(
  "Trh"  = PAL_HETERO[["Trh/Tgo"]],
  "Sim"  = PAL_HETERO[["Sim/Tgo"]],
  "Dys"  = PAL_HETERO[["Dys/Tgo"]],
  "Sima" = PAL_HETERO[["Sima/Tgo"]],
  "Tgo"  = "#DDCC77"
)

pal_venn_tgo <- function(alpha_subunit) {
  if (!alpha_subunit %in% names(PAL_SUBUNIT)) {
    stop("pal_venn_tgo(): expected an alpha-subunit name (",
         paste(setdiff(names(PAL_SUBUNIT), "Tgo"), collapse = ", "),
         "), got ", deparse(alpha_subunit),
         ". For Venn transparency use PAL_VENN_ALPHA.", call. = FALSE)
  }
  unname(c(PAL_SUBUNIT[[alpha_subunit]], PAL_SUBUNIT[["Tgo"]]))
}

PAL_CONTEXT <- c(
  prox_peak    = "#4477AA", # promoter-proximal peak (non-annotated)   blue
  prox_enh     = "#332288", # promoter-proximal enhancer (annotated)   indigo
  distal_enh   = "#DDAA33", # distal enhancer (annotated)  <- KEY CLASS  gold
  distal_unann = "#994455", # distal peak (non-annotated)              maroon
  all_peaks    = "#BBBBBB", # pooled / reference set                   grey
  ## POOLED positional classes, supersets of the four subsets above, used by
  ## figures that split peaks by POSITION only (Ch3 df$Context, Ch4 Subset).
  ## They keep the family reading (cool = proximal, warm = distal) but are
  ## distinct from every subset colour, so a reader cannot carry a
  ## subset-specific claim across to a pooled figure. Chosen by measurement over
  ## the three figures that use them: worst-case dE 34.6.
  prox_all     = "#0077BB", # all promoter-proximal peaks (pooled)      deep blue
  distal_all   = "#EE7733"  # all distal peaks (pooled)                 orange
)

pal_context <- function(keys) {
  stopifnot(all(keys %in% names(PAL_CONTEXT)))
  unname(PAL_CONTEXT[keys])
}

PAL_CONTEXT_LABELS <- c(
  ## --- (A) pooled positional ---
  "Promoter"                               = PAL_CONTEXT[["prox_all"]],
  "promoter-proximal"                          = PAL_CONTEXT[["prox_all"]],
  "distal (>1 kb)"                         = PAL_CONTEXT[["distal_all"]],
  "all peaks"                              = PAL_CONTEXT[["all_peaks"]],
  ## --- (B) 2x2 position-by-annotation subsets ---
  "Promoter-proximal peak (non-annotated)" = PAL_CONTEXT[["prox_peak"]],
  "Promoter-proximal enhancer (annotated)" = PAL_CONTEXT[["prox_enh"]],
  "Distal enhancer (annotated)"            = PAL_CONTEXT[["distal_enh"]],
  "Distal peak (non-annotated)"            = PAL_CONTEXT[["distal_unann"]]
)

PAL_FEATURE <- c(
  "Promoter (<=1kb)"  = "#332288",  # dark indigo - nearest the TSS (darkest blue)
  "Promoter (1-2kb)"  = "#66CCEE",  # light sky   - still proximal, further out
  "5' UTR"            = "#117733",  # green       - transcript begins
  "Exon"              = "#E69F00",  # amber       - coding body (1st + other)
  "3' UTR"            = "#997700",  # dark gold   - transcript ends
  "1st Intron"        = "#DDCC77",  # sand
  "Other Intron"      = "#EE6677",  # rose
  "Downstream"        = "#CC79A7",  # pink        - past the gene
  "Distal Intergenic" = "#AA3377"   # magenta     - furthest from any gene
)

pal_feature <- function(x) {
  levs <- if (inherits(x, "ggplot")) {
            f <- x$data$Feature
            if (is.factor(f)) levels(f) else unique(as.character(f))
          } else if (is.factor(x)) levels(x) else unique(as.character(x))

  ## Label -> palette key. Which distinctions survive is a deliberate choice:
  ## promoter distance bins are preserved (the text argues about distance from
  ## the TSS), exon classes are merged, and downstream distance bins are merged
  ## (distance past the 3' end carries no claim). Anything not handled here must
  ## error, never be quietly folded into a neighbour.
  key_of <- function(lab) {
    lab <- trimws(lab)
    ## Promoter: keep the bin, but normalise the many spellings ChIPseeker and
    ## its versions emit ("<=1kb", "<1kb", "1-2kb", "2-3kb").
    if (grepl("^Promoter", lab)) {
      q <- sub("^Promoter\\s*", "", lab)
      q <- gsub("[()<= ]", "", q)                    # "(<=1kb)" -> "1kb"
      if (q %in% c("", "1kb", "0-1kb"))  return("Promoter (<=1kb)")
      if (q %in% c("1-2kb", "2kb"))      return("Promoter (1-2kb)")
      return(lab)                                    # unknown bin -> error below
    }
    ## Everything else: drop the distance qualifier, then merge the classes the
    ## palette treats as one.
    lab <- trimws(sub("\\s*\\(.*", "", lab))
    lab <- sub("^(1st|Other) Exon$", "Exon", lab)     # exon classes combined
    lab <- sub("^Exon$",   "Exon",   lab)
    lab <- sub("^Intron$", "Other Intron", lab)
    lab <- sub("^Intergenic$", "Distal Intergenic", lab)
    lab <- sub("^fiveUTR$",  "5' UTR", lab)
    lab <- sub("^threeUTR$", "3' UTR", lab)
    lab
  }
  keys <- vapply(levs, key_of, character(1))
  bad  <- levs[!keys %in% names(PAL_FEATURE)]
  if (length(bad)) {
    stop("pal_feature(): unrecognised annotation level(s): ",
         paste(sQuote(bad), collapse = ", "),
         "\n  Add them to PAL_FEATURE rather than letting ",
         "ggplot fall back to its default scale.", call. = FALSE)
  }
  setNames(unname(PAL_FEATURE[keys]), levs)
}

PAL_SEQ <- function(n = 9) hcl.colors(n, "viridis")

seq_limits <- function(x, domain = NULL, min_span = NULL) {
  x <- x[is.finite(x)]
  if (!length(x)) return(if (is.null(domain)) c(0, 1) else range(domain))
  lo <- min(x); hi <- max(x)
  if (hi > lo) return(c(lo, hi))            # normal case: data-derived, untouched
  if (!is.null(domain)) return(range(domain))   # degenerate: use natural domain
  if (is.null(min_span)) min_span <- if (hi == 0) 1 else abs(hi)
  sort(c(hi, hi + min_span))                # degenerate, no domain: expand upward
}

.seq_args <- function(dots, values, domain) {
  if (is.null(values)) return(dots)
  if (!is.null(dots$limits))
    stop("scale_*_seq(): pass either `values` (guarded limits) or `limits`, not both.",
         call. = FALSE)
  dots$limits <- seq_limits(values, domain = domain)
  if (is.null(dots$oob)) dots$oob <- scales::squish
  dots
}

scale_fill_seq   <- function(..., rev = FALSE, values = NULL, domain = NULL) {
  cols <- PAL_SEQ(9); if (rev) cols <- base::rev(cols)
  args <- .seq_args(list(...), values, domain)
  do.call(ggplot2::scale_fill_gradientn, c(list(colours = cols), args))
}

geom_point_seq <- function(..., stroke_colour = PAL_INK[["outline"]],
                           stroke_width = 0.35) {
  ggplot2::geom_point(..., shape = 21, stroke = stroke_width,
                      colour = stroke_colour)
}

PAL_BINARY <- c(pos = "#CC6677", neg = "#4477AA")

pal_binary <- function(pos_label, neg_label) {
  setNames(c(PAL_BINARY[["pos"]], PAL_BINARY[["neg"]]), c(pos_label, neg_label))
}

PAL_OCCUPANCY <- c(
  alpha_only = "#66CCEE",                  # Trh-unique  (no Tgo): unpartnered = cool
  joint      = PAL_HETERO[["Trh/Tgo"]],    # Trh/Tgo joint peaks, the reference set
  beta_only  = PAL_SUBUNIT[["Tgo"]],       # Tgo-unique  (no Trh): the Tgo sand
  genome_ref = "#888888"                   # genome-wide enhancers: a darker baseline
)

PAL_TISSUE <- c(
  "Tracheal+SG" = "#94A680",   # tracheal system + salivary gland  sage-green
  "CNS+midline" = "#A8DCD1"    # CNS + ventral midline             pale teal
)

PAL_VENN_ALPHA <- 0.8

pal_venn_tissue <- function(heterodimer, tissue) {
  stopifnot(heterodimer %in% names(PAL_HETERO))
  i <- match(tolower(tissue), tolower(names(PAL_TISSUE)))
  if (is.na(i)) stop("unknown tissue reference list: ", tissue,
                     " (expected one of ", paste(names(PAL_TISSUE), collapse = ", "), ")")
  unname(c(PAL_HETERO[[heterodimer]], PAL_TISSUE[[i]]))
}

pal_disttotss <- function(x) {
  levs <- if (inherits(x, "ggplot")) {
            f <- x$data$Feature
            if (is.factor(f)) levels(f) else unique(as.character(f))
          } else if (is.factor(x)) levels(x) else unique(as.character(x))
  n <- length(levs)
  if (!n) stop("pal_disttotss(): no distance-bin levels found", call. = FALSE)
  ## PAL_SEQ runs dark -> light, so the first level (nearest TSS) is darkest.
  ## plotDistToTSS reverses its own level order before plotting, so the caller
  ## must apply this to the levels as they appear on the built plot.
  setNames(PAL_SEQ(n), levs)
}


## Bar chart
thesis_text_col <- function(fill) {
  rgb <- grDevices::col2rgb(fill) / 255
  lum <- 0.299 * rgb[1, ] + 0.587 * rgb[2, ] + 0.114 * rgb[3, ]
  ifelse(lum > 0.5, "black", "white")
}

thesis_p_expr <- function(p, digits = 2) {
  if (is.na(p)) return(NA_character_)
  ## The formatted value is QUOTED, so plotmath prints it as formatted rather
  ## than re-formatting it as a number: unquoted, a Fisher p of 1.000 renders
  ## as "1" and 2.10 as "2.1", losing the fixed significant figures that
  ## must be preserved at every mention.
  if (p < 1e-3) {
    parts <- strsplit(formatC(p, format = "e", digits = digits), "e")[[1]]
    sprintf("italic(p)=='%s' %%*%% 10^{%d}", parts[1], as.integer(parts[2]))
  } else {
    ## `digits` governs both branches. "fg" formats to significant digits
    ## rather than decimal places; flag "#" keeps trailing zeros (p = 1
    ## prints 1.0, not 1).
    sprintf("italic(p)=='%s'",
            formatC(signif(p, digits), format = "fg", digits = digits,
                    flag = "#"))
  }
}

thesis_bar <- function(data, x, y, fill = x,
                        value = NULL, value_fmt = "%.1f%%",
                        count = NULL, sig = NULL,
                        facet = NULL, n_lab = NULL, n_lab_size = 3.0,
                        fills = NULL, y_max = NULL, y_lab = NULL,
                        dashed_at = NULL, global_p = NULL, caption_sig = NULL,
                        inside_frac = 0.28, bar_width = 0.62, base_size = 11,
                        title = NULL, facet_free_x = TRUE) {

  d <- as.data.frame(data)
  d$.x <- d[[x]]; d$.y <- d[[y]]; d$.fill <- d[[fill]]

  ## primary value label
  if (!is.null(value)) {
    d$.value <- as.character(d[[value]])
  } else if (!is.na(value_fmt)) {
    d$.value <- sprintf(value_fmt, d$.y)
  } else {
    d$.value <- NA_character_
  }
  d$.count <- if (!is.null(count)) as.character(d[[count]]) else NA_character_
  d$.sig   <- if (!is.null(sig))   as.character(d[[sig]])   else NA_character_

  ## y range + headroom (space for the star that always sits above the bar)
  y_top_data <- max(d$.y, na.rm = TRUE)
  if (is.null(y_max)) y_max <- y_top_data * 1.32
  gap  <- 0.045 * y_max            # one label-line height, scaled to axis
  inside_cut <- inside_frac * y_max

  ## fill -> contrast colour for inside text
  if (!is.null(fills)) {
    missing_fill <- setdiff(unique(as.character(d$.fill)), names(fills))
    if (length(missing_fill))
      stop("thesis_bar(): fill level(s) not found in `fills` (would render grey): ",
           paste(sprintf("'%s'", missing_fill), collapse = ", "),
           "\n  Available in `fills`: ",
           paste(sprintf("'%s'", names(fills)), collapse = ", "))
  }
  fill_levels <- if (!is.null(fills)) names(fills) else unique(as.character(d$.fill))
  fill_cols   <- if (!is.null(fills)) fills else scales::hue_pal()(length(fill_levels))
  names(fill_cols) <- fill_levels
  d$.txtcol <- thesis_text_col(fill_cols[as.character(d$.fill)])

  ## ---- per-bar label geometry ------------------------------------------
  d$.inside <- d$.y >= inside_cut
  has_count <- !is.na(d$.count)
  # value line: inside -> just below bar top; above -> primary value uppermost
  d$.val_y   <- ifelse(d$.inside, d$.y - gap,
                       ifelse(has_count, d$.y + 2 * gap, d$.y + gap))
  d$.val_col <- ifelse(d$.inside, d$.txtcol, "black")
  # count line: inside -> below the value; above -> nearer the bar than value
  d$.cnt_y   <- ifelse(d$.inside, d$.y - 2 * gap, d$.y + gap)
  d$.cnt_col <- ifelse(d$.inside, d$.txtcol, "black")
  # significance always above the bar; clears above-bar labels when present
  above_labels <- (!d$.inside) & (has_count | !is.na(d$.value))
  d$.sig_y    <- d$.y + ifelse(above_labels, 3 * gap, gap * 1.4)

  ## ---- build plot -------------------------------------------------------
  p <- ggplot(d, aes(x = .x, y = .y, fill = .fill)) +
    geom_col(width = bar_width, colour = "black", linewidth = 0.4,
             show.legend = FALSE)

  if (!is.null(dashed_at))
    p <- p + geom_hline(yintercept = dashed_at, linetype = "dashed",
                        colour = PAL_INK[["neutral"]], linewidth = 0.5)

  ## value label
  if (any(!is.na(d$.value)))
    p <- p + geom_text(aes(y = .val_y, label = .value, colour = .val_col),
                       fontface = "bold", size = 3.5, show.legend = FALSE,
                       na.rm = TRUE)
  ## count label
  if (any(!is.na(d$.count)))
    p <- p + geom_text(aes(y = .cnt_y, label = .count, colour = .cnt_col),
                       size = 3.0, show.legend = FALSE, na.rm = TRUE)
  ## per-bar significance (above, spaced)
  if (any(!is.na(d$.sig)))
    p <- p + geom_text(aes(y = .sig_y, label = .sig),
                       fontface = "italic", size = 3.4, colour = PAL_INK[["annotation"]],
                       na.rm = TRUE)

  p <- p +
    scale_colour_identity() +
    scale_fill_manual(values = fill_cols) +
    scale_y_continuous(limits = c(0, y_max), expand = expansion(mult = c(0, 0)))

  ## facets, free_x so each panel shows only its own x categories
  if (!is.null(facet))
    p <- p + facet_wrap(stats::as.formula(paste0("~", facet)), nrow = 1,
                        scales = if (facet_free_x) "free_x" else "fixed")

  ## per-facet n = ... annotation, top-right
  if (!is.null(n_lab)) {
    key <- if (!is.null(facet)) facet else NULL
    nd  <- if (!is.null(key)) d[!duplicated(d[[key]]), ] else d[1, , drop = FALSE]
    nd$.nlab <- as.character(nd[[n_lab]])
    parse_n  <- any(grepl("==|~", nd$.nlab))
    p <- p + geom_text(data = nd,
                       aes(x = Inf, y = Inf, label = .nlab),
                       hjust = 1.08, vjust = 1.6, size = n_lab_size, colour = PAL_INK[["annotation"]],
                       fontface = "italic", parse = parse_n,
                       inherit.aes = FALSE)
  }

  ## single centred p-value, drawn once (as a plot tag), not per-facet,
  ## so a faceted figure shows one central p-value instead of one per panel.
  plot_tag <- NULL
  if (!is.null(global_p)) plot_tag <- parse(text = thesis_p_expr(global_p))[[1]]

  cap <- caption_sig
  p <- p +
    labs(x = NULL, y = y_lab, title = title, caption = cap, tag = plot_tag) +
    theme_classic(base_size = base_size) +
    theme(
      ## BOLD: the panel letter is the one bold element of the house style.
      plot.tag          = element_text(size = base_size + 2, face = "bold"),
      plot.tag.position  = c(0.5, 0.86),
      strip.text       = element_text(face = "bold", size = base_size),
      strip.background = element_rect(fill = NA, colour = "black", linewidth = 0.6),
      axis.title.y     = element_text(size = base_size - 0.5),
      axis.text.x      = element_text(size = base_size - 2, colour = "black"),
      axis.text.y      = element_text(size = base_size - 2, colour = "black"),
      axis.ticks.x     = element_blank(),
      panel.spacing    = unit(1.0, "cm"),
      plot.caption     = element_text(hjust = 0, size = base_size - 2.5,
                                      colour = PAL_INK[["outline"]]),
      plot.title       = element_text(size = base_size, face = "bold"),
      plot.margin      = margin(8, 10, 8, 8)
    )
  p
}


## UpSet plot
upset_class_cols <- function()
  c("Promoter-containing" = PAL_CONTEXT[["prox_all"]],
    "Genic, non-promoter" = PAL_FEATURE[["1st Intron"]],
    "Intergenic"          = PAL_CONTEXT[["distal_all"]])

upset_bars <- function(a, n_int = 20, base_size = 13) {
  y  <- as.matrix(a@detailGenomicAnnotation); nn <- colnames(y)
  sets <- lapply(seq_len(nrow(y)), function(i) nn[y[i, ]])
  res  <- tibble::tibble(anno = sets)
  res$class <- factor(vapply(sets, function(s) {
    if ("Promoter" %in% s) "Promoter-containing"
    else if (any(c("Intergenic", "distal_intergenic") %in% s)) "Intergenic"
    else "Genic, non-promoter" }, character(1)),
    levels = c("Promoter-containing", "Genic, non-promoter", "Intergenic"))
  ## DISPLAY LABELS. ChIPseeker's detailGenomicAnnotation columns are code
  ## identifiers (fiveUTR, threeUTR, distal_intergenic, lowercase genic). The
  ## text, tables and PAL_FEATURE all use "5' UTR", "3' UTR",
  ## "Distal intergenic". Relabel for DISPLAY only, set membership and every
  ## count come from the untouched column names above, so nothing quantitative
  ## depends on this map.
  feat_labels <- c(genic             = "Genic",
                   Promoter          = "Promoter",
                   Intron            = "Intron",
                   Exon              = "Exon",
                   fiveUTR           = "5\u2032 UTR",
                   threeUTR          = "3\u2032 UTR",
                   Intergenic        = "Intergenic",
                   distal_intergenic = "Distal intergenic",
                   downstream        = "Downstream")
  miss <- setdiff(nn, names(feat_labels))
  if (length(miss))
    stop("upset_bars(): no display label for ChIPseeker column(s) ",
         paste(sQuote(miss), collapse = ", "),
         ". Add them to feat_labels rather than letting the raw name print.",
         call. = FALSE)
  ## scale_x_upset() takes `sets` but has no `labels` argument (checked against
  ## ggupset's formals), so the relabel is applied to the set members in the
  ## data. res$class was already computed from the RAW names above, so the
  ## grouping is unaffected by this rename.
  res$anno <- lapply(res$anno, function(s) unname(feat_labels[s]))

  ggplot2::ggplot(res, ggplot2::aes(x = anno, fill = class)) +
    ggplot2::geom_bar(width = 0.72) +
    ggplot2::scale_fill_manual(values = upset_class_cols(), name = NULL) +
    ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.06))) +
    ggupset::scale_x_upset(n_intersections = n_int, order_by = "freq") +
    ggplot2::labs(x = NULL, y = "Peak count") +
    ggplot2::theme_minimal(base_size = base_size) +
    ggplot2::theme(legend.position = "none",
          panel.grid.major.x = ggplot2::element_blank(),
          ## FULL BLACK TEXT. theme_minimal() (like theme_grey()) sets axis.text
          ## to #4D4D4D, a 30% grey. Measured in the rendered png, the y tick
          ## numbers and the genomic-feature names to the left of the dot matrix
          ## came out #3C3C3C-#4B4B4B. Those elements print at ~8.6 pt, so they
          ## need maximum contrast rather than 30% grey. Set explicitly on each
          ## element: axis.text.x/.y do not inherit a colour set only on
          ## axis.text once the child elements are themselves specified.
          axis.text    = ggplot2::element_text(colour = "black"),
          axis.text.x  = ggplot2::element_text(colour = "black"),
          axis.text.y  = ggplot2::element_text(colour = "black"),
          axis.title   = ggplot2::element_text(colour = "black"),
          axis.title.y = ggplot2::element_text(size = 13.5, colour = "black",
                                               margin = ggplot2::margin(r = 5)),
          plot.margin = ggplot2::margin(t = 4, r = 2, b = 2, l = 2, unit = "pt"))
}

thesis_upset <- function(a, label, n_int = 20, base_size = 13) {
  upset_bars(a, n_int = n_int, base_size = base_size) +
    ggplot2::theme(
      legend.position      = c(0.74, 0.88),
      legend.justification = c(0, 1),
      legend.background    = ggplot2::element_rect(
                               fill = grDevices::adjustcolor("white", alpha.f = 0.85),
                               colour = NA),
      legend.key.size      = grid::unit(13, "pt"),
      legend.text          = ggplot2::element_text(size = 13, colour = "black"),
      plot.title           = ggplot2::element_text(face = "bold", size = 16,
                                                   colour = "black", hjust = 0)) +
    ggplot2::labs(title = label)
}
```

# Set-up and data import
```{r setup, include=TRUE, eval=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
# Output locations
REFERENCE_LISTS <- file.path(PROJECT_DIR,"data/reference_gene_lists")
FIGURES_DIR <- file.path(PROJECT_DIR, "results/figures")
#dir.create(FIGURES_DIR, recursive = TRUE, showWarnings = FALSE)
EXPORT_DIR <- file.path(PROJECT_DIR, "data/chapter3")
dir.create(EXPORT_DIR, recursive = TRUE, showWarnings = FALSE)


library(ChIPseeker)
library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
library(BSgenome.Dmelanogaster.UCSC.dm6)
library(clusterProfiler)
library(AnnotationDbi)
library(org.Dm.eg.db)
library(Biostrings)
library(seqinr)
library(rtracklayer)
library(ChIPpeakAnno)
library(ggplot2)
library(ggplotify)
library(DiffBind)
library(gridExtra)
library(motifStack)
library(reactome.db)
library(DBI)
library(GenomicRanges)
library(GenomicFeatures)
library(cowplot)
library(GenomeInfoDb)
#Loading data
tgo_narrowPeak_file <- file.path(PROJECT_DIR, "data/final_peaks/TGO_IDR0.05.final.narrowPeak")
trh_narrowPeak_file <- file.path(PROJECT_DIR, "data/final_peaks/TRH_IDR0.05.final.narrowPeak")

# Focusing on standard assemblies of chromosomes & Deduplication
extraCols_narrowPeak <- c(signalValue = "numeric", pValue = "numeric",
                       qValue = "numeric", peak = "integer")
Trh_ChIP_peaks <- import(trh_narrowPeak_file, format = "BED", extraCols = extraCols_narrowPeak
                       ); Trh_ChIP_peaks
Tgo_ChIP_peaks <- import(tgo_narrowPeak_file, format = "BED", extraCols = extraCols_narrowPeak
                       ); Tgo_ChIP_peaks #
table(seqnames(Trh_ChIP_peaks)); table(seqnames(Tgo_ChIP_peaks))
txdb <- TxDb.Dmelanogaster.UCSC.dm6.ensGene
files <- list(Trh_ChIP_peaks, Tgo_ChIP_peaks)
names(files)<- c("Trh ChIP-seq", "Tgo ChIP-seq")
# Remove non-standard chromosome/scaffold entries (e.g. 'chrX_DS484057v1_random')
seqlevels(files[[1]]); seqlevels(files[[2]])
good_chr <- c("chr3L","chr3R","chrX","chr2L","chr2R","chr4")
for(i in 1:length(files)){
  gr <- files[[i]]
files[[i]] <- gr[seqnames(gr) %in% good_chr]
files[[i]] <- keepSeqlevels(files[[i]], 
                            good_chr,
                            pruning.mode = "coarse"); unique(seqnames(files[[i]]))
}
# Deduplication
Trh_clean <- files[[1]][!duplicated(files[[1]])]
Tgo_clean <- files[[2]][!duplicated(files[[2]])]
files <- list(Trh_clean, Tgo_clean)
names(files)<- c("Trh ChIP-seq", "Tgo ChIP-seq")
seqlevels(files[[1]]); seqlevels(files[[2]])
length(files[[1]]); length(files[[2]])
```

# Global binding architecture

## Per-chromosome distribution

```{r, include=TRUE, eval=FALSE, 'Per-chromosome distributions Trh vs Tgo binding (percent per chromosome)'}
# Table 3.1. Per chromosome distribution of Trh and Tgo ChIP-seq peaks (genome version: dm6)

# Ensure seqlevels are identical and ordered
common_seqs <- intersect(seqlevels(Trh_clean), seqlevels(Tgo_clean))
Trh <- keepSeqlevels(Trh_clean, common_seqs, pruning.mode = "coarse")
Tgo <- keepSeqlevels(Tgo_clean, common_seqs, pruning.mode = "coarse")
trh_chr_counts <- table(seqnames(Trh))
tgo_chr_counts <- table(seqnames(Tgo))
trh_chr_percent <- 100 * trh_chr_counts / sum(trh_chr_counts)
tgo_chr_percent <- 100 * tgo_chr_counts / sum(tgo_chr_counts)
chr_distribution <- data.frame(
  chromosome = common_seqs,
  Trh_percent = as.numeric(trh_chr_percent[common_seqs]),
  Tgo_percent = as.numeric(tgo_chr_percent[common_seqs])
); chr_distribution
library(tidyr)
chr_distribution_long <- pivot_longer(
  chr_distribution,
  cols = c(Trh_percent, Tgo_percent),
  names_to = "factor",
  values_to = "percent"
); chr_distribution_long
chr_distribution$Trh_Tgo_ratio <-
  chr_distribution$Trh_percent / chr_distribution$Tgo_percent
```

```{r, include=TRUE, eval=FALSE, 'Figure 3.2 Trh and Tgo ChIP peaks over chromosomes'}
# Figure 3.2 Trh/Tgo ChIP peaks by chromosome. covplot() facets on chromosome,
# free y scales, n.breaks = 4; at 1500px/300dpi (6 facets, ~0.8in each) this keeps
# labels legible, e.g. chr4 gets 0/10/20/30. Scale must be added before as.ggplot() (wraps in annotation_custom); covplot sets fill only, no colour scale.
cov_y_scale <- scale_y_continuous(n.breaks = 4,
                                  expand = expansion(mult = c(0, 0.08)))

png(file = file.path(FIGURES_DIR, "fig3_2_A.png"),
    width = 2000, height = 1200, res = 300)
covp1 <- ggplotify::as.ggplot(
  covplot(files[[1]], weightCol = "signalValue",
          title = "Trh-ChIP peaks over chromosomes",
          fill_color = PAL_SUBUNIT[["Trh"]]) + cov_y_scale
)
covp1
while(!is.null(dev.list()))
dev.off()

png(file = file.path(FIGURES_DIR, "fig3_2_B.png"),
    width = 2000, height = 1200, res = 300)
covp2 <- ggplotify::as.ggplot(
  covplot(files[[2]], weightCol = "signalValue",
          title = "Tgo-ChIP peaks over chromosomes",
          fill_color = PAL_SUBUNIT[["Tgo"]]) + cov_y_scale
)
covp2
while(!is.null(dev.list()))
dev.off()
```

## Binding around transcription start sites (TSS)

```{bash, include=TRUE, eval=FALSE, "Figure 3.2 will be made using DeepTools and bamCompare in bash"}
# Figure 3.2
#!/usr/bin/env bash
# ============================================================
# VISUALISATION PIPELINE, Trh, Tgo ChIP-seq
# TSS enrichment profiles and heatmaps
# Genome: dm6
# Run in the ChIP-seq processing directory: it reads the blacklist-filtered
# BAM files in 07_blacklist/ (not in the repository; see 01_chipseq_processing/).
# ============================================================
set -euo pipefail
mkdir -p 05_visualisation/{bigwigs,matrices,plots}
mkdir -p gene_references
# ============================================================
# STAGE 1, INDEX BAMs
# ============================================================
# Files downloaded from ENCODE (accessions SRR28099027, SRR28099028, SRR28099029,
# SRR28102500, SRR28099453, SRR28099454, SRR28100325) are renamed to sample labels.
for BAM in \
07_blacklist/TGO_rep1.clean.bam \
07_blacklist/TGO_rep2.clean.bam \
07_blacklist/TGO_rep3.clean.bam \
07_blacklist/TGO_control.clean.bam \
07_blacklist/TRH_rep1.clean.bam \
07_blacklist/TRH_rep2.clean.bam \
07_blacklist/TRH_control.clean.bam
do
[ ! -f "${BAM}.bai" ] && samtools index "$BAM"
done
# ============================================================
# STAGE 2, BIGWIG GENERATION
# ============================================================
# --- TGO (paired-end) ---
for REP in 1 2 3
do
bamCompare \
-b1 07_blacklist/TGO_rep${REP}.clean.bam \
-b2 07_blacklist/TGO_control.clean.bam \
-o 05_visualisation/bigwigs/TGO_rep${REP}_over_input.bw \
--binSize 20 \
--normalizeUsing BPM \
--scaleFactorsMethod None \
--smoothLength 60 \
--centerReads \
--skipNAs \
-p 8
done
# --- TRH (single-end) ---
for REP in 1 2
do
bamCompare \
-b1 07_blacklist/TRH_rep${REP}.clean.bam \
-b2 07_blacklist/TRH_control.clean.bam \
-o 05_visualisation/bigwigs/TRH_rep${REP}_over_input.bw \
--binSize 20 \
--normalizeUsing BPM \
--scaleFactorsMethod None \
--smoothLength 60 \
--extendReads 210 \
--skipNAs \
-p 8
done
# ============================================================
# STAGE 3, GENE FILTERING
# ============================================================
echo -e "chr2L\nchr2R\nchr3L\nchr3R\nchrX\nchr4" \
> gene_references/confirmed_chromosomes.txt

awk 'NR==FNR {c[$1]; next} $1 in c' \
gene_references/confirmed_chromosomes.txt \
gene_references/dm6.ensGene.gtf \
> gene_references/filtered.dm6.ensGene.gtf

# ============================================================
# STAGE 4, COMPUTE MATRICES
# ============================================================
computeMatrix reference-point \
--referencePoint TSS \
-b 3000 -a 3000 \
-R gene_references/filtered.dm6.ensGene.gtf \
-S 05_visualisation/bigwigs/TGO_rep1_over_input.bw \
05_visualisation/bigwigs/TGO_rep2_over_input.bw \
05_visualisation/bigwigs/TGO_rep3_over_input.bw \
05_visualisation/bigwigs/TRH_rep1_over_input.bw \
05_visualisation/bigwigs/TRH_rep2_over_input.bw \
--missingDataAsZero \
--skipZeros \
-o 05_visualisation/matrices/matrix_TSS_all_TFs_3kb.gz \
--outFileSortedRegions 05_visualisation/matrices/regions_TSS_all_TFs_3kb.bed \
-p 8

# --- ZOOM MATRIX ---
computeMatrix reference-point \
--referencePoint TSS \
-b 500 -a 500 \
-R gene_references/filtered.dm6.ensGene.gtf \
-S 05_visualisation/bigwigs/TGO_rep1_over_input.bw \
05_visualisation/bigwigs/TGO_rep2_over_input.bw \
05_visualisation/bigwigs/TGO_rep3_over_input.bw \
05_visualisation/bigwigs/TRH_rep1_over_input.bw \
05_visualisation/bigwigs/TRH_rep2_over_input.bw \
--missingDataAsZero \
--skipZeros \
-o 05_visualisation/matrices/matrix_TSS_all_TFs_500bp.gz \
--outFileSortedRegions 05_visualisation/matrices/regions_TSS_all_TFs_500bp.bed \
-p 8

# ============================================================
# Colours mirror PAL_SUBUNIT via colorspace lighten/darken (deepTools cannot source R); use matrix_TSS_all_TFs_{3kb,500bp}.gz (TGOx3+TRHx2) with --colors to match TF colouring elsewhere.
# ============================================================

# ============================================================
# STAGE 5, PROFILE PLOTS (A, B)
# ============================================================

plotProfile \
-m 05_visualisation/matrices/matrix_TSS_all_TFs_3kb.gz \
-o $PROJECT_DIR/results/figures/fig3_3_A.png \
--perGroup \
--samplesLabel \
"Tgo rep1" "Tgo rep2" "Tgo rep3" \
"Trh rep1" "Trh rep2" \
--colors '#DDCC77' '#CCBB64' '#BBAA50' '#117733' '#368C4C' \
--dpi 300

plotProfile \
-m 05_visualisation/matrices/matrix_TSS_all_TFs_500bp.gz \
-o $PROJECT_DIR/results/figures/fig3_3_B.png \
--perGroup \
--samplesLabel \
"Tgo rep1" "Tgo rep2" "Tgo rep3" \
"Trh rep1" "Trh rep2" \
--colors '#DDCC77' '#CCBB64' '#BBAA50' '#117733' '#368C4C' \
--dpi 300

# ============================================================
# STAGE 6, HEATMAP (C)
# ============================================================

plotHeatmap \
-m 05_visualisation/matrices/matrix_TSS_all_TFs_3kb.gz \
-o $PROJECT_DIR/results/figures/fig3_3_C.png \
--samplesLabel \
"Tgo rep1" "Tgo rep2" "Tgo rep3" \
"Trh rep1" "Trh rep2" \
--whatToShow "heatmap and colorbar" \
--colorMap RdBu \
--zMin -1 --zMax 1 \
--sortRegions descend \
--sortUsingSamples 1 \
--heatmapHeight 12 \
--heatmapWidth 3 \
--dpi 300 \
--xAxisLabel "" \
--refPointLabel "TSS"

# ============================================================
# OPTIONAL, PEAK-CENTRED ANALYSIS (RECOMMENDED FOR PAPER)
# ============================================================

computeMatrix reference-point \
--referencePoint center \
-b 2000 -a 2000 \
-R data/final_peaks/TGO_IDR0.05.final.narrowPeak \
data/final_peaks/TRH_IDR0.05.final.narrowPeak \
-S 05_visualisation/bigwigs/TGO_rep1_over_input.bw \
05_visualisation/bigwigs/TGO_rep2_over_input.bw \
05_visualisation/bigwigs/TGO_rep3_over_input.bw \
05_visualisation/bigwigs/TRH_rep1_over_input.bw \
05_visualisation/bigwigs/TRH_rep2_over_input.bw \
-o 05_visualisation/matrices/matrix_peaks.gz \
--outFileSortedRegions 05_visualisation/matrices/regions_peaks_centred.bed \
-p 8

# ============================================================
echo "=== DONE ==="
```

# Genomic feature annotation

## Distribution across genomic features

```{r, include=TRUE, eval=FALSE, 'Figure 3.4. Distribution of Trh and Tgo TF binding sites relative to the transcription start site (TSS)'}
#ChIP peak annotation comparision
peakAnnoList <- lapply(files, annotatePeak, TxDb=txdb,
                       tssRegion=c(-2000, 2000), verbose=FALSE)
png(file = file.path(FIGURES_DIR, "fig3_4.png"), width= 2000, height = 1000,  res=300)
## Override ChIPseeker's internal qualitative palette: the bins are an ORDERED
## distance progression, so they take a sequential ramp (near-TSS darkest).
p_dtss <- plotDistToTSS(peakAnnoList)
print(p_dtss + scale_fill_manual(values = pal_disttotss(p_dtss),
                                 guide = guide_legend(reverse = TRUE)))
while(!is.null(dev.list()))
dev.off()
```

```{r, include=TRUE, eval=FALSE, 'Figure 3.5. Annobar Distribution of Trh and Tgo ChIP-seq reads across annotated genomic features'}
png(file = file.path(FIGURES_DIR, "fig3_4.png"),  width = 2000, height = 1500, res = 300)
## Override ChIPseeker's getCols(): it assigns Paired-12 BY LEVEL ORDER, so the
## same feature can change colour between chapters. pal_feature() pins one colour
## per feature class and errors loudly on an unrecognised level.
p_anno <- plotAnnoBar(peakAnnoList)
print(p_anno + scale_fill_manual(values = pal_feature(p_anno),
                                 guide = guide_legend(reverse = TRUE)))
while(!is.null(dev.list()))
dev.off()
```

```{r, include=TRUE, eval=FALSE, 'Figure 3.6 UpSet Plot Distribution of genomic features within ChIP-peaks'}
# Figure 3.6A
peakAnno1 <- annotatePeak(files[[1]], tssRegion=c(-2000, 2000), TxDb=txdb, annoDb="org.Dm.eg.db")
peakAnno2 <- annotatePeak(files[[2]], tssRegion=c(-2000, 2000), TxDb=txdb, annoDb="org.Dm.eg.db")

# UPSET PLOTS. thesis_upset() is upsetplot() with ONE change: the bars carry
# genomic-context colour (promoter-containing / genic non-promoter / intergenic).
# upsetplot() itself cannot do this, it calls geom_bar() with no fill mapping,
# so every bar comes out the same dark grey.

png(file = file.path(FIGURES_DIR, "fig3_5_A.png"),  width = 2000, height = 1000, res = 210)
print(thesis_upset(peakAnno1, "Trh ChIP-seq"))
while(!is.null(dev.list()))
dev.off()

# Figure 3.6B
png(file = file.path(FIGURES_DIR, "fig3_5_B.png"),  width = 2000, height = 1000, res = 210)
print(thesis_upset(peakAnno2, "Tgo ChIP-seq"))
while(!is.null(dev.list()))
dev.off()
```

# Defining the Trh/Tgo heterodimer peak set

```{r, include=TRUE, eval=FALSE }
# pintersect(Trh_ChIP_peaks, Tgo_ChIP_peaks) and pintersect(Trh_clean, Tgo_clean) both yield 614 peaks: the two paths are equivalent for the Trh/Tgo intersection.
ovlp <- findOverlaps(Trh_ChIP_peaks, Tgo_ChIP_peaks
, minoverlap = 1, ignore.strand = TRUE); length(ovlp)
joint_peaks <- pintersect( # Joint Trh/Tgo binding regions were defined as the genomic intersections of independently called Trh and Tgo ChIP-seq peaks. Only loci where peaks from both datasets physically overlapped were retained, yielding a conservative set of candidate heterodimer binding regions.
  Trh_ChIP_peaks
[queryHits(ovlp)],
  Tgo_ChIP_peaks
[subjectHits(ovlp)]
); cat("joint_peaks count:", length(joint_peaks), "\n")
mcols(joint_peaks) <- mcols(joint_peaks)[, c("signalValue","qValue")]
```


## Overlap and statistical significance
```{r, include=TRUE, eval=FALSE, 'Figure 3.7a Venn diagram illustrating 614 overlapping genomic coordinates between Trh and Tgo ChIP-seq binding peaks (Trh-Tgo venn diagram overlap)'}
# Figure 3.7A.

# Venn diagram of the intersection of Trh and Tgo Peaks:
library(VennDiagram)
library(extrafont)
#font_import()
png(file = file.path(FIGURES_DIR, "fig3_6_A.png"), width = 1500, height = 1000, res = 200)
draw.pairwise.venn( 
    area1=length(Trh_clean),
    area2=length(Tgo_clean), 
    cross.area=length(joint_peaks), 
    category=c("Trh ChIP peaks", "Tgo ChIP peaks"), 
    fill=pal_venn_tgo("Trh"), # '0 degrees' - is at the bottom
    cex=2,
    cat.cex=2,
    cat.pos = c(198, 30), #c(-19, 30)
    #cat.pos = c(198, 150),
    cat.dist = c(0.075, 0.04))
    #cat.dist = 0.06)
while(!is.null(dev.list()))
dev.off()
```

```{r, include=FALSE, eval=FALSE, 'Writing a 614 interval "joint_peaks" object to a BED file'}
library(GenomicRanges)
write.table(
  {joint_bed<- data.frame(
  chr   = as.character(seqnames(joint_peaks)),
  start = start(joint_peaks) - 1,   # BED is 0-based
  end   = end(joint_peaks),
  name  = paste0("joint_", seq_along(joint_peaks)),
  score = 0,
  strand = "*")
  joint_bed
  },
  file = file.path(EXPORT_DIR, "Trh_Tgo_joint_peaks.bed"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
); View(joint_bed)
```


```{r load-perm-overlap-trh-tgo, include=TRUE, eval=TRUE}
# Loads the pre-computed Trh/Tgo overlap peakPermTest result.
obj <- readRDS(file.path(EXPORT_DIR, "perm_overlap_trh_tgo.rds"))
pt2_ <- obj$pt2_
cat(sprintf('Loaded pt2_ from HPC: observed=%d  null_mean=%.2f  p=%.3e\n',
            pt2_$cntOverlaps$observed,
            mean(pt2_$cntOverlaps$permuted),
            pt2_$cntOverlaps$pval))
```

```{r load-perm-enhancers-trh-tgo, include=TRUE, eval=TRUE}
# Loads pre-computed enhancer permutation results (all/promoter/distal).
obj <- readRDS(file.path(EXPORT_DIR, "perm_enhancers_trh_tgo.rds"))
pt_all      <- obj$pt_all
pt_promoter <- obj$pt_promoter
pt_distal   <- obj$pt_distal
print(obj$summary, row.names = FALSE)
```

```{r load-perm-cme, include=TRUE, eval=TRUE}
# Loads the pre-computed CME positional permTest result.
obj <- readRDS(file.path(EXPORT_DIR, "perm_cme.rds"))
pt <- obj$pt
print(obj$summary, row.names = FALSE)
```

```{r load-chipenrich-ch3, include=TRUE, eval=TRUE}
# Loads pre-computed ChIP-Enrich results (six tables).
obj <- readRDS(file.path(EXPORT_DIR, "chipenrich_ch3.rds"))
res_custom       <- obj$res_custom
res_custom_fet   <- obj$res_custom_fet
res_BP           <- obj$res_BP
res_MF           <- obj$res_MF
res_CC           <- obj$res_CC
res_reactome     <- obj$res_reactome
res_sig_BP       <- obj$res_sig_BP
res_sig_MF       <- obj$res_sig_MF
res_sig_CC       <- obj$res_sig_CC
res_sig_reactome <- obj$res_sig_reactome
# tracheal geneset size, recorded in results/reported_values.tsv
trach_entrez_n   <- obj$trach_entrez_n
comp_entrez_n    <- obj$comp_entrez_n
cat(sprintf('Loaded ChIP-Enrich: custom=%d rows | BP=%d (sig=%d) | MF=%d | CC=%d | reactome=%d\n',
            nrow(res_custom), nrow(res_BP), nrow(res_sig_BP),
            nrow(res_MF), nrow(res_CC), nrow(res_reactome)))
```

```{r load-chipenrich-ch3-distalsplit, include=TRUE, eval=TRUE}
# Loads the promoter-vs-distal diagnostic (translation-artefact test).
obj_ds <- readRDS(file.path(EXPORT_DIR, "chipenrich_ch3_distalsplit.rds"))
res_prox_bp_sig   <- obj_ds$res_prox_bp_sig
res_distal_bp_sig <- obj_ds$res_distal_bp_sig
n_promoter_ds     <- obj_ds$n_promoter
n_distal_ds       <- obj_ds$n_distal
cat(sprintf("Distal split loaded: proximal n=%d (%d sig terms) | distal n=%d (%d sig terms)\n",
            n_promoter_ds, nrow(res_prox_bp_sig),
            n_distal_ds,   nrow(res_distal_bp_sig)))
```


```{r, include=TRUE, eval=FALSE, 'Figure 3.7b Validating the Trh/Tgo overlap significance using permutation test using peakPermTest'}
# Figure 3.7B
# ---- Dependencies ----
library(regioneR)
# ---- Run the permutation test ----
# pt2_ is loaded by the `load-perm-overlap-trh-tgo` chunk earlier in this Rmd
# ---- Extract components ----
perm_counts  <- pt2_$cntOverlaps$permuted      # numeric vector, length 5000
obs_count    <- pt2_$cntOverlaps$observed       # single integer (614)
pval         <- pt2_$cntOverlaps$pval
zscore       <- pt2_$cntOverlaps$zscore
n_perm       <- length(perm_counts)
# ---- Density of null distribution ----
null_dens <- density(perm_counts, adjust = 1.2)
# 95th percentile of null (Ev_perm0.05 threshold)
ev_perm_95 <- quantile(perm_counts, 0.95)

# ---- Open PNG device ----
png(
  file   = file.path(FIGURES_DIR, "fig3_6_B.png"),
  width  = 3000,
  height = 1200,
  res    = 350
)
# ---- Layout: two panels side by side ----
# Left panel: null distribution zoomed in
# Right panel: observed value alone
# This avoids forcing both onto one compressed axis
layout(matrix(c(1, 2), nrow = 1), widths = c(5.5, 1))
# ---- Panel 1: Null distribution (zoomed) ----
par(mar = c(4, 4, 2, 0.5))
null_xlim <- c(0, max(perm_counts) * 1.4)
plot(
  null_dens$x, null_dens$y,
  type  = "n",
  xlim  = null_xlim,
  ylim  = c(0, max(null_dens$y) * 1.35),
  xlab  = "Count of overlaps (permuted)",
  ylab  = "Density",
  main  = paste0("Null distribution (", formatC(n_perm, big.mark = ""), " permutations)"),
  cex.lab  = 1,
  cex.axis = 1,
  las   = 1
)
# Shade null density
polygon(
  c(null_dens$x, rev(null_dens$x)),
  c(null_dens$y, rep(0, length(null_dens$y))),
  col    = adjustcolor(PAL_INK[["neutral"]], alpha.f = 0.5),
  border = NA
)
lines(null_dens$x, null_dens$y, col = PAL_INK[["outline"]], lwd = 1.5)
# 95th percentile line
abline(v = ev_perm_95, col = PAL_INK[["threshold"]], lwd = 1.5, lty = 2)
text(
  x      = ev_perm_95,
  y      = max(null_dens$y) * 1.25,
  labels = bquote(Ev[perm]^{0.05} == .(round(ev_perm_95))),
  col    = PAL_INK[["threshold"]],
  cex    = 1,
  pos    = 4   # place label to the right of the line
)
# Stats annotation, top-left, not competing with density
legend(
  "topright",
  legend = list(
    bquote(italic(Z) == .(round(zscore, 2))),
    bquote(italic(p) == .(formatC(pval, format = "e", digits = 2))),
    bquote(n[perm] == .(n_perm))
  ),
  bty  = "n",
  cex  = 1
)
# ---- Panel 2: Observed value ----
par(mar = c(4, 0.5, 2, 0))
plot(
  0, 0,
  type = "n",
  xlim = c(0, 1),
  ylim = c(0, 1),
  xaxt = "n", yaxt = "n",
  xlab = "", ylab = "",
  main = "Observed",
  bty  = "n",
  cex.main = 0.85
)
# Draw a single vertical bar representing Ev_obs
rect(
  xleft   = 0.3,
  ybottom = 0.05,
  xright  = 0.7,
  ytop    = 0.85,
  col     = PAL_HETERO[["Trh/Tgo"]],
  border  = NA
)
text(
  x      = 0.5,
  y      = 0.93,
  labels = bquote(Ev[obs] == .(obs_count)),
  col    = PAL_HETERO[["Trh/Tgo"]],
  cex    = 1,
  font   = 2
)
while (!is.null(dev.list()))
dev.off()
```

The overlap drawn in the Venn diagram (Figure 3.6) can read 690 while the number of overlapping peaks is 655, because a peak may overlap several peaks of the opposing set; the diagram counts each such configuration once, on the coordinates of the larger merged interval. The two lines below compare the two counts (655 and 690):

```{r,include=TRUE, eval=FALSE}
simple_overlap<-length(findOverlapsOfPeaks(Trh_clean, Tgo_clean)$peaklist$`Trh_clean///Tgo_clean`); simple_overlap
manual_overlap_of_every_instance <- length(unique(queryHits(findOverlaps(Trh_clean, Tgo_clean)))); manual_overlap_of_every_instance
```


# Functional contextualisation of Trh/Tgo binding

## Enhancer and promoter context

```{r, include=TRUE, eval=FALSE, 'Figure 3.7A. Screenshot of overlaps within ±1 kb of TSS'}
# Overlapping Trh/Tgo binding peaks with known Drosophila enhancers (Co-localisation with late embryonic enhancers (st10-17))
```

Permutation test for non-random overlap with promoters and enhancers:
```{r, include=TRUE, eval=FALSE, 'Figure 3.7B. Permutation test to show non-random overlap with promoters and enhancers'}
enhancers <- read.table(file.path(REFERENCE_LISTS, "merged_embryo_5-20_enhancers_dm6.bed"), header = FALSE, stringsAsFactors = FALSE)
colnames(enhancers) <- c("seqnames", "start", "end")
enhancers_gr <- GRanges(
  seqnames = enhancers$seqnames,
  ranges   = IRanges(start = enhancers$start, end = enhancers$end)
)
# ============================================================
# Figure 3.7B, Permutation test (RERUN with joint_peaks_filt)
# Uses the same 614-peak filtered set as all downstream analyses
# ============================================================

library(GenomicRanges)
library(regioneR)
library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)

# joint_peaks_filt should already exist (614 peaks, standard chromosomes)
# enhancers_gr should already exist from the enhancer bed file load
# If not, recreate:
standard_chrs <- c("chr2L","chr2R","chr3L","chr3R","chr4","chrX")
joint_peaks_filt <- keepSeqlevels(joint_peaks, standard_chrs, pruning.mode="coarse")
enhancers_filt <- keepSeqlevels(
  enhancers_gr,
  c("chr2L","chr2R","chr3L","chr3R","chr4","chrX"),
  pruning.mode = "coarse"
)

# Quick sanity check
cat("Peaks in analysis set:     ", length(joint_peaks_filt), "\n")
cat("Observed enhancer overlaps:", sum(countOverlaps(
  joint_peaks_filt, enhancers_filt, ignore.strand = TRUE) > 0), "\n")

# --- Promoter / distal split using the same TSS method as Fig 3_11 -----------
# Use 1-bp TSS points (consistent with distanceToTSS already on joint_peaks_filt)
# If $distanceToTSS is already annotated, use it directly:

if (!"distanceToTSS" %in% names(mcols(joint_peaks_filt))) {
  txdb   <- TxDb.Dmelanogaster.UCSC.dm6.ensGene
  txs    <- transcripts(txdb)
  tss_pt <- resize(txs, width = 1, fix = "start")
  ov_tss <- distanceToNearest(joint_peaks_filt, tss_pt)
  joint_peaks_filt$distanceToTSS <- mcols(ov_tss)$distance
}
promoter_peaks <- joint_peaks_filt[joint_peaks_filt$distanceToTSS <= 1000]
distal_peaks   <- joint_peaks_filt[joint_peaks_filt$distanceToTSS >  1000]

cat("Promoter peaks (≤1 kb):", length(promoter_peaks), "\n")
cat("Distal peaks   (>1 kb):", length(distal_peaks),   "\n")

# ============================================================
# Permutation tests - no masks are used, which is the more conservative approach
# ntimes = 5000 throughout; empirical p floor = 1/5001
# ============================================================
options(mc.cores = 1)
# --- All 614 peaks vs enhancers ----------------------------------------------
set.seed(777)
pt0 <- proc.time()
# pt_all is loaded by the `load-perm-enhancers-trh-tgo` chunk earlier in this Rmd
cat("pt_all done:", (proc.time() - pt0)["elapsed"], "sec\n")

# --- Promoter subset vs enhancers --------------------------------------------
set.seed(777)
pt0 <- proc.time()
# pt_promoter is loaded by the `load-perm-enhancers-trh-tgo` chunk earlier in this Rmd
cat("pt_promoter done:", (proc.time() - pt0)["elapsed"], "sec\n")

# --- Distal subset vs enhancers ----------------------------------------------
set.seed(777)
pt0 <- proc.time()
# pt_distal is loaded by the `load-perm-enhancers-trh-tgo` chunk earlier in this Rmd
cat("pt_distal done:", (proc.time() - pt0)["elapsed"], "sec\n")

# ============================================================
# Extract and display results
# ============================================================

extract_perm_stats <- function(perm, label) {
  obs  <- perm$numOverlaps$observed
  perm_vals <- perm$numOverlaps$permuted
  data.frame(
    Context   = label,
    Observed  = obs,
    Null_mean = round(mean(perm_vals), 2),
    Null_SD   = round(sd(perm_vals),   2),
    Z_score   = round((obs - mean(perm_vals)) / sd(perm_vals), 2),
    p_emp     = perm$numOverlaps$pval,
    N_perm    = length(perm_vals),
    stringsAsFactors = FALSE
  )
}

perm_results_table <- rbind(
  extract_perm_stats(pt_all,      "All Trh/Tgo peaks (n=614)"),
  extract_perm_stats(pt_promoter, "Promoter-proximal (≤1 kb)"),
  extract_perm_stats(pt_distal,   "Distal (>1 kb)")
)

print(perm_results_table, row.names = FALSE)

# Save results table for export
write.csv(perm_results_table,
          file      = file.path(EXPORT_DIR, "perm_results_table_filt.csv"),
          row.names = FALSE,
          quote     = FALSE)
```


## Construction of a tissue-relevant reference gene set: trh-coexpression dataset

## Nearest-gene context and GO enrichment near Trh/Tgo peaks
```{r, include=TRUE, eval=FALSE, 'Figure 3.10 Venn diagram and Gene Ontology enrichment of TS and SG genes near 614 peaks'}
# ------------------------------------------------------------
# FIGURE 3.9, Trh/Tgo peak proximity analysis, hypergeometric & permutation
# Logic: 614 Trh/Tgo peaks -> genes with TSS within 1 kb -> overlap with
#        TS/SG reference list -> statistical tests
# ------------------------------------------------------------

# ---- 0. Libraries ----
library(GenomicRanges)
library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
library(org.Dm.eg.db)
library(AnnotationDbi)
library(regioneR)
library(VennDiagram)
setwd(file.path(PROJECT_DIR, ""))
txdb          <- TxDb.Dmelanogaster.UCSC.dm6.ensGene
standard_chrs <- c("chr2L","chr2R","chr3L","chr3R","chr4","chrX")

# ---- 1. Filter peaks to standard chromosomes ----
# joint_peaks is the pre-existing 614-peak GRanges object
joint_peaks_filt <- keepSeqlevels(
  joint_peaks,
  standard_chrs,
  pruning.mode = "coarse"
)
stopifnot(length(joint_peaks_filt) == 614)
cat("Peaks after chromosome filtering:", length(joint_peaks_filt), "\n")

# ---- 2. Build single-base TSS coordinates ----
# promoters(upstream=0, downstream=1) gives width-1 ranges exactly at the TSS.
# This avoids the double-window inflation caused by using upstream=1000 as a
# range and then applying a 1 kb distance filter on top of it.

tss_gr <- promoters(
  genes(txdb),
  upstream   = 0,
  downstream = 1
)
tss_gr <- keepSeqlevels(tss_gr, standard_chrs, pruning.mode = "coarse")
tss_gr <- trim(tss_gr)

# Step 1: get all mappings as a list to identify ambiguous Ensembl IDs
flybase_list <- mapIds(
  org.Dm.eg.db,
  keys      = names(tss_gr),
  keytype   = "ENSEMBL",
  column    = "FLYBASE",
  multiVals = "list"
)

# Step 2: identify ambiguous entries (more than one FlyBase ID per Ensembl ID)
ambiguous <- lengths(flybase_list) > 1
cat("Ensembl IDs with ambiguous FlyBase mapping:", sum(ambiguous), "\n")

# Step 3: assign first value for all, then NA-mask the ambiguous ones
flybase_vec <- mapIds(
  org.Dm.eg.db,
  keys      = names(tss_gr),
  keytype   = "ENSEMBL",
  column    = "FLYBASE",
  multiVals = "first"
)
flybase_vec[ambiguous] <- NA   # conservative: treat ambiguous as unmapped

# Step 4: assign to GRanges
tss_gr$flybase <- flybase_vec

# Diagnostic
cat("TSSs with FlyBase ID:   ", sum(!is.na(tss_gr$flybase)), "\n")
cat("TSSs without FlyBase ID:", sum( is.na(tss_gr$flybase)), "\n")

# ---- 3. Load TS/SG reference gene list ----
ts_sg_genes <- unique(na.omit(
  read.csv(file.path(REFERENCE_LISTS, "trh_ts_sg_tissues_reference.csv"),
           stringsAsFactors = FALSE)$genes
))
cat("TS/SG reference genes:", length(ts_sg_genes), "\n")   # expect 2053
#######

# Checks for loss of gene identifiers during mapping between annotation systems
cat("=== Identifier loss check ===\n")
# Step 1: FBgn -> Entrez (used by ChIP-Enrich)
fbgn_to_entrez <- mapIds(org.Dm.eg.db,
  keys = ts_sg_genes, keytype = "FLYBASE",
  column = "ENTREZID", multiVals = "first")
cat("FBgn → Entrez: ", sum(!is.na(fbgn_to_entrez)), 
    "mapped,", sum(is.na(fbgn_to_entrez)), "lost\n")
# Step 2: FBgn -> Ensembl (used by gene length test)
fbgn_to_ensembl <- mapIds(org.Dm.eg.db,
  keys = ts_sg_genes, keytype = "FLYBASE",
  column = "ENSEMBL", multiVals = "first")
cat("FBgn → Ensembl:", sum(!is.na(fbgn_to_ensembl)),
    "mapped,", sum(is.na(fbgn_to_ensembl)), "lost\n")

# Step 3: are the unmapped genes the same across conversions?
lost_to_entrez  <- ts_sg_genes[is.na(fbgn_to_entrez)]
lost_to_ensembl <- ts_sg_genes[is.na(fbgn_to_ensembl)]
cat("Shared unmapped (lost in both):", 
    length(intersect(lost_to_entrez, lost_to_ensembl)), "\n")
cat("Lost only to Entrez:", 
    length(setdiff(lost_to_entrez, lost_to_ensembl)), "\n")
cat("Lost only to Ensembl:", 
    length(setdiff(lost_to_ensembl, lost_to_entrez)), "\n")
######


# ---- 4. Find all genes with a TSS within 1 kb of any Trh/Tgo peak ----
# findOverlaps with maxgap=1000 captures every peak-TSS pair within 1 kb.
# A peak near two genes contributes both genes, this is biologically correct
# for a promoter-proximal transcription factor like Trh/Tgo.
peak_tss_hits <- findOverlaps(
  joint_peaks_filt,
  tss_gr,
  maxgap = 1000L,
  ignore.strand = TRUE
)

# Extract unique FlyBase IDs of all genes near any peak
nearby_flybase <- unique(na.omit(
  tss_gr$flybase[subjectHits(peak_tss_hits)]
))
cat("Unique genes within 1 kb of a peak:", length(nearby_flybase), "\n")

# ---- 5. Overlap with TS/SG reference list ----
overlap_genes <- intersect(nearby_flybase, ts_sg_genes)
cat("Overlap (peak-proximal ∩ TS/SG expressed):", length(overlap_genes), "\n")
print(overlap_genes)

# Save overlap gene list for GO enrichment and the supplementary table
write.csv(
  data.frame(FlyBase_ID = overlap_genes),
  file      = file.path(EXPORT_DIR, "trh_tgo_peaks_vs_ts_sg_overlap_genes.csv"),
  row.names = FALSE,
  quote     = FALSE
)

# ---- 6. Venn diagram ----
png(file.path(FIGURES_DIR, "fig3_10_A_clean.png"),
    width = 1800, height = 1600, res = 400)

VennDiagram::draw.pairwise.venn(
  area1      = length(ts_sg_genes),       # 2053
  area2      = length(nearby_flybase),    # genes near peaks
  cross.area = length(overlap_genes),     # intersection
  category   = c(
    "Genes expressed in tracheal\nsystem and salivary gland",
    paste0("Gene TSSs within\n±1 kb of ",
           length(joint_peaks_filt),
           " Trh/Tgo\nheterodimer peaks")
  ),
  fill    = rev(pal_venn_tissue("Trh/Tgo", "Tracheal+SG")),
  alpha   = rep(PAL_VENN_ALPHA, 2),
  col     = rep(PAL_INK[["outline"]], 2),
  cex     = 1.1,
  cat.cex = 1.1,
  cat.col = rep(PAL_INK[["annotation"]], 2),
  cat.pos = c(-15, 163),
  cat.dist = c(0.06, 0.1),
  margin  = 0.1
)
while (!is.null(dev.list())) dev.off()

# ---- 7. Hypergeometric test ----
# N: all annotated Drosophila genes (background universe)
# A: genes near Trh/Tgo peaks (draws from urn)
# B: TS/SG reference genes (successes in urn)
# k: observed overlap (successes drawn)
N <- length(keys(org.Dm.eg.db, keytype = "FLYBASE"))
A <- length(nearby_flybase)
B <- length(ts_sg_genes)
k <- length(overlap_genes)

p_hyper  <- phyper(k - 1, m = B, n = N - B, k = A, lower.tail = FALSE)
expected <- A * (B / N)
fold_enr <- k / expected

cat(sprintf(
  "\nHypergeometric test\nN=%d  A=%d  B=%d  k=%d\nExpected=%.2f  Fold enrichment=%.2f  p=%.3e\n",
  N, A, B, k, expected, fold_enr, p_hyper
))

# ---- 8. Permutation test ----
# Label each TSS as TS/SG or not, then pass only the TS/SG subset to the
# evaluation function to avoid carrying the full tss_gr through forked workers.
tss_gr$is_ts_sg <- (!is.na(tss_gr$flybase)) & (tss_gr$flybase %in% ts_sg_genes)
tss_ts_sg       <- tss_gr[tss_gr$is_ts_sg == TRUE]
cat("TS/SG TSSs for permutation:", length(tss_ts_sg), "\n")

# Evaluation function:
# counts unique TS/SG genes with a TSS within 1 kb of the (randomised) peaks.
# The ... absorbs per.chromosome and genome passed by permTest internally.
perm_eval <- function(A, B, ..., tss_ts_sg) {
  if (length(tss_ts_sg) == 0L) return(0L)
  hits <- findOverlaps(A, tss_ts_sg, maxgap = 1000L, ignore.strand = TRUE)
  length(unique(subjectHits(hits)))
}

options(mc.cores = 1)   # avoids mclapply serialisation failures
set.seed(777)

# pt (CME perm) is loaded by the `load-perm-cme` chunk earlier in this Rmd

# ---- 9. Permutation results table ----
obs  <- pt$perm_eval$observed
null <- pt$perm_eval$permuted

tbl_perm <- data.frame(
  Context        = "Trh/Tgo peaks within 1 kb of TS/SG TSS",
  Observed       = obs,
  Null_mean      = round(mean(null), 2),
  Null_SD        = round(sd(null),   2),
  Z_score        = round((obs - mean(null)) / sd(null), 3),
  p_empirical    = mean(null >= obs),
  N_permutations = length(null)
)
print(tbl_perm)
# p < 0.0002
```

## enrichGO on the 75 tracheal/salivary-gland genes near peaks (not reported)
```{r, include=TRUE, eval=FALSE, 'Figure3.9b Gene Ontology enrichment of genes near peaks'}
# Figure3.9b Gene Ontology enrichment of 75 (TS and SG) genes near 614 peaks
png(file = file.path(FIGURES_DIR, "fig3_9B.png"), width = 1500, height = 1000, res = 210)
ego <- enrichGO(gene = overlap_genes, OrgDb = org.Dm.eg.db, keyType = "FLYBASE", ont = "ALL", pAdjustMethod = "BH")
dotplot(ego)
while(!is.null(dev.list()))
dev.off()
```

## Gene-length bias of the tracheal/salivary-gland gene set
```{r, include=TRUE, eval=FALSE}
library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
gene_lengths <- width(genes(TxDb.Dmelanogaster.UCSC.dm6.ensGene))
ts_sg_ensembl <- mapIds(
  org.Dm.eg.db,
  keys     = ts_sg_genes,
  keytype  = "FLYBASE",
  column   = "ENSEMBL",
  multiVals = "first"
)
ts_sg_ensembl <- na.omit(ts_sg_ensembl)
all_gene_names <- names(genes(TxDb.Dmelanogaster.UCSC.dm6.ensGene))

ts_sg_lengths  <- gene_lengths[all_gene_names %in% ts_sg_ensembl]
other_lengths   <- gene_lengths[!all_gene_names %in% ts_sg_ensembl]

cat("Median TS/SG gene length:  ", median(ts_sg_lengths), "\n")# Median TS/SG gene length:   4685

cat("Median other gene length:  ", median(other_lengths), "\n")# Median other gene length:   1512

wilcox.test(ts_sg_lengths, other_lengths)
# Wilcoxon rank sum test with continuity correction
# data:  ts_sg_lengths and other_lengths
# W = 24279486, p-value < 2.2e-16
# alternative hypothesis: true location shift is not equal to 0
```
TS/SG genes are significantly longer than background genes, so the hypergeometric p-value is inflated; the enrichment claim therefore rests on ChIP-Enrich, which corrects for locus length.

## Quantitative enrichment near tracheal loci
```{r, include=TRUE, eval=FALSE, 'Table 3.2 Gene length and total locus size adjusted enrichment'}
# ============================================================
# ChIP-Enrich analysis, complete script in execution order
# Section 3.5.4 (main text) + Appendix robustness table
# All uses joint_peaks_filt (614 peaks, standard chromosomes)
# ============================================================
suppressPackageStartupMessages({
  library(chipenrich)
  library(org.Dm.eg.db)
  library(AnnotationDbi)
  library(readr)
  library(dplyr)
  library(GenomicRanges)
  library(stringr)
  library(ggplot2)
})
# ============================================================
# 0. Build peaks_df
# ============================================================
peaks_df <- data.frame(
  chr   = as.character(seqnames(joint_peaks_filt)),
  start = start(joint_peaks_filt),
  end   = end(joint_peaks_filt),
  stringsAsFactors = FALSE
)
cat("Peaks in analysis:", nrow(peaks_df), "\n")
# ============================================================
# PART A, MAIN TEXT (Section 3.5.4 + Table 3.2)
# ============================================================
# --- A1. Build custom tracheal geneset ---------------------------------------
all_dm6 <- unique(as.character(keys(org.Dm.eg.db, keytype = "ENTREZID")))
read_fbgn <- function(path) {
  df <- if (grepl("\\.csv$", path, ignore.case = TRUE))
    suppressMessages(readr::read_csv(path, show_col_types = FALSE))
  else
    suppressMessages(readr::read_tsv(path, show_col_types = FALSE))
  v <- if ("genes" %in% names(df)) df[["genes"]] else df[[1]]
  v <- unique(trimws(as.character(v)))
  v[grepl("^FBgn\\d+$", v)]
}
map_fbgn_to_entrez <- function(fbgn_vec) {
  if (length(fbgn_vec) == 0) return(character(0))
  m <- AnnotationDbi::select(org.Dm.eg.db,
                             keys    = fbgn_vec,
                             keytype = "FLYBASE",
                             columns = "ENTREZID")
  m <- dplyr::filter(m, !is.na(ENTREZID)) |> dplyr::distinct(ENTREZID)
  as.character(m$ENTREZID)
}
tracheal_file <- file.path(REFERENCE_LISTS, "trh_ts_sg_tissues_reference.csv")
trach_fbgn    <- read_fbgn(tracheal_file)
trach_entrez  <- map_fbgn_to_entrez(trach_fbgn)
trach_entrez  <- intersect(trach_entrez, all_dm6)
comp_entrez   <- setdiff(all_dm6, trach_entrez)
stopifnot(length(intersect(trach_entrez, comp_entrez)) == 0)
# Print geneset sizes: tracheal genes and complement genes
cat("Tracheal geneset size:", length(trach_entrez), "Entrez IDs\n")
cat("Complement geneset size:", length(comp_entrez), "Entrez IDs\n")
genesets_file <- file.path(EXPORT_DIR, "geneset_tracheal_vs_complement.tsv")
custom_gs <- bind_rows(
  tibble(geneset = "tracheal_custom",        gene_id = trach_entrez),
  tibble(geneset = "All_dm6_minus_tracheal", gene_id = comp_entrez)
)
readr::write_delim(custom_gs, genesets_file, delim = "\t", col_names = FALSE)
# --- A2. PRIMARY ANALYSIS, logistic regression (main text, Table 3.2) -------

# res_custom is loaded by the `load-chipenrich-ch3` chunk earlier in this Rmd

cat("\n===== TABLE 3.2 NUMBERS (main text) =====\n")
print(res_custom[, c("Description", "Odds.Ratio", "P.value", "FDR",
                     "N.Geneset.Peak.Genes")])
# or = 2.27, p = 8.93e-9 for tracheal_custom, reported in Table 3.2
# or = 0.44, p = 8.93e-9 for complement, reported in Table 3.2
# FDR equals the p-value because only 2 terms are tested (BH correction of 2 identical values)

# ============================================================
# PART B, APPENDIX (robustness check, Fisher exact test)
# ============================================================
# --- B1. Fisher exact test (no locus-length correction) ----------------------
# res_custom_fet is loaded by the `load-chipenrich-ch3` chunk earlier in this Rmd
# A warning about locusdef may appear here; the analysis uses a 1kb locus definition
# --- B2. Build Appendix comparison table -------------------------------------
robustness_df <- data.frame(
  Method     = c("ChIP-Enrich (logistic regression, locus-length corrected)",
                 "Fisher exact test (count-based, no locus-length correction)"),
  Odds_Ratio = c(
    res_custom[res_custom$Description == "tracheal_custom", "Odds.Ratio"],
    res_custom_fet[res_custom_fet$Description == "tracheal_custom", "Odds.Ratio"]
  ),
  P_value    = c(
    res_custom[res_custom$Description == "tracheal_custom", "P.value"],
    res_custom_fet[res_custom_fet$Description == "tracheal_custom", "P.value"]
  ),
  stringsAsFactors = FALSE
)
robustness_df$Significant <- ifelse(robustness_df$P_value < 0.05, "Yes", "No")

cat("\n===== APPENDIX TABLE: ROBUSTNESS CHECK =====\n")
print(robustness_df, row.names = FALSE)

all_enriched <- all(robustness_df$Odds_Ratio > 1)
all_sig      <- all(robustness_df$P_value < 0.05) 

if (all_enriched & all_sig) {
  cat("\nCONCLUSION: Both methods agree — enrichment is significant and",
      "consistent. Robustness sentence in main text is justified.\n")
} else {
  cat("\nWARNING: Methods disagree. Review before keeping robustness sentence.\n")
}

write.csv(robustness_df,
          file.path(EXPORT_DIR, "chipenrich_robustness_comparison.csv"),
          row.names = FALSE, quote = FALSE)
cat("Saved: chipenrich_robustness_comparison.csv\n")
# ============================================================
# PART C, MAIN TEXT (Figure 3.11 dot plots, GO and Reactome enrichment)
# ============================================================
# Shared dot plot function
plot_chipenrich_dotplot <- function(res_sig, title_str, outfile,
                                    dpi = 400,
                                    max_terms = 20,
                                    base_height_per_term = 120,   # was 120
                                    min_height = 900,            # was 900
                                    max_height = 2400,            # was 2400
                                    width = 2400,
                                    fdr_cut = 0.05) {  # FDR cut res_sig was
                                                       # filtered at; sets the
                                                       # colourbar's natural axis
  if (nrow(res_sig) == 0) {
    message("No significant terms for: ", title_str, " — skipping plot")
    return(invisible(NULL))
  }
  res_plot <- head(res_sig, max_terms)
  res_plot$Description <- str_wrap(res_plot$Description, width = 45)  # was 45
  res_plot$Description <- factor(res_plot$Description,
                                 levels = rev(res_plot$Description))
  n_terms   <- nrow(res_plot)
  n_lines   <- sum(vapply(strsplit(as.character(res_plot$Description), "\n"),
                          length, integer(1)))

  few_terms <- n_terms <= 8
  base_size <- if (n_terms <= 8) 11 else if (n_terms <= 15) 10 else 9

  # ---- ADD THESE TWO LINES ----
  legend_pad <- if (few_terms) 350 else 0
  height     <- min(max(n_lines * base_height_per_term, min_height) + legend_pad,
                    max_height)
  
  p <- ggplot(res_plot,
              aes(x = log2(Odds.Ratio), y = Description,
                  fill = FDR, size = N.Geneset.Peak.Genes)) +
    geom_point_seq() +
    # FDR: low end is interesting, rev = FALSE puts dark at low FDR.
    # values/domain fix colourbar range to [0, fdr_cut] (res_sig filtered at FDR < fdr_cut); avoids flat bar when all sig terms share one FDR (GOMF n=1; Reactome n=10, FDR=0.00696).
    # n.breaks = 3, default six tick labels overprint on the 3 cm bar.
    scale_fill_seq(values = res_plot$FDR, domain = c(0, fdr_cut),
                     n.breaks = 3, name = "FDR") +
    scale_size_continuous(name = "Peak-associated genes (n)", range = c(2, 6)) +
    geom_vline(xintercept = 0, linetype = "dashed",
               colour = PAL_INK[["annotation"]]) +
    scale_y_discrete(expand = expansion(add = 0.8)) +
    labs(x = expression(log[2]~"(Odds ratio)"), y = NULL) +
    theme_bw(base_size = base_size)

  if (few_terms) {
    p <- p +
      theme(plot.title = element_blank(), plot.subtitle = element_blank(),
            plot.margin = margin(t = 12, r = 10, b = 12, l = 8),
            legend.position  = "bottom",
            legend.box       = "horizontal",
            legend.key.height = unit(0.30, "cm"),
            legend.key.width  = unit(0.30, "cm"),
            legend.title = element_text(size = rel(0.8)),
            legend.text  = element_text(size = rel(0.72))) +
      guides(
        colour = guide_colourbar(barwidth = unit(3, "cm"),
                                 barheight = unit(0.35, "cm"),
                                 title.position = "top"),
        size   = guide_legend(nrow = 1, override.aes = list(colour = PAL_INK[["outline"]]),
                              title.position = "top")
      )
  } else {
    p <- p +
      theme(plot.title = element_blank(), plot.subtitle = element_blank(),
            plot.margin = margin(t = 14, r = 8, b = 10, l = 8),
            legend.position = "right",
            legend.key.height = unit(0.35, "cm"),
            legend.key.width  = unit(0.35, "cm"),
            legend.title = element_text(size = rel(0.8)),
            legend.text  = element_text(size = rel(0.75)),
            legend.spacing.y = unit(0.15, "cm")) +
      guides(
        colour = guide_colourbar(barheight = unit(2.5, "cm"),
                                 barwidth  = unit(0.4, "cm")),
        size   = guide_legend(override.aes = list(colour = PAL_INK[["outline"]]))
      )
  }
  png(file.path(FIGURES_DIR, outfile),
      width = width, height = height, res = dpi)
  print(p)
  while (!is.null(dev.list())) dev.off()
  message(sprintf("Saved: %s [%d x %d px @ %d dpi, %d terms, %d label-lines]",
                  outfile, width, height, dpi, n_terms, n_lines))
}
# ---- GO Biological Process -----------------------------------------------
# res_BP is loaded by the `load-chipenrich-ch3` chunk earlier in this Rmd
print(res_sig_BP[, c("Description","Odds.Ratio","P.value","FDR","N.Geneset.Peak.Genes")])
plot_chipenrich_dotplot(res_sig_BP, "GO Biological Process enrichment near Trh/Tgo peaks",
                        "fig3_11_chipenrich_GOBP.png")

# --- GO Molecular Function -----------------------------------------------
# res_MF is loaded by the `load-chipenrich-ch3` chunk earlier in this Rmd
print(res_sig_MF[, c("Description","Odds.Ratio","P.value","FDR","N.Geneset.Peak.Genes")])
plot_chipenrich_dotplot(res_sig_MF, "GO Molecular Function enrichment near Trh/Tgo peaks",
                        "figD_1_chipenrich_GOMF.png")
# --- 3. GO Cellular Component -----------------------------------------------
# res_CC is loaded by the `load-chipenrich-ch3` chunk earlier in this Rmd
print(res_sig_CC[, c("Description","Odds.Ratio","P.value","FDR","N.Geneset.Peak.Genes")])
plot_chipenrich_dotplot(res_sig_CC, "GO Cellular Component enrichment near Trh/Tgo peaks",
                        "fig3_12_chipenrich_GOCC.png")
# ---- Reactome ---------------------------------------------------------
# res_reactome is loaded by the `load-chipenrich-ch3` chunk earlier in this Rmd
print(res_sig_reactome[, c("Description","Odds.Ratio","P.value","FDR","N.Geneset.Peak.Genes")])
plot_chipenrich_dotplot(res_sig_reactome, "Reactome pathway enrichment near Trh/Tgo peaks",
                        "fig3_13_chipenrich_reactome.png")
```

# Motif architecture within Trh/Tgo-bound regions

## Definition of regulatory context and CME stratification

```{r, include=TRUE, eval=FALSE, 'regulatory-classification-nearest-TSS'}
# Regulatory annotation
library(GenomicRanges)
library(dplyr)
library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
txdb <- TxDb.Dmelanogaster.UCSC.dm6.ensGene
# 1-bp TSS points (strand-aware)
txs    <- transcripts(txdb)
tss_pt <- resize(txs, width = 1, fix = "start")
# Distance to nearest TSS *point*
joint_peaks_filt <- keepSeqlevels(
  joint_peaks_filt,   # use the 614-peak object explicitly
  c("chr2L","chr2R","chr3L","chr3R","chr4","chrX"),
  pruning.mode = "coarse"
)
#stopifnot(length(joint_peaks_filt) == 614)
ov_tss <- distanceToNearest(joint_peaks_filt, tss_pt)
joint_peaks_filt$distanceToTSS <- mcols(ov_tss)$distance
# Enhancer overlap
enhancers <- read.table(
  file.path(REFERENCE_LISTS, "merged_embryo_5-20_enhancers_dm6.bed"),
  header = FALSE,
  stringsAsFactors = FALSE
)
colnames(enhancers) <- c("seqnames", "start", "end")
enhancers_gr <- GRanges(
  seqnames = enhancers$seqnames,
  ranges   = IRanges(start = enhancers$start, end = enhancers$end)
)
joint_peaks_filt$overlaps_enhancer <-
  countOverlaps(joint_peaks_filt, enhancers_gr, ignore.strand = TRUE) > 0
# Regulatory class (canonical)
joint_peaks_filt$Regulatory_Class <- case_when(
  joint_peaks_filt$distanceToTSS <= 1000 & joint_peaks_filt$overlaps_enhancer ~
    "Promoter-proximal enhancer (annotated)",
  joint_peaks_filt$distanceToTSS <= 1000 ~
    "Promoter-proximal peak (non-annotated)",
  joint_peaks_filt$distanceToTSS > 1000 & joint_peaks_filt$overlaps_enhancer ~
    "Distal enhancer (annotated)",
  TRUE ~
    "Distal peak (non-annotated)"
)
# Summary table
# dplyr::count() is qualified explicitly because matrixStats::count() can mask
# the dplyr version when a package that Depends on SummarizedExperiment (such as
# DiffBind or chipenrich) is attached; matrixStats is pulled in transitively via
# MatrixGenerics/SummarizedExperiment/DelayedArray. Qualifying the call makes the
# chunk independent of attach order, consistent with dplyr::filter being qualified
# further down this chapter.
summary_df <- as.data.frame(joint_peaks_filt) %>%
  dplyr::count(Regulatory_Class) %>%
  dplyr::mutate(
    Fraction = n / sum(n),
    Percent  = 100 * Fraction
  ); summary_df
## Recode for plotting (same logic as §3.4.1)
bar_df_35 <- summary_df %>%
  dplyr::filter(Regulatory_Class %in% c(
    "Promoter-proximal peak (non-annotated)",
    "Promoter-proximal enhancer (annotated)",
    "Distal enhancer (annotated)",
    "Distal peak (non-annotated)"
  )) %>%
  dplyr::mutate(
    Category = dplyr::recode(
      Regulatory_Class,
      "Promoter-proximal peak (non-annotated)"  = "Promoter (≤1 kb, nearest TSS)",
      "Promoter-proximal enhancer (annotated)"  = "Promoter + enhancer overlap",
      "Distal enhancer (annotated)"             = "Distal enhancer",
      "Distal peak (non-annotated)"             = "Distal peak"
    ),
    Regulatory_Domain = ifelse(
      Category %in% c(
        "Promoter (≤1 kb, nearest TSS)",
        "Promoter + enhancer overlap"
      ),
      "Promoter-proximal",
      "Distal"
    )
  )
#---------------------------- Plotting Figure 3.12 -----------------------------
rc <- table(joint_peaks_filt$Regulatory_Class)
lv311 <- c("Distal enhancer\n(annotated)", "Distal peak\n(non-annotated)",
           "Promoter\n(\u22641 kb, nearest TSS)", "Promoter + enhancer\noverlap")
df311 <- data.frame(
  Context = factor(lv311, levels = lv311),
  Panel   = factor(c("Distal","Distal","Promoter-proximal","Promoter-proximal"),
                   levels = c("Distal","Promoter-proximal")),
  n       = c(as.integer(rc["Distal enhancer (annotated)"]),
              as.integer(rc["Distal peak (non-annotated)"]),
              as.integer(rc["Promoter-proximal peak (non-annotated)"]),
              as.integer(rc["Promoter-proximal enhancer (annotated)"])),
  stringsAsFactors = FALSE
)
df311$Pct   <- 100 * df311$n / length(joint_peaks_filt)   # denominator = 614
df311$Count <- sprintf("n = %d", df311$n)
fills311 <- pal_context(c("distal_enh","distal_unann","prox_peak","prox_enh"))
names(fills311) <- lv311
p311 <- thesis_bar(df311, x = "Context", y = "Pct", fill = "Context",
                   value_fmt = "%.1f%%", count = "Count", facet = "Panel",
                   fills = fills311, y_max = 78,
                   y_lab = "Percentage of Trh/Tgo peaks (%)")
ggsave(file.path(FIGURES_DIR, "fig3_7.png"), p311,
       width = 8, height = 4.3, dpi = 300, bg = "white")
```

## Adding CME and Distance to TSS to joint_peaks_filt
```{r, include=TRUE, eval=FALSE, fig-cme-strength, fig.cap="Figure 3.13. Occurrence of the canonical CNS midline element (CME)"}
# fig3_13.png, CME vs Trh/Tgo binding strength (Figure 3.13 in the text)
library(BSgenome.Dmelanogaster.UCSC.dm6)
library(Biostrings)
genome <- BSgenome.Dmelanogaster.UCSC.dm6
# --- Canonical CME definition (E-box ACGTG), scanned on both strands ----------
# CME presence = a match to ACGTG (plus) or its reverse complement CACGT
# (minus) on the FILTERED joint peak set. This is the single canonical CME
# definition used throughout the chapter and by the saveRDS() checkpoint later
# in this file, which recomputes the identical both-strand result.
# The stopifnot() check below checks that scanning is performed on both
# strands, giving 186 CME-positive peaks.
seqs   <- getSeq(genome, joint_peaks_filt)
cme_plus  <- vcountPattern("ACGTG", seqs, fixed = TRUE)
cme_minus <- vcountPattern("CACGT", seqs, fixed = TRUE)   # rev-comp of ACGTG
joint_peaks_filt$CME <- (cme_plus + cme_minus) > 0
stopifnot(length(joint_peaks_filt) == 614L,
          sum(joint_peaks_filt$CME) == 186L)              # both-strand canonical
# CME BED export
cme_peaks    <- joint_peaks_filt[joint_peaks_filt$CME]
no_cme_peaks <- joint_peaks_filt[!joint_peaks_filt$CME]
to_bed <- function(gr){
  df <- as.data.frame(gr)[, c("seqnames","start","end")]
  df$start <- pmax(0, df$start - 1L)  # BED is 0-based
  df
}
write.table(
  to_bed(cme_peaks),
  file.path(EXPORT_DIR, "CME_positive_joint_peaks_filt.bed"),
  sep = "\t", quote = FALSE,
  row.names = FALSE, col.names = FALSE
); length(cme_peaks)
write.table(
  to_bed(no_cme_peaks),
  file.path(EXPORT_DIR, "CME_negative_joint_peaks_filt.bed"),
  sep = "\t", quote = FALSE,
  row.names = FALSE, col.names = FALSE
); length(no_cme_peaks)
#------------------------------#
library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
library(GenomicFeatures)
txdb <- TxDb.Dmelanogaster.UCSC.dm6.ensGene
# ---- 1. Distance to nearest TSS ----
tss <- promoters(txdb, upstream = 0, downstream = 1)
tss <- keepSeqlevels(tss, 
                     c("chr2L","chr2R","chr3L","chr3R","chr4","chrX"),
                     pruning.mode = "coarse")

dist_hits <- distanceToNearest(joint_peaks_filt, tss, 
                                ignore.strand = TRUE)

# Initialise to NA then fill matched rows
joint_peaks_filt$distanceToTSS <- NA_integer_
joint_peaks_filt$distanceToTSS[queryHits(dist_hits)] <- 
    mcols(dist_hits)$distance

# ---- 2. Enhancer overlap flag ----
# enhancers_filt must exist, with keepSeqlevels applied, before this step is run
joint_peaks_filt$overlaps_enhancer <- 
    countOverlaps(joint_peaks_filt, 
                  enhancers_filt, 
                  ignore.strand = TRUE) > 0
# ---- 3. Sanity check before proceeding ----
table(is.na(joint_peaks_filt$distanceToTSS))  # should be all FALSE
table(joint_peaks_filt$overlaps_enhancer)      # should give TRUE/FALSE counts
#-------------------- CME enrichment by context---------------------------------
df <- as.data.frame(joint_peaks_filt) %>%
  dplyr::mutate(
    Context = case_when(
      distanceToTSS <= 1000 ~ "Promoter",
      overlaps_enhancer     ~ "Distal enhancer (annotated)",
      TRUE                  ~ "Distal peak (non-annotated)"
    ),
    CME = as.factor(CME)
  )
#-----------------Fisher test of 'Promoter vs distal enhancer'------------------
df_pe <- df %>% dplyr::filter(Context %in% c("Promoter","Distal enhancer (annotated)"))
tab_context_cme <- table(df_pe$Context, df_pe$CME)
tab_context_cme
fisher.test(tab_context_cme)
#----------Fisher test of 'Distal enhancer vs distal non-annotated--------------
df_distal <- df %>% dplyr::filter(Context != "Promoter")
tab_distal_cme <- table(
  df_distal$Context == "Distal enhancer (annotated)",
  df_distal$CME
)
tab_distal_cme
fisher.test(tab_distal_cme)
#---------------- Plotting Figure 3.13. CME vs ChIP signal----------------------
png(file = file.path(FIGURES_DIR, "fig3_13.png"),
  width = 1500, height = 1000, res = 300)
library(ggplot2)
pval <- wilcox.test(signalValue ~ CME, data = df)$p.value
p_fmt_12 <- if (pval < 0.001) paste0("italic(p)==", formatC(pval, format = "e", digits = 2)) else paste0("italic(p)==", round(pval, 3))
ggplot(df, aes(x = CME, y = signalValue)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  scale_x_discrete(labels = c("FALSE" = "CME-negative", "TRUE" = "CME-positive")) +
  ylab("Trh/Tgo ChIP signal (signalValue)") +
  annotate(
    "text",
    x     = 1.5,
    y     = max(df$signalValue, na.rm = TRUE),
    label = p_fmt_12,
    parse = TRUE,
    size  = 5
  ) +
  theme_bw()
while (!is.null(dev.list())) dev.off()
```


```{r, include=TRUE, eval=FALSE, 'Figure 3.14. Promoter vs distal enhancer signal'}
# fig3_15.png, CME-containing Trh/Tgo peaks, distal enrichment (Figure 3.14)
png(file = file.path(FIGURES_DIR, "fig3_15.png"),
  width = 1500, height = 1000, res = 300
)
library(ggplot2)
library(dplyr)
df_signal <- df %>%
  dplyr::filter(Context %in% c("Promoter", "Distal enhancer (annotated)"))
## Sample sizes. The labels are named so that each is bound to its class
## (an unnamed vector would be matched to the sorted levels).
lev_13 <- c("Promoter", "Distal enhancer (annotated)")
legend_labels <- c(
  "Promoter" = paste0("Promoter (n=", sum(df_signal$Context == "Promoter"), ")"),
  "Distal enhancer (annotated)" = paste0("Distal enhancer (n=", sum(df_signal$Context == "Distal enhancer (annotated)"), ")")
)
stopifnot(setequal(names(legend_labels), unique(df_signal$Context)))
## Statistics
pval <- wilcox.test(signalValue ~ Context, data = df_signal)$p.value
# ---- Live p-value formatting for ggplot annotate() ----
  pval_13 <- wilcox.test(signalValue ~ Context, data = df_signal)$p.value

# Format for parse = TRUE: must be valid R expression syntax
  if (pval_13 < 0.001) {
  parts <- strsplit(formatC(pval_13, format = "e", digits = 2), "e")[[1]]
  p_label_13 <- paste0("italic(p)==",
                       parts[1], " %*% 10^{", as.integer(parts[2]), "}")
} else {
  p_label_13 <- paste0("italic(p)==", round(pval_13, 3))
}
ggplot(df_signal, aes(x = signalValue, colour = Context, fill = Context)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  theme_bw() +
  xlab("Trh/Tgo ChIP signal") +
  ylab("Density") +
  scale_colour_manual(
    values = PAL_CONTEXT_LABELS[lev_13], breaks = lev_13,
    labels = legend_labels
  ) +
  scale_fill_manual(
    values = PAL_CONTEXT_LABELS[lev_13], breaks = lev_13,
    labels = legend_labels
  ) +
  theme(
    legend.position = c(0.7, 0.7),
    legend.background = element_rect(fill = "white", colour = "black"),
    legend.title = element_blank()
  ) +
  ## p-value annotation
    annotate(
    "text",
    x      = Inf,
    y      = Inf,
    label  = p_label_13,
    parse  = TRUE,
    hjust  = 1.1,
    vjust  = 1.5,
    size   = 5
  )
while (!is.null(dev.list())) dev.off()
## Class medians: promoter-proximal peaks carry the higher signal.
med_sig <- tapply(df_signal$signalValue, df_signal$Context, median)
stopifnot(med_sig[["Promoter"]] > med_sig[["Distal enhancer (annotated)"]])
save_values("ch3_signal_context", list(
  medSignalPromoter  = formatC(med_sig[["Promoter"]], format = "f", digits = 1),
  medSignalDistalEnh = formatC(med_sig[["Distal enhancer (annotated)"]], format = "f", digits = 1)
))
```


```{r,include=TRUE, eval=FALSE}
library(BSgenome.Dmelanogaster.UCSC.dm6)
genome <- BSgenome.Dmelanogaster.UCSC.dm6

seqs_cme_pos <- getSeq(genome, cme_peaks)
names(seqs_cme_pos) <- paste0("peak_", seq_along(seqs_cme_pos))
writeXStringSet(seqs_cme_pos, file.path(EXPORT_DIR, "CME_positive_sequences.fa"))

seqs_cme_neg <- getSeq(genome, no_cme_peaks)
names(seqs_cme_neg) <- paste0("peak_", seq_along(seqs_cme_neg))
writeXStringSet(seqs_cme_neg, file.path(EXPORT_DIR, "CME_negative_sequences.fa"))
```


## *De novo* motif discovery and triage
```{bash, include=TRUE, eval=FALSE}
cd reference
wget https://hgdownload.soe.ucsc.edu/goldenPath/dm6/database/rmsk.txt.gz
gunzip rmsk.txt.gz
# Extract just the BED coordinates (columns 6,7,8 are chrom, start, end)
awk '{print $6"\t"$7"\t"$8}' rmsk.txt > dm6_repeats.bed
```

```{r, bed-export, include=TRUE, eval=FALSE}
# Generating a BED file to upload to MEME-ChIP for PWM motif discovery
library(GenomicRanges)
## Drop peaks shorter than 5 bp
length(joint_peaks_filt)        # before
joint_peaks_meme <- joint_peaks_filt[width(joint_peaks_filt) >= 5]
joint_peaks_unique <- unique(joint_peaks_meme)
length(joint_peaks_unique)

library(BSgenome.Dmelanogaster.UCSC.dm6)
genome <- BSgenome.Dmelanogaster.UCSC.dm6
seqs <- getSeq(genome, joint_peaks_unique)
names(seqs) <- paste0("peak_", seq_along(seqs))
writeXStringSet(seqs, file.path(EXPORT_DIR, "Trh_Tgo_joint_peaks_filt_raw.fa"))
getwd()
```

```{r}
joint_bed <- data.frame(
  chr    = as.character(seqnames(joint_peaks_unique)),
  start  = start(joint_peaks_unique) - 1,
  end    = end(joint_peaks_unique),
  name   = paste0("joint_peak_", seq_along(joint_peaks_unique)),
  score  = 0,
  strand = "*",
  stringsAsFactors = FALSE
)
write.table(joint_bed,
  file = file.path(EXPORT_DIR, "Trh_Tgo_joint_peaks_native_width_for_MEME_ChIP.bed"),
  sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

```{r, include=TRUE, eval=FALSE}
# Genomic-coordinate BED (peak_N names) -> repeat-masked DISCOVERY input only.
jb <- data.frame(as.character(seqnames(joint_peaks_unique)),
                 start(joint_peaks_unique) - 1L, end(joint_peaks_unique),
                 paste0("peak_", seq_along(joint_peaks_unique)), 0L, ".")
write.table(jb, file.path(EXPORT_DIR, "Trh_Tgo_joint_peaks_filt.bed"),
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

```{bash, include=TRUE, eval=FALSE}
cd $PROJECT_DIR
meme-chip --version          

cat reference/motif_databases/FLY/fly_factor_survey.meme \
reference/motif_databases/JASPAR/JASPAR2024_CORE_insects_non-redundant_v2.meme \
> reference/motif_databases/combined_drosophila.meme
DB="reference/motif_databases/combined_drosophila.meme"   # Fly Factor Survey + JASPAR Insects

meme-chip \
 -oc data/chapter3/memechip_Trh_Tgo_masked  \
 -db "$DB"  \
 -order 0  \
 -meme-nmotifs 10 \
 -minw 6 -maxw 30 \
 -seed 777 \
 data/chapter3/Trh_Tgo_joint_peaks_filt_repeatmasked.fa
```

# Hard-masking for STREME (used in Chapter 4)
```{bash, include=TRUE, eval=FALSE}
# DISCOVERY input ONLY (STREME / MEME-ChIP). NOT a FIMO input.
# Mask repeats on the GENOME, then extract peaks, so chrom names match.
cd $PROJECT_DIR
[ -s reference/dm6_hardmasked.fa ] || bedtools maskfasta -fi reference/dm6.fa -bed reference/dm6_repeats.bed -fo reference/dm6_hardmasked.fa -mc N
bedtools getfasta -nameOnly -fi reference/dm6_hardmasked.fa \
  -bed data/chapter3/Trh_Tgo_joint_peaks_filt.bed \
  -fo  data/chapter3/Trh_Tgo_joint_peaks_filt_repeatmasked.fa
n=$(grep -v '^>' data/chapter3/Trh_Tgo_joint_peaks_filt_repeatmasked.fa | tr -cd N | wc -c)
echo "N bases masked: $n"; [ "$n" -gt 0 ] || { echo "ABORT — masking did nothing"; exit 1; }
```


MEME-ChIP was run on meme-suite.org with the settings below (see the Methods chapter).

| Setting | Value | Rationale |
|---|---|---|
| Tool | MEME-ChIP | Combines MEME, STREME, CentriMo |
| Input | Full 614-peak FASTA | Sufficient n for CentriMo |
| Sequence alphabet | DNA | — |
| Motif discovery | MEME + STREME both enabled | STREME catches lower-frequency motifs |
| Min motif width | 6 | Captures minimal E-box |
| Max motif width | 20 | Covers composite TF sites |
| Number of motifs | 10 | Enough without noise inflation |
| Background model | 0-order Markov from input sequences | Matched to the composition of the input |
| CentriMo | Enabled | Tests central enrichment relative to background |
| Motif databases | JASPAR Insects 2024 + Fly Factor Survey | Most relevant for *Drosophila* |
| FIMO | Enabled | FIMO output is used by the downstream chunks |


The following motifs were rescored with FIMO against all 614 Trh/Tgo
heterodimer peak sequences. The CME was included as a manually specified
reference motif; of the two motifs MEME returned, Motif1 was excluded as a
low-complexity artefact and Motif2 was the sole de novo motif retained.

```
MEME output (memechip_Trh_Tgo_masked/meme_out/meme.xml), verbatim:

  MEME-1  AWWWWWWWWWWWAAHAWATWWAAAAWWWW   width 29  sites 268  E = 4.0e-114  EXCLUDED
  MEME-2  GTRCACTGARARAAA                 width 15  sites  39  E = 1.1e-023  RETAINED (= Motif2)

  Reference motif, manually specified:
  CME     ACGTG                           canonical CNS midline element core
```

```{r motif-fimo-inputs, include=TRUE, eval=FALSE, echo=TRUE}
# The FIMO runs use MEME's own PWM files rather than a typed consensus string.
# Motif2's PWM is STRCACTGARARAAA.meme.
# EXPORT_DIR is this chapter's data/chapter3 (set in the setup chunk)
motif2_pwm <- file.path(EXPORT_DIR, "STRCACTGARARAAA.meme")
meme_xml   <- file.path(EXPORT_DIR, "memechip_Trh_Tgo_masked", "meme_out", "meme.xml")
stopifnot(file.exists(motif2_pwm), file.exists(meme_xml))

# Confirm the retained motif is the one the text quotes, so a re-run cannot
# silently pick up a different motif. Read from the MEME header directly, no
# package dependency, so this check works in any session.
hdr <- readLines(motif2_pwm, warn = FALSE)
mline <- grep("^MOTIF ", hdr, value = TRUE)
wline <- grep("w=\\s*[0-9]+", hdr, value = TRUE)[1]
m2_name <- sub("^MOTIF\\s+(\\S+).*", "\\1", mline[1])
m2_w    <- as.integer(sub(".*w=\\s*([0-9]+).*", "\\1", wline))
cat("Motif2 PWM        :", mline[1], "\n")
cat("consensus / width :", m2_name, "/", m2_w, "\n")
stopifnot(identical(m2_name, "STRCACTGARARAAA"), m2_w == 15L)

# Confirm MEME returned this as its second motif.
xml <- readLines(meme_xml, warn = FALSE)
mnames <- regmatches(xml, gregexpr('<motif id="motif_[0-9]+" name="[^"]+"', xml))
mnames <- unlist(mnames)[1:2]
cat("MEME motifs 1-2   :\n"); cat(paste0("  ", mnames, collapse = "\n"), "\n")
stopifnot(any(grepl("GTRCACTGARARAAA", mnames)))
```

## Contextual enrichment of retained motifs
```{r, include=TRUE, eval=FALSE}
library(GenomicRanges)

fimo <- read.table(
  file.path(EXPORT_DIR, "fimo_joint","fimo.tsv"),
  header       = TRUE,
  sep          = "\t",
  comment.char = "#"
)

# Inspect
unique(fimo$motif_id)
head(fimo)

# Filter
fimo_filt <- fimo[fimo$q.value < 0.05, ]
cat("Motif2 hits after q < 0.05 filter:", nrow(fimo_filt), "\n")

# Convert peak-local FIMO coordinates to genomic coordinates
peak_idx <- as.integer(sub("peak_", "", fimo_filt$sequence_name))
peak_starts <- start(joint_peaks_filt)[peak_idx]
peak_chr    <- as.character(seqnames(joint_peaks_filt))[peak_idx]
fimo_gr <- GRanges(
  seqnames = peak_chr,
  ranges = IRanges(
    start = peak_starts + fimo_filt$start - 1L,
    end   = peak_starts + fimo_filt$stop  - 1L
  ),
  strand   = fimo_filt$strand,
  motif_id = fimo_filt$motif_id,
  score    = fimo_filt$score,
  qvalue   = fimo_filt$q.value
)

# Verify seqlevels match joint_peaks_filt
intersect(seqlevels(joint_peaks_filt), seqlevels(fimo_gr))
# Expect: "chr2L" "chr2R" "chr3L" "chr3R" "chr4" "chrX"

# Annotate
joint_peaks_filt$Motif2 <- countOverlaps(
  joint_peaks_filt,
  fimo_gr,
  ignore.strand = TRUE
) > 0

cat("Peaks with Motif2:", sum(joint_peaks_filt$Motif2), "\n")
cat("Unique peaks with Motif2:",
    sum(countOverlaps(joint_peaks_filt, fimo_gr, ignore.strand = TRUE) > 0), "\n")
```

```{r motif2-contextual-enrichment, include=TRUE, eval=FALSE}
#---------- Motif2 (TTTYTYHCAGTGTA): contextual enrichment ----------
joint_peaks_filt$Enhancer_Status <- case_when(
  joint_peaks_filt$Regulatory_Class == "Distal enhancer (annotated)"            ~ "Annotated_enhancer",
  joint_peaks_filt$Regulatory_Class == "Promoter-proximal enhancer (annotated)" ~ "Annotated_enhancer",
  joint_peaks_filt$Regulatory_Class == "Promoter-proximal peak (non-annotated)" ~ "Non_annotated",
  joint_peaks_filt$Regulatory_Class == "Distal peak (non-annotated)"            ~ "Non_annotated"
)

# ---- Rebuild tab_refined from live columns ----
stopifnot("Regulatory_Class" %in% names(mcols(joint_peaks_filt)))
stopifnot("Motif2"          %in% names(mcols(joint_peaks_filt)))

df_motif_refined <- as.data.frame(mcols(joint_peaks_filt))

tab_refined <- table(
  df_motif_refined$Regulatory_Class,
  df_motif_refined$Motif2
)

cat("tab_refined:\n")
print(tab_refined)
fisher.test(tab_refined)
```

```{r,include=TRUE, eval=FALSE}
# ---- Figure 3.15 --------------------------------------------------------
get_count <- function(tbl, row, col) {
  if (row %in% rownames(tbl) && col %in% colnames(tbl)) as.integer(tbl[row, col]) else 0L
}
lv314 <- c("Distal enhancer\n(annotated)", "Promoter-proximal\nenhancer (annotated)",
           "Promoter-proximal\npeak (non-annotated)", "Distal peak\n(non-annotated)")
n_de <- as.integer(rc["Distal enhancer (annotated)"])
n_pe <- as.integer(rc["Promoter-proximal enhancer (annotated)"])
n_pn <- as.integer(rc["Promoter-proximal peak (non-annotated)"])
n_dn <- as.integer(rc["Distal peak (non-annotated)"])
m_de <- get_count(tab_refined, "Distal enhancer (annotated)",            "TRUE")
m_pe <- get_count(tab_refined, "Promoter-proximal enhancer (annotated)", "TRUE")
m_pn <- get_count(tab_refined, "Promoter-proximal peak (non-annotated)", "TRUE")
m_dn <- get_count(tab_refined, "Distal peak (non-annotated)",            "TRUE")
df314 <- data.frame(
  Context = factor(lv314, levels = lv314),
  Panel   = factor(c("Annotated enhancers","Annotated enhancers",
                     "Non-annotated peaks","Non-annotated peaks"),
                   levels = c("Annotated enhancers","Non-annotated peaks")),
  Pct     = c(100*m_de/n_de, 100*m_pe/n_pe, 100*m_pn/n_pn, 100*m_dn/n_dn),
  Count   = sprintf("(%d/%d)", c(m_de,m_pe,m_pn,m_dn), c(n_de,n_pe,n_pn,n_dn)),
  stringsAsFactors = FALSE
)
fills314 <- pal_context(c("distal_enh","prox_enh","prox_peak","distal_unann"))
names(fills314) <- lv314
p314 <- thesis_bar(df314, x = "Context", y = "Pct", fill = "Context",
                   value_fmt = "%.1f%%", count = "Count", facet = "Panel",
                   fills = fills314, y_max = 58, y_lab = "Peaks with Motif2 (%)",
                   global_p = fisher.test(tab_refined)$p.value)
ggsave(file.path(FIGURES_DIR, "fig3_16_motif2_context.png"), p314,
       width = 8, height = 4.3, dpi = 300, bg = "white")
```
## Adding CME to joint_peak_filt
```{r cme-genomic-coordinates,include=TRUE, eval=FALSE}
library(BSgenome.Dmelanogaster.UCSC.dm6)

genome <- BSgenome.Dmelanogaster.UCSC.dm6

# Extract genomic peak sequences
peak_seqs <- getSeq(genome, joint_peaks_filt)

# Search both strands
cme_plus  <- vmatchPattern("ACGTG", peak_seqs)
cme_minus <- vmatchPattern("CACGT", peak_seqs)

# Helper
hits_to_granges <- function(hits, peaks) {

  idx <- rep(seq_along(peaks), elementNROWS(hits))

  GRanges(
    seqnames = seqnames(peaks)[idx],
    ranges = IRanges(
      start = start(peaks)[idx] +
              unlist(start(hits)) - 1,
      width = width(unlist(hits))
    )
  )
}

cme_gr_plus  <- hits_to_granges(cme_plus, joint_peaks_filt)
cme_gr_minus <- hits_to_granges(cme_minus, joint_peaks_filt)

cme_genomic <- c(cme_gr_plus, cme_gr_minus)

# Annotate peaks
joint_peaks_filt$CME <- countOverlaps(
  joint_peaks_filt,
  cme_genomic,
  ignore.strand = TRUE
) > 0

# Verify
table(joint_peaks_filt$CME)
table(joint_peaks_filt$Motif2)

# Peaks with both
peaks_both <- joint_peaks_filt[
  joint_peaks_filt$CME &
  joint_peaks_filt$Motif2
]
length(peaks_both)
saveRDS(joint_peaks_filt, file.path(EXPORT_DIR, "joint_peaks_filt.rds"))
```

# Motif2 deconfounding tests (tests 1 and 3)
```{r test1-motif2-deconfounding-regression, include=TRUE, eval=FALSE, echo=TRUE}
# ---- TEST 1: does Motif2 dosage predict primary Trh/Tgo occupancy, ----
# ---- independent of GC, width, TSS distance, CME? ----
# Outcome: MACS3 fold-enrichment over input (data/chapter3/trh_joint_macs3_fe.csv,
# the value at the summit of the IDR-paired replicate peaks, averaged over the two
# replicates). signalValue is mean pooled-ChIP coverage and is not input-normalised;
# the same model on signalValue is kept as fit_signal for comparison.
# Runs on the LIVE objects produced earlier in this Rmd:
#   joint_peaks_filt (614 peaks, with signalValue/distanceToTSS/CME mcols)
#   fimo_gr          (motif instances; must be subset to Motif2)
EXPORT_DIR <- file.path(PROJECT_DIR, "data/chapter3")

library(GenomicRanges); library(Biostrings)
library(BSgenome.Dmelanogaster.UCSC.dm6)

stopifnot(exists("joint_peaks_filt"), length(joint_peaks_filt) == 614)
stopifnot(exists("fimo_gr"))

# --- isolate Motif2 from CME / Motif_1 contamination --------------
cat("motif_id contents of fimo_gr:\n"); print(table(fimo_gr$motif_id))
motif2_gr <- fimo_gr[fimo_gr$motif_id == "STRCACTGARARAAA"]
stopifnot(length(motif2_gr) > 0)   # 0 means the FIMO motif labels differ
cat("Motif2 instances used for dosage:", length(motif2_gr), "\n")

genome <- BSgenome.Dmelanogaster.UCSC.dm6
seqs   <- getSeq(genome, joint_peaks_filt)

fe_tab <- read.csv(file.path(EXPORT_DIR, "trh_joint_macs3_fe.csv"))
stopifnot(nrow(fe_tab) == length(joint_peaks_filt),
          all(fe_tab$seqnames == as.character(seqnames(joint_peaks_filt))),
          all(fe_tab$start == start(joint_peaks_filt)), all(fe_tab$end == end(joint_peaks_filt)),
          !anyNA(fe_tab$macs3_fold_enrichment))

df <- data.frame(
  peak_score   = fe_tab$macs3_fold_enrichment,
  signal_value = joint_peaks_filt$signalValue,
  width        = width(joint_peaks_filt),
  dist_tss     = joint_peaks_filt$distanceToTSS,
  has_cme      = as.integer(joint_peaks_filt$CME),
  n_motif2     = countOverlaps(joint_peaks_filt, motif2_gr, ignore.strand = TRUE),
  gc_content   = as.numeric(letterFrequency(seqs, "GC", as.prob = TRUE))
)
df$has_motif2    <- as.integer(df$n_motif2 > 0)
df$log_dist_tss  <- log10(df$dist_tss + 1)

# --- full model + collinearity ----------------------------------------------
fit_full <- lm(peak_score ~ gc_content + width + log_dist_tss + has_cme + n_motif2,
               data = df)
fit_signal <- lm(signal_value ~ gc_content + width + log_dist_tss + has_cme + n_motif2,
                 data = df)   # raw coverage as outcome, for comparison
if (!requireNamespace("car", quietly = TRUE)) install.packages("car")
cat("\n--- VIF (flag > 5) ---\n");   print(car::vif(fit_full))
cat("\n--- full model ---\n");       print(summary(fit_full))

# --- nested F-test: does Motif2 add over confounders? -----------------------
fit_reduced <- lm(peak_score ~ gc_content + width + log_dist_tss + has_cme, data = df)
cat("\n--- nested F-test ---\n");     print(anova(fit_reduced, fit_full))
cat("adj R^2 without Motif2:", summary(fit_reduced)$adj.r.squared, "\n")
cat("adj R^2 with    Motif2:", summary(fit_full)$adj.r.squared,    "\n")

# --- binary robustness ------------------------------------------------------
fit_bin <- lm(peak_score ~ gc_content + width + log_dist_tss + has_cme + has_motif2,
              data = df)
cat("\n--- binary Motif2 ---\n")
print(summary(fit_bin)$coefficients["has_motif2", , drop = FALSE])

# === REFEREE PRE-EMPTS (added after TEST1 review) ===========================
# (a) log-outcome: fold-enrichment is right-skewed; does the sign/significance hold?
fit_log <- lm(log1p(peak_score) ~ gc_content + width + log_dist_tss + has_cme + n_motif2,
              data = df)
cat("\n--- (a) log-outcome ---\n")
print(summary(fit_log)$coefficients["n_motif2", , drop = FALSE])

# (b) GC-only adjustment: strips possible mediator conditioning (TSS dist/width)
fit_gc <- lm(peak_score ~ gc_content + n_motif2, data = df)
cat("\n--- (b) GC-only adjustment ---\n")
print(summary(fit_gc)$coefficients["n_motif2", , drop = FALSE])

# (c) heteroscedasticity-robust SEs (HC3) on the full model
if (!requireNamespace("sandwich", quietly = TRUE)) install.packages("sandwich")
if (!requireNamespace("lmtest",   quietly = TRUE)) install.packages("lmtest")
cat("\n--- (c) HC3-robust SEs, full model ---\n")
print(lmtest::coeftest(fit_full, vcov. = sandwich::vcovHC(fit_full, type = "HC3")))

saveRDS(list(full = fit_full, reduced = fit_reduced, binary = fit_bin,
             log = fit_log, gc_only = fit_gc, signal = fit_signal,
             vif = car::vif(fit_full), df = df,
             outcome = "MACS3 fold-enrichment over input; element signal holds the same model on raw coverage"),
        file.path(EXPORT_DIR, "test1_regression_binding.rds"))
```

```{r test3-motif2-phylop-conservation, include=TRUE, eval=FALSE, echo=TRUE}
# ---- TEST 3: is Motif2 under purifying selection relative to its OWN ----
# ---- immediate flanking sequence? (within-locus paired design) ----
# PREREQUISITE: download the dm6 phyloP27way conservation track from UCSC:
#   cd $PROJECT_DIR/data/chapter3
#   curl -O https://hgdownload.soe.ucsc.edu/goldenPath/dm6/phyloP27way/dm6.phyloP27way.bw
EXPORT_DIR <- file.path(PROJECT_DIR, "data/chapter3")

library(GenomicRanges)
library(rtracklayer)

BW_PATH <- file.path(EXPORT_DIR, "dm6.phyloP27way.bw")
FLANK_W <- 50L
stopifnot(file.exists(BW_PATH))
stopifnot(exists("joint_peaks_filt"), exists("fimo_gr"))

# --- isolate Motif2 instances; this is the MASTER object & ordering ----------
cat("motif_id contents of fimo_gr:\n"); print(table(fimo_gr$motif_id))
motif2_gr <- fimo_gr[fimo_gr$motif_id == "STRCACTGARARAAA"]
motif2_gr <- sort(unique(motif2_gr))           # keep strand (needed for core)
stopifnot(length(motif2_gr) > 0, all(width(motif2_gr) == 15L))
cat("Motif2 instances entering phyloP test:", length(motif2_gr), "\n")

# unstranded copy for scoring (phyloP is strand-agnostic); same order/coords
motif2_us <- motif2_gr; strand(motif2_us) <- "*"

# --- chromosome-naming compatibility check -----------------------------------
bwf    <- BigWigFile(BW_PATH)
bw_chr <- seqnames(seqinfo(bwf))
if (!all(as.character(seqnames(motif2_us)) %in% bw_chr)) {
  warning("seqlevel mismatch between motif2_gr and bigWig; check chr naming")
  print(setdiff(unique(as.character(seqnames(motif2_us))), bw_chr))
}
sl <- seqlengths(seqinfo(bwf))

# --- matched flanks: same-locus, immediately adjacent, clamped to chr ---------
left_fl  <- GRanges(seqnames(motif2_us),
                    IRanges(start(motif2_us) - FLANK_W, start(motif2_us) - 1L))
right_fl <- GRanges(seqnames(motif2_us),
                    IRanges(end(motif2_us) + 1L, end(motif2_us) + FLANK_W))
clamp <- function(g) {
  L <- sl[as.character(seqnames(g))]
  start(g) <- pmax(start(g), 1L); end(g) <- pmin(end(g), L); g
}
left_fl <- clamp(left_fl); right_fl <- clamp(right_fl)

# drop any instance whose flank abuts a NEIGHBOURING motif (contamination check)
fl_touch_motif <- countOverlaps(left_fl,  motif2_us, ignore.strand = TRUE) +
  countOverlaps(right_fl, motif2_us, ignore.strand = TRUE)

# --- per-region mean phyloP (version-robust extractor) ------------------------
region_mean <- function(bwf, gr) {
  s <- summary(bwf, gr, size = 1L, type = "mean")
  if (methods::is(s, "RleList")) {
    v <- vapply(s, function(r){x <- as.numeric(r); if(length(x)) x[1] else NA_real_}, numeric(1))
  } else {
    v <- vapply(s, function(g){if(length(g)) as.numeric(score(g))[1] else NA_real_}, numeric(1))
  }
  v[is.nan(v)] <- NA_real_; as.numeric(v)
}

motif_pp <- region_mean(bwf, motif2_us)
left_pp  <- region_mean(bwf, left_fl)
right_pp <- region_mean(bwf, right_fl)
flank_pp <- rowMeans(cbind(left_pp, right_pp), na.rm = TRUE)
flank_pp[is.nan(flank_pp)] <- NA_real_

# --- CACTG core (positions 4-8 of the 15-mer, strand-aware) -------------------
# Derived from motif2_gr -> identical ordering to flank_pp / fl_touch_motif.
core <- motif2_gr
plus <- as.logical(strand(motif2_gr) == "+")
start(core)[plus]  <- start(motif2_gr)[plus] + 3L
end(core)[plus]    <- start(motif2_gr)[plus] + 7L
start(core)[!plus] <- end(motif2_gr)[!plus]  - 7L
end(core)[!plus]   <- end(motif2_gr)[!plus]  - 3L
strand(core) <- "*"
core_pp <- region_mean(bwf, core)

# --- assemble paired tables, drop NA pairs and motif-adjacent flanks ----------
ok       <- is.finite(flank_pp) & fl_touch_motif == 0
pp_k     <- data.frame(motif = motif_pp, flank = flank_pp)[ok & is.finite(motif_pp), ]
pp_core  <- data.frame(core  = core_pp,  flank = flank_pp)[ok & is.finite(core_pp),  ]
cat("Full-motif pairs retained:", nrow(pp_k),
    "| CACTG-core pairs retained:", nrow(pp_core),
    "of", length(motif2_gr), "\n")
stopifnot(nrow(pp_k) >= 10)

# --- paired tests -------------------------------------------------------------
wt      <- wilcox.test(pp_k$motif,   pp_k$flank,   paired = TRUE)
wt_core <- wilcox.test(pp_core$core, pp_core$flank, paired = TRUE)
prop_up <- mean(pp_k$motif > pp_k$flank)

cat("\n=== TEST 3: Motif2 phyloP vs own ±", FLANK_W, "bp flank ===\n", sep = "")
cat("full motif : median", round(median(pp_k$motif), 4),
    "vs flank", round(median(pp_k$flank), 4),
    "| diff", round(median(pp_k$motif - pp_k$flank), 4),
    "| prop>flank", round(prop_up, 3),
    "| p", formatC(wt$p.value, format = "e", digits = 2), "\n")
cat("CACTG core : median", round(median(pp_core$core), 4),
    "vs flank", round(median(pp_core$flank), 4),
    "| diff", round(median(pp_core$core - pp_core$flank), 4),
    "| p", formatC(wt_core$p.value, format = "e", digits = 2), "\n")

saveRDS(list(full = list(table = pp_k,    wilcox = wt,      prop_up = prop_up),
             core = list(table = pp_core, wilcox = wt_core),
             flank_w = FLANK_W, n_instances = length(motif2_gr)),
        file.path(EXPORT_DIR, "test3_phylop_conservation.rds"))
```

```{r cme-motif2-spacing,include=TRUE, eval=FALSE}
# Get spacing between CME and Motif2 centres within co-occurring peaks
cme_in_both    <- subsetByOverlaps(cme_genomic, peaks_both, ignore.strand = TRUE)
motif2_in_both <- subsetByOverlaps(fimo_gr, peaks_both, ignore.strand = TRUE)

# For each peak, find minimum CME-to-motif2 centre distance
spacings <- sapply(seq_along(peaks_both), function(i) {
  cme_i <- subsetByOverlaps(
    cme_in_both, peaks_both[i], ignore.strand = TRUE)
  m3_i  <- subsetByOverlaps(
    motif2_in_both, peaks_both[i], ignore.strand = TRUE)
  if (length(cme_i) == 0 || length(m3_i) == 0) return(NA)
  min(abs(outer(mid(ranges(cme_i)), mid(ranges(m3_i)), "-")))
})

spacings <- spacings[!is.na(spacings)]
cat("Median CME-Motif2 spacing:", median(spacings), "bp\n")
cat("Range:", min(spacings), "to", max(spacings), "bp\n")

png(file.path(FIGURES_DIR, "fig3_17_cme_motif2_spacing.png"),
    width = 2400, height = 1200, res = 300)
hist(spacings,
     breaks = 21,
     main = "",  # <- removes title
     xlab = "Minimum centre-to-centre distance (bp)",
     xlim = c(0, 300),
     col  = PAL_HETERO[["Trh/Tgo"]],
     border = PAL_INK[["on_mark"]])
while(!is.null(dev.list()))
dev.off()
```
```{r,include=TRUE, eval=FALSE}
# Identify which peaks have CME-motif2 spacing <= 50 bp
close_peaks_idx <- which(spacings <= 250)
# Extract those peaks from peaks_both
close_peaks <- peaks_both[close_peaks_idx]
# Print their genomic coordinates and metadata
as.data.frame(close_peaks)
library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
library(org.Dm.eg.db)
txdb <- TxDb.Dmelanogaster.UCSC.dm6.ensGene
genes_gr <- genes(txdb)
# Find nearest gene for each close-spacing peak
nearest_idx <- nearest(close_peaks, genes_gr)
nearest_genes <- genes_gr[nearest_idx]
# Get gene symbols
gene_ids <- nearest_genes$gene_id
gene_symbols <- mapIds(
  org.Dm.eg.db,
  keys      = gene_ids,
  column    = "SYMBOL",
  keytype   = "FLYBASE",
  multiVals = "first"
)
data.frame(
  peak          = as.character(granges(close_peaks)),
  spacing       = spacings[close_peaks_idx],
  gene          = gene_symbols,
  distanceToTSS = close_peaks$distanceToTSS
)
```

## Motif2 in CME-positive vs CME-negative peaks

```{r cme-stratified-motif2, include=TRUE, eval=FALSE}
# ============================================================
# FIGURE 3.16, Motif2 frequency in CME-positive vs CME-negative peaks
# Tests whether CACTGA (Motif2) substitutes for ACGTG (CME) in CME-negative
# peaks, or whether the two motifs are independent of each other.
# Requires: joint_peaks_filt$CME and joint_peaks_filt$Motif2 both TRUE/FALSE
# ============================================================
library(ggplot2)
library(dplyr)

stopifnot("CME"    %in% names(mcols(joint_peaks_filt)))
stopifnot("Motif2" %in% names(mcols(joint_peaks_filt)))

df_cme_m2 <- as.data.frame(mcols(joint_peaks_filt)) %>%
  dplyr::select(CME, Motif2, Regulatory_Class) %>%
  dplyr::mutate(
    CME_status = ifelse(CME,     "CME-positive", "CME-negative"),
    M2_status  = ifelse(Motif2, "Motif2 present", "Motif2 absent")
  )

# ---- 1. Primary contingency table: CME status x Motif2 ----
tab_cme_m2 <- table(CME_status = df_cme_m2$CME_status,
                    Motif2    = df_cme_m2$Motif2)
cat("=== CME status x Motif2 contingency table ===\n")
print(tab_cme_m2)

freq_cme_m2 <- prop.table(tab_cme_m2, margin = 1) * 100
cat("\nMotif2 prevalence (%) within each CME class:\n")
print(round(freq_cme_m2, 1))

# Fisher exact test
ft_cme_m2 <- fisher.test(tab_cme_m2)
cat("\nFisher exact test (CME status x Motif2):\n")
cat("  OR =", round(ft_cme_m2$estimate, 3), "\n")
cat("  p  =", formatC(ft_cme_m2$p.value, format = "e", digits = 2), "\n")
cat("\nInterpretation:\n")
cat("  OR > 1 -> Motif2 enriched in CME-positive (co-occurrence)\n")
cat("  OR < 1 -> Motif2 enriched in CME-negative (substitution)\n")
cat("  OR ~ 1 -> Motif2 independent of CME status\n")

# ---- 2. Store key values for export ----
cme_pos_m2_n   <- as.integer(tab_cme_m2["CME-positive", "TRUE"])
cme_neg_m2_n   <- as.integer(tab_cme_m2["CME-negative", "TRUE"])
cme_pos_total  <- sum(tab_cme_m2["CME-positive", ])
cme_neg_total  <- sum(tab_cme_m2["CME-negative", ])
cme_pos_m2_pct <- round(100 * cme_pos_m2_n / cme_pos_total, 1)
cme_neg_m2_pct <- round(100 * cme_neg_m2_n / cme_neg_total, 1)
cme_m2_OR      <- round(ft_cme_m2$estimate, 2)
cme_m2_pval    <- ft_cme_m2$p.value

# ---- 3. Figure 3.17 ----
plot_df_cme <- data.frame(
  CME_status = c("CME-positive", "CME-negative"),
  n_total    = c(cme_pos_total,  cme_neg_total),
  n_motif2   = c(cme_pos_m2_n,  cme_neg_m2_n),
  pct_motif2 = c(cme_pos_m2_pct, cme_neg_m2_pct),
  col        = unname(pal_binary("CME-positive", "CME-negative"))
)
plot_df_cme$label <- paste0(plot_df_cme$pct_motif2, "%\n(",
                             plot_df_cme$n_motif2, "/",
                             plot_df_cme$n_total, ")")

pval_fmt_cme <- if (cme_m2_pval < 0.001) {
  parts <- strsplit(formatC(cme_m2_pval, format = "e", digits = 2), "e")[[1]]
  paste0("italic(p)==", parts[1], " %*% 10^{", as.integer(parts[2]), "}")
} else {
  paste0("italic(p)==", round(cme_m2_pval, 3))
}

p316 <- (function() {
  plot_df_cme$Count <- sprintf("(%d/%d)", plot_df_cme$n_motif2, plot_df_cme$n_total)
  fills316 <- pal_binary("CME-positive", "CME-negative")
  thesis_bar(plot_df_cme, x = "CME_status", y = "pct_motif2", fill = "CME_status",
             value_fmt = "%.1f%%", count = "Count", fills = fills316,
             y_lab = "Peaks containing Motif2 (%)", global_p = cme_m2_pval)
})()
ggsave(file.path(FIGURES_DIR, "fig3_17_cme_motif2_enrichment.png"), p316,
       width = 5, height = 4.3, dpi = 300, bg = "white")
cat("Saved: fig3_17_cme_motif2_enrichment.png\n")
```


## Motif2 across Trh-unique, Tgo-unique, joint peaks and genome-wide enhancers


```{r trh-tgo-unique-fasta, include=TRUE, eval=FALSE}
# ============================================================
# Trh-unique and Tgo-unique peak sets (IDR peaks with no overlap from partner factor, Tgo checked against Sim, Sima, Ss, Dys); export FASTA for FIMO.
# ============================================================
library(BSgenome.Dmelanogaster.UCSC.dm6)
genome <- BSgenome.Dmelanogaster.UCSC.dm6

# Trh_clean and Tgo_clean from setup chunk; joint_peaks_filt from Section 3.4
trh_unique_peaks <- Trh_clean[!Trh_clean %over% joint_peaks_filt]
tgo_unique_peaks <- Tgo_clean[!Tgo_clean %over% joint_peaks_filt]

cat("Trh-unique peaks (no Tgo overlap):", length(trh_unique_peaks), "\n")
cat("Tgo-unique peaks (no Trh overlap):", length(tgo_unique_peaks), "\n")
cat("Trh/Tgo joint peaks (reference):  ", length(joint_peaks_filt), "\n")

# Sanity check: a 15-bp PWM cannot be found in an interval narrower than 15 bp.
cat("Min width  -- Trh-unique:", min(width(trh_unique_peaks)),
    " Tgo-unique:", min(width(tgo_unique_peaks)),
    " joint:", min(width(joint_peaks_filt)), "\n")

# Export FASTA, unique sequence names (peak_N); name style is irrelevant to
# the prevalence loader as long as names are unique within each file.
seqs_trh_unique <- getSeq(genome, trh_unique_peaks)
names(seqs_trh_unique) <- paste0("peak_", seq_along(seqs_trh_unique))
writeXStringSet(seqs_trh_unique, file.path(EXPORT_DIR, "trh_unique.fa"))

seqs_tgo_unique <- getSeq(genome, tgo_unique_peaks)
names(seqs_tgo_unique) <- paste0("peak_", seq_along(seqs_tgo_unique))
writeXStringSet(seqs_tgo_unique, file.path(EXPORT_DIR, "tgo_unique.fa"))

cat("\nWrote trh_unique.fa and tgo_unique.fa to EXPORT_DIR.\n")
```


```{r joint-fasta-export, include=TRUE, eval=FALSE}
# ============================================================
# Export the Trh/Tgo joint peak set as joint.fa for the FIMO loop.
#
# The joint peaks must be scanned with the same FIMO parameters as the three
# other sets, so joint.fa is regenerated here rather than reusing any earlier
# joint FASTA built for a different purpose (e.g. MEME-ChIP discovery input).
# ============================================================
library(BSgenome.Dmelanogaster.UCSC.dm6)
genome <- BSgenome.Dmelanogaster.UCSC.dm6

cat("Min width -- joint_peaks_filt:", min(width(joint_peaks_filt)), "\n")

seqs_joint <- getSeq(genome, joint_peaks_filt)
names(seqs_joint) <- paste0("peak_", seq_along(seqs_joint))
writeXStringSet(seqs_joint, file.path(EXPORT_DIR, "joint.fa"))
cat("Wrote", length(seqs_joint), "joint peak sequences to joint.fa\n")
```


```{r enhancer-fasta-export, include=TRUE, eval=FALSE}
# ============================================================
# Build genome-wide annotated-enhancer set, export enhancers_genomewide.fa for full-PWM FIMO scan; keep enhancers >=15 bp (Motif2 PWM width).
# ============================================================
library(Biostrings)
library(GenomicRanges)
library(BSgenome.Dmelanogaster.UCSC.dm6)
genome <- BSgenome.Dmelanogaster.UCSC.dm6

# Load the annotated-enhancer BED if not already in the session.
if (!exists("enhancers_filt")) {
  enhancers_raw <- read.table(
    file.path(REFERENCE_LISTS, "merged_embryo_5-20_enhancers_dm6.bed"),
    header = FALSE, stringsAsFactors = FALSE
  )
  colnames(enhancers_raw) <- c("seqnames", "start", "end")
  enhancers_gr <- GRanges(
    seqnames = enhancers_raw$seqnames,
    ranges   = IRanges(start = enhancers_raw$start, end = enhancers_raw$end)
  )
  standard_chrs <- c("chr2L", "chr2R", "chr3L", "chr3R", "chr4", "chrX")
  enhancers_filt <- keepSeqlevels(enhancers_gr, standard_chrs,
                                  pruning.mode = "coarse")
}

# Full-PWM enhancer set: width >= 15 (PWM length).
enhancers_filt_pwm <- enhancers_filt[width(enhancers_filt) >= 15]
cat("Enhancers retained for 15-bp PWM scan (width >= 15):",
    length(enhancers_filt_pwm), "of", length(enhancers_filt), "\n")

enh_seqs_export <- getSeq(genome, enhancers_filt_pwm)
names(enh_seqs_export) <- paste0(
  as.character(seqnames(enhancers_filt_pwm)), ":",
  start(enhancers_filt_pwm), "-", end(enhancers_filt_pwm)
)
writeXStringSet(enh_seqs_export,
                file.path(EXPORT_DIR, "enhancers_genomewide.fa"))
cat("Wrote", length(enh_seqs_export),
    "enhancer sequences to enhancers_genomewide.fa\n")
```


```{bash fimo-four-way-scan, include=TRUE, eval=FALSE}
# ============================================================
# FIMO scan of four sets with PWM STRCACTGARARAAA.meme, --thresh 1.0E-4, --bgfile --nrdb--, run on the HPC.
# Matches footer of original Trh/Tgo joint-peak fimo.tsv; feeds Figure 3.17 bars.
# Requires in EXPORT_DIR (data/chapter3): STRCACTGARARAAA.meme, joint.fa, trh_unique.fa, tgo_unique.fa, enhancers_genomewide.fa
# ============================================================
cd $PROJECT_DIR/data/chapter3

for SET in joint trh_unique tgo_unique enhancers_genomewide; do
  conda run -n phd fimo \
    --oc fimo_${SET} \
    --verbosity 1 \
    --bgfile --nrdb-- \
    --thresh 1.0E-4 \
    STRCACTGARARAAA.meme \
    ${SET}.fa
  echo "Done: fimo_${SET}/fimo.tsv"
done
```


```{r motif2-fimo-prevalence}
# ============================================================
# Motif2 prevalence loader.
# FIMO collapses headers to chromosome names in sequence_name and reports hits in absolute genome coordinates, so unique(sequence_name) counts chromosomes; enhancer set is counted by genomic overlap instead.
# q-value is corrected per scan, not comparable between the 614-peak scan and the 35798-enhancer scan; cross-set comparison uses p-value < 1e-4.
# ============================================================
library(GenomicRanges)

PVAL_THRESH <- 1e-4   # matches the FIMO --thresh used for all four scans

# --- Loader A: peak sets scanned with peak_N FASTA headers -------------------
# sequence_name == "peak_<i>"; one row per hit; count distinct peaks hit.
motif2_prev_peakindexed <- function(fimo_tsv, n_total, label) {
  stopifnot(file.exists(fimo_tsv))
  f <- read.table(fimo_tsv, header = TRUE, sep = "\t", comment.char = "#")
  # FIMO writes the p-value column as "p.value" after read.table name-munging
  f <- f[f$p.value < PVAL_THRESH, ]
  n_pos <- length(unique(f$sequence_name))
  pct   <- round(100 * n_pos / n_total, 1)
  cat(label, ": Motif2 in", n_pos, "/", n_total,
      paste0("(", pct, "%)\n"))
  list(n_pos = n_pos, n_total = n_total, pct = pct)
}

# --- Loader B: enhancer set scanned with genomic FASTA headers ---------------
# sequence_name == chromosome; start/stop are ABSOLUTE genome coordinates.
# Build a GRanges of hits and count target intervals with >=1 overlap.
motif2_prev_genomic <- function(fimo_tsv, target_gr, label) {
  stopifnot(file.exists(fimo_tsv))
  f <- read.table(fimo_tsv, header = TRUE, sep = "\t", comment.char = "#")
  f <- f[f$p.value < PVAL_THRESH, ]
  hits_gr <- GRanges(
    seqnames = f$sequence_name,
    ranges   = IRanges(start = f$start, end = f$stop),
    strand   = f$strand
  )
  hit_target <- countOverlaps(target_gr, hits_gr, ignore.strand = TRUE) > 0
  n_pos <- sum(hit_target)
  n_tot <- length(target_gr)
  pct   <- round(100 * n_pos / n_tot, 1)
  cat(label, ": Motif2 in", n_pos, "/", n_tot,
      paste0("(", pct, "%)\n"))
  list(n_pos = n_pos, n_total = n_tot, pct = pct)
}

# --- Four scans --------------------------------------------------------------
# confirm the joint / unique fimo.tsv files use peak_N naming.
# tail their footers; if any used genomic headers, switch it to loader B.
joint_m2 <- motif2_prev_peakindexed(
  file.path(EXPORT_DIR, "fimo_joint", "fimo.tsv"),
  length(joint_peaks_filt), "Trh/Tgo joint")

trh_u_m2 <- motif2_prev_peakindexed(
  file.path(EXPORT_DIR, "fimo_trh_unique", "fimo.tsv"),
  length(trh_unique_peaks), "Trh-unique")

tgo_u_m2 <- motif2_prev_peakindexed(
  file.path(EXPORT_DIR, "fimo_tgo_unique", "fimo.tsv"),
  length(tgo_unique_peaks), "Tgo-unique")

enh_m2 <- motif2_prev_genomic(
  file.path(EXPORT_DIR, "fimo_enhancers_genomewide", "fimo.tsv"),
  enhancers_filt_pwm, "Genome-wide enhancers")

# --- Variables consumed by Figure 3.17 and the computed-values export --------
joint_m2_n         <- joint_m2$n_pos
joint_m2_total     <- joint_m2$n_total
joint_m2_pct       <- joint_m2$pct

trh_unique_m2_n    <- trh_u_m2$n_pos
trh_unique_n_total <- trh_u_m2$n_total
trh_unique_m2_pct  <- trh_u_m2$pct

tgo_unique_m2_n    <- tgo_u_m2$n_pos
tgo_unique_n_total <- tgo_u_m2$n_total
tgo_unique_m2_pct  <- tgo_u_m2$pct

mspec_n_enh_m2_pos <- enh_m2$n_pos
mspec_n_enh_total  <- enh_m2$n_total
mspec_enh_pct      <- enh_m2$pct

# --- Fisher tests: each unique set vs joint reference ------------------------
ft_trh_unique <- fisher.test(matrix(
  c(joint_m2_n,      joint_m2_total - joint_m2_n,
    trh_unique_m2_n, trh_unique_n_total - trh_unique_m2_n), nrow = 2))
ft_tgo_unique <- fisher.test(matrix(
  c(joint_m2_n,      joint_m2_total - joint_m2_n,
    tgo_unique_m2_n, tgo_unique_n_total - tgo_unique_m2_n), nrow = 2))

cat("\nFisher (joint vs Trh-unique):  OR =", round(ft_trh_unique$estimate, 2),
    " p =", formatC(ft_trh_unique$p.value, format = "e", digits = 2), "\n")
cat("Fisher (joint vs Tgo-unique):  OR =", round(ft_tgo_unique$estimate, 2),
    " p =", formatC(ft_tgo_unique$p.value, format = "e", digits = 2), "\n")

ft_trh_OR <- round(ft_trh_unique$estimate, 2); ft_trh_p <- ft_trh_unique$p.value
ft_tgo_OR <- round(ft_tgo_unique$estimate, 2); ft_tgo_p <- ft_tgo_unique$p.value

motif2_summary <- data.frame(
  Peak_set   = c("Trh/Tgo joint (reference)", "Trh-unique (no Tgo)",
                 "Tgo-unique (no Trh)", "Genome-wide enhancers"),
  N_total    = c(joint_m2_total, trh_unique_n_total,
                 tgo_unique_n_total, mspec_n_enh_total),
  N_motif2   = c(joint_m2_n, trh_unique_m2_n,
                 tgo_unique_m2_n, mspec_n_enh_m2_pos),
  Pct_motif2 = c(joint_m2_pct, trh_unique_m2_pct,
                 tgo_unique_m2_pct, mspec_enh_pct),
  stringsAsFactors = FALSE)
cat("\n=== Motif2 prevalence (full-PWM FIMO, p < 1e-4) ===\n")
print(motif2_summary)

# --- Fisher test: joint peaks vs genome-wide enhancer baseline ---------------
# This is the tier-1 test: is Motif2 enriched at bHLH-PAS binding sites
# relative to the genome-wide annotated-enhancer average?
ft_joint_vs_enh <- fisher.test(matrix(
  c(joint_m2_n,                  joint_m2_total - joint_m2_n,
    mspec_n_enh_m2_pos,          mspec_n_enh_total - mspec_n_enh_m2_pos),
  nrow = 2))

cat("\nFisher (joint vs genome-wide enhancers):  OR =",
    round(ft_joint_vs_enh$estimate, 2),
    " p =", formatC(ft_joint_vs_enh$p.value, format = "e", digits = 2), "\n")

mspec_peaks_vs_enh_or <- round(ft_joint_vs_enh$estimate, 2)
mspec_peaks_vs_enh_p  <- ft_joint_vs_enh$p.value
```

```{r length-check}
# --- Width sanity check: PWM is 15 bp; any interval < 15 bp cannot hit -------
check_widths <- function(gr, label) {
  w <- width(gr)
  n_short <- sum(w < 15)
  cat(sprintf("%-22s n=%6d  min=%5d  median=%6.0f  <15bp: %d\n",
              label, length(gr), min(w), median(w), n_short))
  if (n_short > 0) {
    cat("  WARNING:", n_short, "intervals <15bp in", label,
        "-- these are forced negatives; consider a width>=15 filter.\n")
  }
  invisible(n_short)
}
check_widths(joint_peaks_filt,   "Trh/Tgo joint")
check_widths(trh_unique_peaks,   "Trh-unique")
check_widths(tgo_unique_peaks,   "Tgo-unique")
check_widths(enhancers_filt_pwm, "Enhancers (PWM set)")
```

```{r fig3-17-four-bar, include=TRUE, eval=FALSE}
# ============================================================
# Figure 3.17: Motif2 prevalence in Trh-unique, Tgo-unique, joint peaks and genome-wide enhancer baseline; same FIMO scan (q<0.05). Requires objects from `motif2-fimo-prevalence`
library(ggplot2)

stopifnot(exists("trh_unique_m2_pct"), exists("tgo_unique_m2_pct"),
          exists("joint_m2_pct"),      exists("mspec_enh_pct"))

fig_df <- data.frame(
  Peak_set = factor(
    c("Trh-unique\n(no Tgo)",
      "Trh/Tgo joint\n(reference)",
      "Tgo-unique\n(no Trh)",
      "Genome-wide\nenhancers"),
    levels = c("Trh-unique\n(no Tgo)",
               "Trh/Tgo joint\n(reference)",
               "Tgo-unique\n(no Trh)",
               "Genome-wide\nenhancers")
  ),
  Pct_motif2 = c(trh_unique_m2_pct, joint_m2_pct,
                 tgo_unique_m2_pct, mspec_enh_pct),
  N_motif2   = c(trh_unique_m2_n,  joint_m2_n,
                 tgo_unique_m2_n,  mspec_n_enh_m2_pos),
  N_total    = c(trh_unique_n_total, joint_m2_total,
                 tgo_unique_n_total, mspec_n_enh_total),
  # Occupancy classes inherit their defining subunit's colour; the genome-wide
  # set stays neutral grey because it is a reference, not a peak class.
  Fill = unname(PAL_OCCUPANCY[c("alpha_only", "joint", "beta_only", "genome_ref")])
)
fig_df$Label <- paste0(fig_df$Pct_motif2, "%\n(",
                       fig_df$N_motif2, "/", fig_df$N_total, ")")

p317 <- (function() {
  fig_df$Count <- sprintf("(%d/%d)", fig_df$N_motif2, fig_df$N_total)
  fills317 <- fig_df$Fill; names(fills317) <- as.character(fig_df$Peak_set)
  thesis_bar(fig_df, x = "Peak_set", y = "Pct_motif2", fill = "Peak_set",
             value_fmt = "%.1f%%", count = "Count",
             fills = fills317, dashed_at = mspec_enh_pct,
             y_lab = "Sequences containing Motif2 (full PWM, %)")
})()
ggsave(file.path(FIGURES_DIR, "fig3_18_trh_tgo_unique_motif2.png"), p317,
       width = 7, height = 4.3, dpi = 300, bg = "white")
cat("Saved: fig3_18_trh_tgo_unique_motif2.png (four bars, full-PWM FIMO)\n")
```



```{r, include=FALSE, eval=FALSE}
# ====
# Exports Chapter 3 reported values to results/reported_values.tsv; run last, after analyses.
# ====

# ---- Helper: format p-values for scientific notation ----
fmt_p <- function(p, digits = 2) {
  formatted <- formatC(p, format = "e", digits = digits)
  parts <- strsplit(formatted, "e")[[1]]
  mantissa <- parts[1]
  exponent <- as.integer(parts[2])
  paste0(mantissa, "e", exponent)
}

fmt_sci <- function(x, digits = 2) {
  # Wrapper for already-known values passed as strings (no recomputation needed)
  x
}

# ============================================================
# SECTION 1, Peak counts (from IDR-filtered narrowPeak files)
# Objects: Trh_clean, Tgo_clean, joint_peaks_filt
# ============================================================
trh_peaks_n  <- length(Trh_clean)           # 1033
tgo_peaks_n  <- length(Tgo_clean)           # 1403
joint_n      <- length(joint_peaks_filt)    # 614

# Percentages for Trh and Tgo participating in joint set.
# joint_n counts pintersect intervals, not peaks; dividing by tgo_peaks_n overstates Tgo share (tgo_unique.fa: 1403 - 609 = 794).
# Count distinct peaks per side.
hits_tt        <- findOverlaps(Trh_clean, Tgo_clean, minoverlap = 1L)
trh_joint_n    <- length(unique(queryHits(hits_tt)))     # 614
tgo_joint_n    <- length(unique(subjectHits(hits_tt)))   # 609
tgo_spanning_n <- sum(table(subjectHits(hits_tt)) > 1L)  # 5

stopifnot(length(hits_tt) == joint_n,          # pairs == intersection intervals
          all(table(queryHits(hits_tt)) == 1L),# every Trh peak has one partner
          joint_n - tgo_joint_n == tgo_spanning_n)

trh_pct_joint <- round(100 * trh_joint_n / trh_peaks_n, 1)   # 59.4
tgo_pct_joint <- round(100 * tgo_joint_n / tgo_peaks_n, 1)   # 43.4

# Trh-unique peaks (Trh peaks not in joint set)
trh_unique_n  <- trh_peaks_n - joint_n                    # 419

# ============================================================
# SECTION 2, Permutation test: Trh vs Tgo overlap (peakPermTest)
# Object: pt2_ (from Figure 3.6B chunk)
# ============================================================
perm_obs_overlap  <- pt2_$cntOverlaps$observed            # 614
perm_zscore       <- round(pt2_$cntOverlaps$zscore, 2)    # 114.58
perm_pval_raw     <- pt2_$cntOverlaps$pval                # 2.00e-4 (= floor 1/5001)
perm_n            <- length(pt2_$cntOverlaps$permuted)    # 5000
perm_ev95         <- round(quantile(pt2_$cntOverlaps$permuted, 0.95))  # 35
perm_fold         <- round(perm_obs_overlap / mean(pt2_$cntOverlaps$permuted), 0)  # ~23

# ============================================================
# SECTION 3, Enhancer overlap counts (from Figure 3.7B chunk)
# Objects: pt_all, pt_promoter, pt_distal
# ============================================================
enh_obs_all       <- pt_all$numOverlaps$observed           # 169
enh_obs_promoter  <- pt_promoter$numOverlaps$observed      # 55
enh_obs_distal    <- pt_distal$numOverlaps$observed        # 114

enh_null_mean_all <- round(mean(pt_all$numOverlaps$permuted),      2)  # 202.13
enh_null_sd_all   <- round(sd(pt_all$numOverlaps$permuted),        2)  # 11.19
enh_z_all         <- round(pt_all$numOverlaps$zscore,              2)  # -2.96
enh_p_all         <- round(pt_all$numOverlaps$pval, 5)             # 0.0026

enh_null_mean_prom <- round(mean(pt_promoter$numOverlaps$permuted), 2) # 154.46
enh_null_sd_prom   <- round(sd(pt_promoter$numOverlaps$permuted),   2) # 9.99
enh_z_prom         <- round(pt_promoter$numOverlaps$zscore,         2) # -9.95
enh_p_prom         <- round(pt_promoter$numOverlaps$pval,5)                     # 2e-04

enh_null_mean_dist <- round(mean(pt_distal$numOverlaps$permuted),   2) # 46.79
enh_null_sd_dist   <- round(sd(pt_distal$numOverlaps$permuted),     2) # 5.77
enh_z_dist         <- round(pt_distal$numOverlaps$zscore,           2) # 11.64
enh_p_dist         <- round(pt_distal$numOverlaps$pval,5)                       # 2e-04

# ============================================================
# SECTION 4, Reference gene set
# Object: ts_sg_genes, trach_entrez, comp_entrez
# ============================================================
ts_sg_total      <- length(ts_sg_genes)                   # 2053
ts_sg_entrez_n   <- length(trach_entrez)                  # 2049
ts_sg_unmapped_n <- ts_sg_total - ts_sg_entrez_n          # 4

# comp_entrez is a live setdiff() against whatever org.Dm.eg.db is installed
# in the session, so length(comp_entrez) drifts with the annotation package
# and need not match the geneset file that ChIP-Enrich actually read. The
# size is therefore read off the file that was passed to chipenrich(), which
# is the only number the reported ORs correspond to.
.gs_written   <- read.delim(file.path(EXPORT_DIR, "geneset_tracheal_vs_complement.tsv"),
                            header = FALSE, stringsAsFactors = FALSE)
.gs_tab       <- table(.gs_written$V1)
comp_entrez_n <- unname(.gs_tab[["All_dm6_minus_tracheal"]])   # 23030
stopifnot(unname(.gs_tab[["tracheal_custom"]]) == ts_sg_entrez_n)

# ChIP-Enrich further restricts each geneset to genes it can place under
# its own locus definition. Those are the denominators the reported
# peak-gene counts (peak genes, tracheal and comparison) belong to.
.ce_custom <- read.csv(file.path(EXPORT_DIR, "chipenrich_ch3_custom.csv"),
                       stringsAsFactors = FALSE)
rownames(.ce_custom)  <- .ce_custom$Geneset.ID
ce_trach_modelled_n   <- .ce_custom["tracheal_custom",        "N.Geneset.Genes"]  # 1986
ce_comp_modelled_n    <- .ce_custom["All_dm6_minus_tracheal", "N.Geneset.Genes"]  # 11495

# ============================================================
# SECTION 5, Hypergeometric & permutation: peak-proximal TS/SG genes
# Requires: N, A, B, k, p_hyper, fold_enr from Fig 3.9 chunk
#           pt from permutation chunk
# ============================================================

nearby_genes_n     <- A
overlap_n          <- k                          # 114
hyper_fold         <- round(fold_enr, 2)         # 2.01
hyper_p_val        <- p_hyper                    # 4.39e-13

# Permutation test
perm_obs_tsg       <- pt$perm_eval$observed      # 114
perm_null_mean_tsg <- round(mean(pt$perm_eval$permuted), 2)
perm_null_sd_tsg   <- round(sd(pt$perm_eval$permuted),   2)
perm_z_tsg         <- round(
  (pt$perm_eval$observed - mean(pt$perm_eval$permuted)) /
    sd(pt$perm_eval$permuted), 2)
perm_n_tsg         <- length(pt$perm_eval$permuted)

# Empirical p-value, +1 continuity correction prevents exact zero
perm_p_tsg_raw <- (sum(pt$perm_eval$permuted >= pt$perm_eval$observed) + 1) /
                  (perm_n_tsg + 1)

# Format both p-values identically in scientific notation
fmt_p_sci <- function(p) {
  s        <- formatC(p, format = "e", digits = 2)
  parts    <- strsplit(s, "e")[[1]]
  mantissa <- parts[1]
  exponent <- as.integer(parts[2])
  paste0(mantissa, "e", exponent)
}

hyper_p_fmt    <- fmt_p_sci(hyper_p_val)     # 4.39e-13

perm_p_tsg_fmt <- if (perm_p_tsg_raw <= (1 / perm_n_tsg + 1e-10)) {
  # Floor hit, no permuted value reached observed; report as bound
  bound <- formatC(1 / perm_n_tsg, format = "e", digits = 2)
  parts    <- strsplit(bound, "e")[[1]]
  mantissa <- parts[1]
  exponent <- as.integer(parts[2])
  paste0("< ", mantissa, "e", exponent)
  # e.g. "< 2.00e-4"
} else {
  fmt_p_sci(perm_p_tsg_raw)
}
# ============================================================
# SECTION 6, Gene length bias test
# Objects: ts_sg_lengths, other_lengths, wilcox test
# ============================================================
median_tsg_length   <- median(ts_sg_lengths)               # 4685
median_other_length <- median(other_lengths)               # 1512
length_fold         <- round(median_tsg_length / median_other_length, 1)  # 3.1
wtest               <- wilcox.test(ts_sg_lengths, other_lengths)
wilcox_W            <- wtest$statistic                     # 24279486
# p < 2.2e-16, reported as-is

# ============================================================
# SECTION 7, ChIP-Enrich (Table 3.2)
# Object: res_custom (from chipenrich chunk)
# ============================================================
trach_row   <- res_custom[res_custom$Description == "tracheal_custom", ]
comp_row    <- res_custom[res_custom$Description == "All_dm6_minus_tracheal", ]

ce_OR_trach   <- round(trach_row$Odds.Ratio, 2)            # 1.56
ce_p_trach    <- trach_row$P.value                         # 1.82e-4
ce_OR_comp    <- round(comp_row$Odds.Ratio, 2)             # 0.64
ce_peakgenes_trach <- trach_row$N.Geneset.Peak.Genes       # 107
ce_peakgenes_comp  <- comp_row$N.Geneset.Peak.Genes        # 319

trach_row_fet <- res_custom_fet[res_custom_fet$Description == "tracheal_custom", ]
ce_OR_fisher <- round(trach_row_fet$Odds.Ratio, 2)    # 1.99
ce_p_fisher  <- trach_row_fet$P.value                  # 1.26e-8
# ============================================================
# SECTION 8, Regulatory context classification (Figure 3.11)
# Object: summary_df (from fig3_12 chunk)
# ============================================================
# These are computed inside bar_df_35 / df_plot in the fig3_12 chunk
# Recompute from joint_peaks_filt$Regulatory_Class for safety
rc_counts <- table(joint_peaks_filt$Regulatory_Class)

n_prom_nonanno   <- as.integer(rc_counts["Promoter-proximal peak (non-annotated)"])   # 413
n_prom_enh       <- as.integer(rc_counts["Promoter-proximal enhancer (annotated)"])   # 55
n_dist_enh       <- as.integer(rc_counts["Distal enhancer (annotated)"])              # 114
n_dist_nonanno   <- as.integer(rc_counts["Distal peak (non-annotated)"])              # 32

pct_prom_nonanno <- round(100 * n_prom_nonanno / joint_n, 1)  # 67.3
pct_prom_enh     <- round(100 * n_prom_enh / joint_n, 1)      # 9
pct_dist_enh     <- round(100 * n_dist_enh / joint_n, 1)      # 18.6
pct_dist_nonanno <- round(100 * n_dist_nonanno / joint_n, 1)  # 5.2
pct_promoter_total <- round(pct_prom_nonanno + pct_prom_enh, 1) # 76.3

# ============================================================
# SECTION 9, CME analysis (Figure 3.12)
# Object: joint_peaks_filt$CME
# ============================================================
n_cme_pos  <- sum(joint_peaks_filt$CME)           # 186 (both-strand ACGTG)
n_cme_neg  <- sum(!joint_peaks_filt$CME)           # 428
pct_cme_pos <- round(100 * n_cme_pos / joint_n, 1) # 30.3
pct_cme_neg <- round(100 * n_cme_neg / joint_n, 1) # 69.7

cme_wilcox_p <- wilcox.test(signalValue ~ CME,
                             data = as.data.frame(joint_peaks_filt))$p.value  # 0.328

# ============================================================
# SECTION 10, Distal vs promoter ChIP signal (Figure 3.13)
# Object: df_signal
# ============================================================
n_promoter_signal  <- sum(df_signal$Context == "Promoter")                     # 468
n_distal_enh_signal <- sum(df_signal$Context == "Distal enhancer (annotated)") # 114
signal_wilcox_p    <- wilcox.test(signalValue ~ Context, data = df_signal)$p.value  # 7.40e-6

# ============================================================
# SECTION 11, Motif2 contextual enrichment (Figure 3.14)
# Object: tab_refined, joint_peaks_filt$Motif2, Regulatory_Class
# ============================================================
n_motif2_total        <- sum(joint_peaks_filt$Motif2)             # 178
pct_motif2_total      <- round(100 * n_motif2_total / joint_n, 1)  # 29

n_motif2_dist_enh     <- tab_refined["Distal enhancer (annotated)", "TRUE"]          # 60
n_motif2_prom_enh     <- tab_refined["Promoter-proximal enhancer (annotated)", "TRUE"] # 20
n_motif2_prom_nonanno <- tab_refined["Promoter-proximal peak (non-annotated)", "TRUE"] # 86
n_motif2_dist_nonanno <- tab_refined["Distal peak (non-annotated)", "TRUE"]            # 12

pct_motif2_dist_enh     <- round(100 * n_motif2_dist_enh / n_dist_enh, 1)       # 52.6
pct_motif2_prom_enh     <- round(100 * n_motif2_prom_enh / n_prom_enh, 1)       # 36.4
pct_motif2_prom_nonanno <- round(100 * n_motif2_prom_nonanno / n_prom_nonanno, 1) # 20.8
pct_motif2_dist_nonanno <- round(100 * n_motif2_dist_nonanno / n_dist_nonanno, 1) # 37.5
motif2_fisher_p <- fisher.test(tab_refined)$p.value                              # 6.86e-10
# Motif2 fold-enrichment comparisons (for Discussion 3.7.5).
#
# Fold-enrichments are computed separately for the distal-enhancer class
# and the promoter-enhancer class, using the counts defined just above,
# so that the two classes are not pooled into a single claim.
.f <- function(a, na, b, nb, dp = 1) round((a/na) / (b/nb), dp)
motif2_fold_enh_vs_prom <- .f(n_motif2_dist_enh, n_dist_enh,
                              n_motif2_prom_nonanno, n_prom_nonanno)    #2.5
motif2_fold_enh_vs_dist <- .f(n_motif2_dist_enh, n_dist_enh,
                              n_motif2_dist_nonanno, n_dist_nonanno)    #1.4
motif2_fold_pe_vs_pn    <- .f(n_motif2_prom_enh, n_prom_enh,
                              n_motif2_prom_nonanno, n_prom_nonanno)    #1.7
motif2_fold_pe_vs_dn    <- .f(n_motif2_prom_enh, n_prom_enh,
                              n_motif2_dist_nonanno, n_dist_nonanno, 2) #0.97
# High-confidence peak percentage (for §3.7.7)
pct_high_conf <- round(100 * (n_dist_enh + n_prom_enh) / joint_n, 1) #27.5
# ============================================================
# SECTION 12, CME-motif2 inter-motif spacing (Figure 3.15)
# Object: spacings, peaks_both
# ============================================================
n_peaks_both         <- length(peaks_both)            # 55
spacing_median       <- median(spacings)              # 72
spacing_min          <- min(spacings)                 # 2
spacing_max          <- max(spacings)                 # 275
n_close_apposition   <- sum(spacings <= 50)           # 17

# ============================================================
# SECTION 13, Per-chromosome table (Table 3.1), these are computed not hardcoded
# Object: chr_distribution
# ============================================================

# ============================================================
# RECORD computed values in results/reported_values.tsv
# ============================================================

# ---- Distal/proximal GO diagnostic values for results/reported_values.tsv ----
# Objects come from the load-chipenrich-ch3-distalsplit loader chunk.
stopifnot(exists("res_distal_bp_sig"), exists("res_prox_bp_sig"),
          exists("n_distal_ds"), exists("n_promoter_ds"))

# Top distal term (ordered by FDR ascending in the saved object)
.dtop <- res_distal_bp_sig[1, ]
dist_go_top_term <- as.character(.dtop$Description)            # e.g. "embryo development"
dist_go_top_or   <- formatC(.dtop$Odds.Ratio, format = "f", digits = 2)
dist_go_top_fdr  <- .dtop$FDR                                  # fmt_p() applied at write time

# --- Motif2 functional-test scalars (TEST1 regression, TEST3 phyloP) --------
AUX_CV <- file.path(PROJECT_DIR, "data/chapter3")
t1 <- readRDS(file.path(AUX_CV, "test1_regression_binding.rds"))
t3 <- readRDS(file.path(AUX_CV, "test3_phylop_conservation.rds"))
if (!requireNamespace("sandwich", quietly = TRUE)) install.packages("sandwich")
if (!requireNamespace("lmtest",   quietly = TRUE)) install.packages("lmtest")

t1f <- summary(t1$full)$coefficients
t1b <- summary(t1$binary)$coefficients
m2_beta     <- round(t1f["n_motif2", "Estimate"], 2)
m2_beta_ci  <- sprintf("%.2f", t1f["n_motif2", "Estimate"] + c(-1.96, 1.96) * t1f["n_motif2", "Std. Error"])
m2_outcome_median <- sprintf("%.2f", median(t1$df$peak_score))
m2_reg_p    <- t1f["n_motif2", "Pr(>|t|)"]
m2_reg_p_hc <- lmtest::coeftest(t1$full,
                 vcov. = sandwich::vcovHC(t1$full, type = "HC3"))["n_motif2", "Pr(>|t|)"]
m2_bin_beta <- round(t1b["has_motif2", "Estimate"], 2)
m2_bin_p    <- t1b["has_motif2", "Pr(>|t|)"]
adjr_red    <- round(summary(t1$reduced)$adj.r.squared, 3)
adjr_full   <- round(summary(t1$full)$adj.r.squared, 3)
ppk <- t3$full$table; ppc <- t3$core$table
m2_pp_n      <- nrow(ppk)
flank_w      <- t3$flank_w
m2_pp_motif  <- round(median(ppk$motif), 2)
m2_pp_flank  <- round(median(ppk$flank), 2)
m2_pp_diff   <- round(median(ppk$motif - ppk$flank), 2)
m2_pp_propup <- round(100 * mean((ppk$motif - ppk$flank) > 0), 1)
m2_pp_p      <- t3$full$wilcox$p.value
core_med     <- round(median(ppc$core), 2)
core_diff    <- round(median(ppc$core - ppc$flank), 2)
core_p       <- t3$core$wilcox$p.value
```

```{r ch3-motif2-dbscan, eval=FALSE}
# ============================================================
# TOMTOM scan of Motif2 PWM (data/chapter3/STRCACTGARARAAA.meme) against combined Fly Factor Survey + JASPAR2024 CORE insects database; result written to results/reported_values.tsv
# ============================================================

qry <- file.path(PROJECT_DIR, "data/chapter3/STRCACTGARARAAA.meme")
db  <- file.path(PROJECT_DIR, "reference/motif_databases/combined_drosophila.meme")
ffs <- file.path(PROJECT_DIR, "reference/motif_databases/FLY/fly_factor_survey.meme")
jas <- file.path(PROJECT_DIR,
        "reference/motif_databases/JASPAR/JASPAR2024_CORE_insects_non-redundant_v2.meme")
stopifnot(file.exists(qry), file.exists(ffs), file.exists(jas))

# Rebuild the combined database (same concatenation as the MEME-ChIP step).
if (!file.exists(db)) writeLines(c(readLines(ffs), readLines(jas)), db)
n_db <- sum(grepl("^MOTIF", readLines(db)))            # 942 matrices

# Run 1, canonical significance count. TOMTOM writes only matches <= thresh,
# so an empty table == no significant match.
oc_sig <- file.path(PROJECT_DIR, "data/chapter3/tomtom_motif2_vs_db_q05")
system2("tomtom", c("-no-ssc","-thresh","0.05","-min-overlap","5",
                    "-dist","pearson","-oc", oc_sig, qry, db))
tt_sig <- read.delim(file.path(oc_sig, "tomtom.tsv"),
                     comment.char = "#", check.names = FALSE)
tt_sig <- tt_sig[!is.na(tt_sig$Query_ID) & tt_sig$Query_ID != "", ]
n_hits <- nrow(tt_sig)                                 # expect 0

# Run 2, nearest neighbour (relaxed, E-value scored) for the appendix only.
oc_all <- file.path(PROJECT_DIR, "data/chapter3/tomtom_motif2_vs_db_all")
system2("tomtom", c("-no-ssc","-thresh","10","-evalue","-min-overlap","5",
                    "-dist","pearson","-oc", oc_all, qry, db))
tt_all <- read.delim(file.path(oc_all, "tomtom.tsv"),
                     comment.char = "#", check.names = FALSE)
tt_all <- tt_all[!is.na(tt_all$Query_ID) & tt_all$Query_ID != "", ]
tt_all <- tt_all[order(tt_all[["E-value"]]), ]

stopifnot(n_hits == 0L)     # CANONICAL FACT: Motif2 matches no catalogued TF

# ---- session objects consumed by the ch3-export chunk ----
motif2_db_nmatrices <- n_db
motif2_db_nhits     <- n_hits
motif2_db_thresh    <- "0.05"
motif2_db_nearestE  <- sprintf("%.1f", tt_all[["E-value"]][1])   # ~2.9

cat(sprintf("Motif2 vs %d TF matrices: %d matches at q<%s; nearest %s E=%.2f\n",
            n_db, n_hits, motif2_db_thresh,
            tt_all$Target_ID[1], tt_all[["E-value"]][1]))
```

```{r ch3-export-computed-values, eval=FALSE}
# ============================================================
# Exports chapter 3 reported values via save_values(); run last, after all analysis chunks. Excludes 66 chapter 5 and 20+10 chapter 4 values.
# ============================================================

stopifnot(exists("save_values"),
          length(joint_peaks_filt) == 614L) # canonical-fact check

# The single-coefficient t-test and the nested F-test of the Motif2 dosage
# regression are the same test on 1 df; assert the identity and a plausible
# range for the p-value before recording it.
local({
  .p  <- summary(t1$full)$coefficients["n_motif2", "Pr(>|t|)"]
  .av <- anova(t1$reduced, t1$full)
  stopifnot(abs(.p - .av$`Pr(>F)`[2]) < 1e-12,   # same test, 1 df
            .p > 1e-6, .p < 0.05)                # plausible for n = 614
})

# ---- fmt_p: safe re-use if already defined in session ----
if (!exists("fmt_p")) {
  fmt_p <- function(p, digits = 2) {
    if (is.na(p)) return("NA")
    if (p == 0)   return("< 1e-300")
    s <- formatC(p, format = "e", digits = digits)
    parts <- strsplit(s, "e")[[1]]
    paste0(parts[1], "e", as.integer(parts[2]))
  }
}

ch3_vals <- list(

  # ---- Section 1: Peak counts ----
  trhPeaks              = length(Trh_clean),           # 1033
  tgoPeaks              = length(Tgo_clean),           # 1403
  jointPeaks            = length(joint_peaks_filt),    # 614, canonical
  trhJointPeaks         = trh_joint_n,                 # 614 distinct Trh peaks
  tgoJointPeaks         = tgo_joint_n,                 # 609 distinct Tgo peaks
  tgoSpanningPeaks      = tgo_spanning_n,              # 5 Tgo peaks spanning two Trh peaks
  trhPctJoint           = trh_pct_joint,               # 59.4
  tgoPctJoint           = tgo_pct_joint,               # 43.4
  trhUniquePeaks        = trh_unique_n,                # 419

  # ---- Section 2: Trh/Tgo overlap permutation test ----
  permObsOverlap        = perm_obs_overlap,            # 614
  permZscore            = perm_zscore,                 # 114.58
  permPval              = fmt_p(perm_pval_raw),        # 2.00e-4
  permN                 = perm_n,                      # 5000
  permEv                = perm_ev95,                   # 35
  permFold              = perm_fold,                   # 23

  # ---- Section 3: Enhancer overlap permutation (regioneR) ----
  enhObsAll             = enh_obs_all,
  enhObsPromoter        = enh_obs_promoter,
  enhObsDistal          = enh_obs_distal,
  enhNullMeanAll        = enh_null_mean_all,
  enhNullSdAll          = enh_null_sd_all,
  enhZAll               = enh_z_all,
  enhPAll               = enh_p_all,
  enhNullMeanProm       = enh_null_mean_prom,
  enhNullSdProm         = enh_null_sd_prom,
  enhZProm              = enh_z_prom,
  enhPProm              = enh_p_prom,
  enhNullMeanDist       = enh_null_mean_dist,
  enhNullSdDist         = enh_null_sd_dist,
  enhZDist              = enh_z_dist,
  enhPDist              = enh_p_dist,

  # ---- Section 4: Reference gene set ----
  tsSgTotal             = ts_sg_total,
  tsSgEntrezN           = ts_sg_entrez_n,
  tsSgUnmapped          = ts_sg_unmapped_n,
  compEntrezN           = comp_entrez_n,           # 23030, read off the geneset file
  ceTrachModelledN      = ce_trach_modelled_n,     # 1986, after ChIP-Enrich locus definition
  ceCompModelledN       = ce_comp_modelled_n,      # 11495

  # ---- Section 5: Hypergeometric and permutation (TS/SG vs peaks) ----
  nearbyGenesN          = nearby_genes_n,
  overlapN              = overlap_n,
  hyperFold             = hyper_fold,
  hyperPval             = hyper_p_fmt,
  permObsTsg            = perm_obs_tsg,
  permNullMeanTsg       = perm_null_mean_tsg,
  permNullSdTsg         = perm_null_sd_tsg,
  permZTsg              = perm_z_tsg,
  permPTsg              = perm_p_tsg_fmt,
  permNTsg              = perm_n_tsg,

  # ---- Section 6: Gene length bias ----
  medianTsgLength       = median_tsg_length,
  medianOtherLength     = median_other_length,
  lengthFold            = length_fold,
  wilcoxW               = formatC(wilcox_W, format = "f", big.mark = "", digits = 0),

  # ---- Section 7: ChIP-Enrich (Table 3.2) ----
  ceORTrach             = ce_OR_trach,
  cePTrach              = fmt_p(ce_p_trach),
  ceORComp              = ce_OR_comp,
  cePeakGenesTrach      = ce_peakgenes_trach,
  cePeakGenesComp       = ce_peakgenes_comp,
  ceORfisher            = ce_OR_fisher,
  cePfisher             = fmt_p(ce_p_fisher),

  # ---- Section 8: Regulatory context (Figure 3.11) ----
  nPromNonAnno          = n_prom_nonanno,
  nPromEnh              = n_prom_enh,
  nDistEnh              = n_dist_enh,
  nDistNonAnno          = n_dist_nonanno,
  pctPromNonAnno        = pct_prom_nonanno,
  pctPromEnh            = pct_prom_enh,
  pctDistEnh            = pct_dist_enh,
  pctDistNonAnno        = pct_dist_nonanno,
  pctPromoterTotal      = pct_promoter_total,

  # ---- Section 9: CME analysis ----
  nCmePos               = n_cme_pos,
  nCmeNeg               = n_cme_neg,
  pctCmePos             = pct_cme_pos,
  pctCmeNeg             = pct_cme_neg,
  cmeWilcoxP            = round(cme_wilcox_p, 3),

  # ---- Section 10: Signal context comparison ----
  nPromoterSignal       = n_promoter_signal,
  nDistalEnhSignal      = n_distal_enh_signal,
  signalWilcoxP         = fmt_p(signal_wilcox_p),

  # ---- Section 11: Motif2 enrichment (distal vs promoter) ----
  nMotifThreeTotal      = n_motif2_total,
  pctMotifThreeTotal    = pct_motif2_total,
  nMotifThreeDistEnh    = n_motif2_dist_enh,
  nMotifThreePromEnh    = n_motif2_prom_enh,
  nMotifThreePromNon    = n_motif2_prom_nonanno,
  nMotifThreeDistNon    = n_motif2_dist_nonanno,
  pctMotifThreeDistEnh  = pct_motif2_dist_enh,
  pctMotifThreePromEnh  = pct_motif2_prom_enh,
  pctMotifThreePromNon  = pct_motif2_prom_nonanno,
  pctMotifThreeDistNon  = pct_motif2_dist_nonanno,
  motifThreeFisherP     = fmt_p(motif2_fisher_p),
  motifTwoFoldEnhVsProm = motif2_fold_enh_vs_prom,
  motifTwoFoldEnhVsDist = motif2_fold_enh_vs_dist,
  motifTwoFoldPromEnhVsPromNon = motif2_fold_pe_vs_pn,
  motifTwoFoldPromEnhVsDistNon = motif2_fold_pe_vs_dn,
  pctHighConfPeaks      = pct_high_conf,

  # ---- Section 12: CME-Motif2 spacing ----
  nPeaksBoth            = n_peaks_both,
  spacingMedian         = spacing_median,
  spacingMin            = spacing_min,
  spacingMax            = spacing_max,
  nCloseApposition      = n_close_apposition,

  # ---- Section 14: CME-stratified Motif2 (Figure 3.16) ----
  cmePosMotifTwoN       = cme_pos_m2_n,
  cmeNegMotifTwoN       = cme_neg_m2_n,
  cmePosMotifTwoPct     = cme_pos_m2_pct,
  cmeNegMotifTwoPct     = cme_neg_m2_pct,
  cmePosTotal           = cme_pos_total,
  cmeNegTotal           = cme_neg_total,
  cmeMotifTwoOR         = cme_m2_OR,
  cmeMotifTwoCIlo       = sprintf("%.2f", ft_cme_m2$conf.int[1]),
  cmeMotifTwoCIhi       = sprintf("%.2f", ft_cme_m2$conf.int[2]),
  cmeMotifTwoPval       = fmt_p(cme_m2_pval),

  # ---- Section 15: Trh-unique / Tgo-unique Motif2 (Figure 3.17) ----
  trhUniqueTotalN       = trh_unique_n_total,
  tgoUniqueTotalN       = tgo_unique_n_total,
  trhUniqueMotifTwoN    = trh_unique_m2_n,
  tgoUniqueMotifTwoN    = tgo_unique_m2_n,
  trhUniqueMotifTwoPct  = trh_unique_m2_pct,
  tgoUniqueMotifTwoPct  = tgo_unique_m2_pct,
  trhUniqueMotifTwoOR   = ft_trh_OR,
  tgoUniqueMotifTwoOR   = ft_tgo_OR,
  trhUniqueMotifTwoP    = fmt_p(ft_trh_p),
  tgoUniqueMotifTwoP    = fmt_p(ft_tgo_p),

  # ---- Section 16: Full-PWM FIMO specificity (Figure 3.17) ----
  mspecEnhPct           = mspec_enh_pct,
  mspecEnhN             = mspec_n_enh_total,
  mspecEnhPosN          = mspec_n_enh_m2_pos,
  jointMotifTwoPct      = joint_m2_pct,
  jointMotifTwoN        = joint_m2_n,
  mspecPeaksPct         = joint_m2_pct,
  mspecPeaksVsEnhOR     = mspec_peaks_vs_enh_or,     # 1.28
  mspecPeaksVsEnhP      = fmt_p(mspec_peaks_vs_enh_p), # 3.52e-3

  # ---- TEST1: Motif2 deconfounding regression ----
  motifTwoBeta          = m2_beta,
  motifTwoRegP          = fmt_p(m2_reg_p),
  motifTwoRegPHC        = fmt_p(m2_reg_p_hc),
  motifTwoRegBetaCIlo   = m2_beta_ci[1],
  motifTwoRegBetaCIhi   = m2_beta_ci[2],
  motifTwoRegOutcomeMedian = m2_outcome_median,
  motifTwoBinBeta       = m2_bin_beta,
  motifTwoBinP          = fmt_p(m2_bin_p),
  adjRReduced           = adjr_red,
  adjRFull              = adjr_full,

  # ---- TEST3: phyloP conservation ----
  nMotifTwoPhyloP       = m2_pp_n,
  flankWidth            = flank_w,
  motifTwoPhyloPMedian  = m2_pp_motif,
  flankPhyloPMedian     = m2_pp_flank,
  motifTwoPhyloPDiff    = m2_pp_diff,
  motifTwoPhyloPPropUp  = m2_pp_propup,
  motifTwoPhyloPP       = fmt_p(m2_pp_p),
  coreCactgPhyloP       = core_med,
  coreCactgPhyloPDiff   = core_diff,
  coreCactgPhyloPP      = fmt_p(core_p),

  # ---- Motif2 database-novelty scan (TOMTOM vs known Drosophila TFs) ----
  nMotifTwoDbMatrices   = motif2_db_nmatrices,   # 942
  nMotifTwoDbHits       = motif2_db_nhits,       # 0
  motifTwoDbThresh      = motif2_db_thresh,      # 0.05
  motifTwoDbNearestE    = motif2_db_nearestE,    # 2.9

  # ---- Distal/proximal GO diagnostic ----
  nProxPeaksGO          = n_promoter_ds,
  nDistPeaksGO          = n_distal_ds,
  nProxGoSigTerms       = nrow(res_prox_bp_sig),
  nDistGoSigTerms       = nrow(res_distal_bp_sig),
  distGoTopTerm         = dist_go_top_term,
  distGoTopOR           = dist_go_top_or,
  distGoTopFDR          = fmt_p(dist_go_top_fdr)

)

save_values("ch3", ch3_vals)
```

# Motif2 attribution: heterodimers vs. enhancers in general

The analyses above compare Trh/Tgo peaks against an enhancer catalogue.
That design cannot separate two accounts of the Motif2 enrichment: that
Motif2 is associated with the heterodimers, or that it is a compositional
feature of distal embryonic enhancers which the heterodimers happen to
bind. The partition test below separates them, by taking ONE enhancer
catalogue and splitting it only by whether a heterodimer peak overlaps the
enhancer. Sequence class, interval length and scan parameters are then
constant by construction.

Two conventions are load-bearing and are enforced in the script:

1. **Equal 201 bp windows for every scanned set.** The joint peaks average
   ~420 bp and the annotated enhancers ~860 bp; a longer sequence carries a
   higher per-sequence hit probability irrespective of biology, which
   understates the enrichment.
2. **`p < 1e-4` only, never FIMO's q.** FIMO computes q within a scan, so
   the burden scales with set size: the 35797-enhancer catalogue retains
   19% of its p-passing hits under `q < 0.05` while a 614-peak set retains
   88%. A q-threshold is therefore not comparable across sets of different
   size, and using one penalises the large comparator.

The analysis lives in its own script rather than inline, because it reads
all four heterodimer peak sets and belongs to no single chapter. It writes
the attribution tables (`data/motif2_attribution/`) and the figure
`results/figures/fig4_12_motif2_attribution.png` used in
the Chapter 4 section on Motif2 attribution (Chapter 4, because the partition
test requires the Sim/Dys/Sima peak sets and the non-bHLH-PAS controls
that Chapter 4 introduces).

```{r motif2-attribution, eval=FALSE}
# Sources: data/final_peaks IDR peak sets (all four heterodimers),
#          data/chapter3/enhancers_genomewide.fa,
#          data/ctrl_tf/fimo/* (Chapter 4 control scans)
# Outputs: data/motif2_attribution/*.csv
#          results/figures/fig4_12_motif2_attribution.png
# Run from the repository root:
#   Rscript 04_chapter_analyses/chapter4/ch4_motif2_attribution.R
```

```{r motif2-attribution-export, eval=FALSE}
# Lists which attribution values are recorded in results/reported_values.tsv.
# The pooled-attribution counts, odds ratios and p-values are in
# data/motif2_attribution/pooled_attribution_tests.csv; the enhancer/factor
# shares (attrEnh*, attrFac*) are recorded by
# 04_chapter_analyses/chapter4/ch4_motif2_nonenh_baseline.R.
A <- file.path(PROJECT_DIR, "data/motif2_attribution")
stopifnot(file.exists(file.path(A, "pooled_attribution_tests.csv")))
rv <- read.delim(VALUES_FILE, colClasses = "character", quote = "")
need <- c("attrPooledOR", "attrPooledCIlo", "attrPooledCIhi", "attrPooledP",
          "attrPooledBoundN", "attrPooledBoundPct", "attrUnboundN",
          "attrUnboundPct", "attrEnhPctShare", "attrFacPctShare",
          "attrEnhOR", "attrFacOR", "attrWindowBp",
          "attrTrhOR", "attrSimOR", "attrDysOR", "attrSimaOR",
          "mspecPeaksPctMatched", "mspecEnhPctMatched", "mspecORMatched")
missing <- setdiff(need, rv$name)
if (length(missing))
  message("attribution values not recorded in reported_values.tsv: ",
          paste(missing, collapse = ", "))
cat(sprintf("attribution values present: %d/%d\n", length(need) - length(missing),
            length(need)))
```

